---
title: "Ch_03_lab"
author: "Holden Jones"
date: '2022-07-22'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse) #loads the ggplot2 package. Could also use library(ggplot2)
library(maps)
library(mapproj)
```

##########################################################
#https://r4ds.had.co.nz/data-visualisation.html#exercises
##########################################################
# 1. Run ggplot(data = mpg). What do you see?
 a blank grey square

#2. How many rows are in mpg? How many columns?
234 rows, 11 columns

#3. What does the drv variable describe? Read the help for ?mpg to find out.
type of drive train

#4. Make a scatterplot of hwy vs cyl.
doesn't look good!
```{r}
ggplot(data = mpg, aes(x = hwy, y = cyl)) +
  geom_point()
```

#5. What happens if you make a scatterplot of class vs drv? Why is the plot not useful? 
overlap of points like the plot above tho
```{r}
ggplot(data = mpg, aes(x = class, y = drv)) +
  geom_point()
```


###########################################################
#https://r4ds.had.co.nz/data-visualisation.html#exercises-1
###########################################################
#1. What’s gone wrong with this code? Why are the points not blue?
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color="blue"))

# here ya go:
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue")
```
not blue b/c within the aes argument, looking for a variable called "blue"

#2. Which variables in mpg are categorical? Which variables are continuous? (Hint: type ?mpg to read the documentation for the dataset). How can you see this information when you run mpg?
running just mpg shows at the top what the variable type is - chr for categorical (ie a character)

#3.  Map a continuous variable to color, size, and shape. How do these aesthetics behave differently for categorical vs. continuous variables?
```{r}
# continuous variables:
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = cty))
# color w/ continuous looks good, but not easiest to read, better for cat.

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = cty))
# size w/ continuous looks great!

#ggplot(data = mpg) + 
  #geom_point(mapping = aes(x = displ, y = hwy, shape = cty))
# shape w/ continuous doesn't work - error cant map cont variable to shape

# categorical variables:
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = manufacturer))
# color looks great - actually better than for continuous

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = manufacturer))
# size is fine, but I think better for continuous

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape = manufacturer))
# only can do shape palette for six discrete values - error message and doesn't work with more than 6 
```

#4.  What happens if you map the same variable to multiple aesthetics?
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = manufacturer, size = manufacturer))
```
repeats information - not good!


#5.  What does the stroke aesthetic do? What shapes does it work with? (Hint: use ?geom_point)
stroke adjusts the thickness of the border for shapes that can take on different colors both inside and outside. It only works for shapes 21-24.

#6. What happens if you map an aesthetic to something other than a variable name, like aes(colour = displ < 5)? Note, you’ll also need to specify x and y.
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = manufacturer, color = displ <5))
```
This is actually pretty cool! Makes two colors, one for tru, one for false - could be a good way to show thresholds / limits for seed size and things like that

###########################################################
#https://r4ds.had.co.nz/data-visualisation.html#exercises-2
###########################################################

facet_wrap() to facet on one variable, facet_grid() to facet on multiple

#1. What happens if you facet on a continuous variable?
```{r}
# this is good
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2)

# w/ continuous variable it gets super hectic super quickly, not worth it too crazy!
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ cty, nrow = 2)
```


#2. What do the empty cells in plot with facet_grid(drv ~ cyl) mean? How do they relate to this plot?
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(drv ~ cyl)
```
means that there isn't a combo of those two variables - ie no 5 cylinder 4 wheel drive, or 5 cyl rear wheel drive, or 4 cylinder rear wheel drive cars in the dataset


#3. What plots does the following code make? What does . do?
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(drv ~ .)

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(. ~ cyl)
```
. controls the orientation of the faceting either horizontal or vertical

#4. Take the first faceted plot in this section:
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2)
```
#What are the advantages to using faceting instead of the colour aesthetic? What are the disadvantages? How might the balance change if you had a larger dataset?
 can see how it would be confusing w/ color right on top of each other - if dataset is super large with lots of different variable values you're thinking about faceting on, then it makes more sense to just do color maybe?
  

#5. Read ?facet_wrap. What does nrow do? What does ncol do? What other options control the layout of the individual panels? Why doesn’t facet_grid() have nrow and ncol arguments?
nrow controls the number of rows, ncol number of columns
facet_grid() is default in 1D - so they recommend facet_wrap

#6. When using facet_grid() you should usually put the variable with more unique levels in the columns. Why?
otherwise it gets pretty hectic / hard to read?

###########################################################
#https://r4ds.had.co.nz/data-visualisation.html#exercises-3
###########################################################
#1. Guess what geom would you use to draw a line chart? A boxplot? A histogram? An area chart?
geom_line, geom_boxplot, geom_histogram, geom_area

#2.  Run this code in your head and predict what the output will look like. Then, run the code in R and check your predictions.
```{r}
ggplot(data = mpg) +
  geom_line(mapping = aes(x = displ, y = hwy))

ggplot(data = mpg) +
  geom_boxplot(mapping = aes(x = drv, y = hwy)) # x has to be categorical

ggplot(data = mpg) +
  geom_histogram(mapping = aes(x = displ))

ggplot(data = mpg) +
  geom_area(mapping = aes(x = displ, y = hwy))
```

#3. What does show.legend = FALSE do? What happens if you remove it? Why do you think I used it earlier in the chapter?
hides the legend, default is tru so will show. maybe not to overwhelm us?

#4. What does the se argument to geom_smooth() do?
adds standard error shading
  
#5. Will these two graphs look different? Why/why not?
no, they will look the same - same aesthetics but first one is better b/c has them in ggplot command instead of geom_point() or geom_smooth()
### this is how you can display multiple geoms in the same plot! ###
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()

ggplot() + 
  geom_point(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy))
```

#6. Recreate the R code necessary to generate the following graphs (see page).

```{r}
#1
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() +
  geom_smooth(se = FALSE)

#2
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, group = drv)) + 
  geom_point() + 
  geom_smooth(se = FALSE)

#3
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) +
  geom_point() + 
  geom_smooth(se = FALSE)

#4
ggplot() + 
  geom_point(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) + 
  geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy), se = FALSE)

#5
ggplot() + 
  geom_point(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) + 
  geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy, linetype = drv), se = FALSE)

#6
ggplot() + 
  geom_point(data = mpg, mapping = aes(x = displ, y = hwy), color = 'white', size = 4) +
  geom_point(data = mpg, mapping = aes(x = displ, y = hwy, color = drv), size = 2)
# super cool! so there's probably a way to create a border of each point, but instead I just made an all white set of points from the same x y data, then overlaid the same points but colored that were a bit smaller - creates the white outline! cool!
```


###########################################################
#https://r4ds.had.co.nz/data-visualisation.html#exercises-4
###########################################################

Can generally use geoms and stats interchangeably b/c each geom has a default stat and each stat has a default geom

#1. What is the default geom associated with stat_summary()? How could you rewrite the previous plot to use that geom function instead of the stat function?
default geom for stat_summary is "pointrange

```{r}
# instead of this
ggplot(data = diamonds) + 
  stat_summary(
    mapping = aes(x = cut, y = depth),
    fun.min = min,
    fun.max = max,
    fun = median
  )

# can do this
ggplot(data = diamonds) + 
  geom_pointrange(
    mapping = aes(x = cut, y = depth),
    stat = "summary", # default stat is identity so need to change it to summary here
    fun.min = min,
    fun.max = max,
    fun = median
  )
```

#2. What does geom_col() do? How is it different to geom_bar()?
two types of bar charts in ggplot - geom_bar() and geom_col()
- bar makes height proportional to # of cases in each group (or can use weight to show the sum of the weights) stat_count() default
- col makes height of bars represent values in data stat_identity() default

#3. Most geoms and stats come in pairs that are almost always used in concert. Read through the documentation and make a list of all the pairs. What do they have in common?
not going to do a complete list - b/c there are so many! but most often will have the name in common ie. geom_boxplot goes with stat_boxplot

#4. What variables does stat_smooth() compute? What parameters control its behaviour?
y: predicted value
ymin: lower value of the confidence interval
ymax: upper value of the confidence interval
se: standard error

control w/
method, formula, se, n, span, fullrange, level
  
#5. In our proportion bar chart, we need to set group = 1. Why? In other words what is the problem with these two graphs?
problem is that the proportions for each graph are calculated within the groups - so will be 1 for each group - this doesn't tell us anything!
```{r}
# instead of
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, y = ..prop..))

# use this - specify group = 1 within mapping = aes()
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y = ..prop.., group = 1))

# instead of 
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, fill = color, y = ..prop..))

# use this - don't clearly understand all of it, but the y value creates a proportion, we can then fill off of this
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y = ..count.. / sum(..count..), fill = color))
```


###########################################################
#https://r4ds.had.co.nz/data-visualisation.html#exercises-5
###########################################################
#1. What is the problem with this plot? How could you improve it?
```{r}
# overplotting
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + 
  geom_point()

# so use jitter to fix this
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + 
  geom_jitter()
```

#2. What parameters to geom_jitter() control the amount of jittering?
width and height arguments control amount of jittering

#3. Compare and contrast geom_jitter() with geom_count().
geom_jitter is basically geom_point() offset - geom_count() counts the number of observations at each point
- can use either to account for overplotting, although geom_count() often harder to read than geom_jitter()

#4. What’s the default position adjustment for geom_boxplot()? Create a visualisation of the mpg dataset that demonstrates it.
default position is dodge2
```{r}
# default here
ggplot(data = mpg, mapping = aes(x = drv, y = hwy, colour = class)) + 
  geom_boxplot()

# changing to identity stacks them on top of each other
ggplot(data = mpg, mapping = aes(x = drv, y = hwy, colour = class)) + 
  geom_boxplot(position = "identity")
```


###########################################################
#https://r4ds.had.co.nz/data-visualisation.html#exercises-6
###########################################################

default coordinate system for ggplot is cartesian - can change this 
coord_flip() flips x and y axes

coord_quickmap() - sets aspect ratio correctly for maps. super important if plotting spatial data w/ ggplot2, not covering this in this book
```{r}
nz <- map_data("nz")

ggplot(nz, aes(long, lat, group = group)) +
  geom_polygon(fill = "white", colour = "black")

ggplot(nz, aes(long, lat, group = group)) +
  geom_polygon(fill = "white", colour = "black") +
  coord_quickmap()
```
coord_polar() uses polar coordinates
```{r}
bar <- ggplot(data = diamonds) + 
  geom_bar(
    mapping = aes(x = cut, fill = cut), 
    show.legend = FALSE,
    width = 1
  ) + 
  theme(aspect.ratio = 1) +
  labs(x = NULL, y = NULL)

bar
bar + coord_flip()
bar + coord_polar() # this is cool!
```

#1. Turn a stacked bar chart into a pie chart using coord_polar().
```{r}
ggplot(mpg, aes(x = factor(1), fill = drv)) +
  geom_bar(width = 1) # stacked bar chart

ggplot(mpg, aes(x = factor(1), fill = drv)) +
  geom_bar(width = 1) +
  coord_polar() # interesting, this makes a bullseye chart

# we need to specify theta = "y within coord_polar()
ggplot(mpg, aes(x = factor(1), fill = drv)) +
  geom_bar(width = 1) +
  coord_polar(theta = "y")
```

#2. What does labs() do? Read the documentation.
labs() modifies the labels with ggplot2

#3. What’s the difference between coord_quickmap() and coord_map()?
coord_map() uses map projections, default is Mercator projection
coord_quickmap() uses an approximate but fast projection ignoring curve of the earth
tried w/ the NZ example above and it didn't take longer, but again this is a pretty simple example

#4.  What does the plot below tell you about the relationship between city and highway mpg? Why is coord_fixed() important? What does geom_abline() do?
```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
  geom_point() + 
  geom_abline() +
  coord_fixed()
```
  
coord_fixed() sets the geom_abline() at a 45 degree angle which is important for humans!