---
title: "Ch20_lab"
author: "Holden Jones"
date: '2022-12-21'
output: html_document
---

# Vectors


--------------------------------------------------------------------------------
# 20.1 Introduction

vectors are objects that underlie vectors

## 20.1.1 Prerequisites

```{r}
library(tidyverse)
```

## 20.1.2 Vector basics

2 main types of vectors:
- Atomic vectors, of which there are six types: logical, integer, double,
  character, complex, and raw. Integer and double vectors are collectively known
  as numeric vectors.
- Lists, which are sometimes called recursive vectors because lists can contain
  other lists.

atomic vectors are homogenous
lists can be heterogenous
NULL is a related object, means absence of vector

every vector has two key properties:
1. type, which determine with typeof()
2. length, which determine with length()
```{r}
# type
typeof(letters)
typeof(1:10)

# length
x <- list("a", "b", 1:10)
length(x)
```

vectors can also contain arbitrary additional metadata in form of attributes
used to create augmented vectors which build on additional behaviour
- factors are built on top of integer vectors
- dates and date-times are built on top of numeric vectors
- data frames and tibbles are built on top of lists


--------------------------------------------------------------------------------
# 20.3 Important types of atomic vectors

4 most important types of atomic vector
- logical
- integer
- double
- character

## 20.3.1 Logical

Logical vectors are simplest type of atomic vector b/c can take only 3 values:
- FALSE
- TRUE
- NA
logical vectors usually constructed with comparison operators, also can use c()
```{r}
# comparison operator
1:10 %% 3 == 0

# c()
c(TRUE, TRUE, FALSE, NA)
```

## 20.3.2 Numeric

Integer and double vectors known collectively as numeric vectors
to make an integer, place L after the number
```{r}
typeof(1)
typeof(1L)
```

distinction b/ween integers and doubles is not usually important, but:
- doubles are approximations
- integers have one special value NA, while doubles have four: NA, NAN, Inf, -Inf

## 20.3.3 Character

most complex type of atomic vector, b/c each element is a string
already learned about strings in strings!
R uses global string pool - saves memory!

## 20.3.4 Missing values

each type of atomic vector has its own missing value
- logical = NA
- integer = NA_integer_
- double = NA_real_
- character = NA_character_
but don't worry about this too much, R will implicitly correct to vector type

## 20.3.5 Exercises

1. Describe the difference between is.finite(x) and !is.infinite(x).
*is.finite tells you which are finite (not infinite and not missing)*
*!is.infinite tells you which are not infinite, but won't cover missing values*
```{r}
x <- c(0, NA, NaN, Inf, -Inf)

is.finite(x)
!is.infinite(x)
```

2. Read the source code for dplyr::near() (Hint: to see the source code, 
  drop the ()). How does it work?
*checks to see if two numbers are within tolerance of each other*
```{r}
near(1000, 999.99999)
```

3. A logical vector can take 3 possible values. How many possible values can 
  an integer vector take? How many possible values can a double take? Use google
  to do some research.
*2^32 for integer vector, 2^64 for double values*

4. Brainstorm at least four functions that allow you to convert a double to an
  integer. How do they differ? Be precise.
*differ with how to treat rounding of double decimals*

5. What functions from the readr package allow you to turn a string into logical,
  integer, and double vector?
*parse_logical, parse_integer, and parse_double*


--------------------------------------------------------------------------------
# 20.4 Using atomic vectors

Important tools for working atomic vectors:
- How to convert from one type to another
- How to tell if an object is a specific type of vector
- What happens when work with vectors of different lengths
- How to name elements of a vector
- How to pull out elements of interest

## 20.4.1 Coercion

2 ways to convert or corece one type of vector to another:
- explicit coercion
- implicit coercion

focus on implicit coercion:
- TRUE converted to 1 and FALSE to 0
```{r}
x <- sample(20, 100, replace = TRUE)
y <- x > 10

sum(y)  # how many are greater than 10?
mean(y) # what proportion are greater than 10?
```

## 20.4.2 Test functions

use is_* functions to specifically test for each type of vector

## 20.4.3 Scalars and recycling rules

shorter vector repeated or recycled to same length as longer vector
most built-in functions are vectorised, meaning will operate on vector of numbers
```{r}
sample(10) + 100
runif(10) > 0.5
```

## 20.4.4 Naming vectors

all types of vectors can be named, useful for subsetting
```{r}
# during creation:
c(x = 1, y = 2, z = 4)
set_names(1:3, c("a", "b", "c"))
```

## 20.4.5 Subsetting

so far we've only used filter, but that only works with tibbles
with vectors need new function: []

Can subset a vector with 4 types of things:

1.numeric vector containing only integers, integers all positive, all neg, or 0
```{r}
x <- c("hi", "how", "are", "you", "doing")

x[c(3, 2, 5)]

# can create longer output than input by repeating positions:
x[c(1, 1, 5, 5, 5, 2)]

# can drop from specific positions
x[c(-1, -3, -5)]

# but don't mix positive and negative values
```

2. subsetting with logical vector keeps all values corresponding to TRUE
- best for comparison functions
```{r}
x <- c(10, 3, NA, 5, 8, 1, NA)

x[!is.na(x)]
```

3. If have named vector, can subset with character vector
```{r}
x <- c(abc = 1, def = 2, xyz = 5)
x[c("xyz", "def")]
```

4. also can subset nothing! leaves everything, useful for matrices 
- ex. all rows / all columns

## 20.4.6 Exercises

1. What does mean(is.na(x)) tell you about a vector x? 
  What about sum(!is.finite(x))?
*1st tells us what percentage of x is NA*
*2nd tells us how many are not finite*
```{r}
x <- c(10, 3, NA, 5, 8, 1, NA)
mean(is.na(x))
sum(!is.finite(x))
```

2. Carefully read the documentation of is.vector(). What does it actually test 
  for? Why does is.atomic() not agree with the definition of atomic vectors 
  above?
*is.vector checks to see if vector has attribute other than a name*
*is.atomic checks to see if vector has an atomic attribute*
*will be atomic even if has extra attributes*

3. Compare and contrast setNames() with purrr::set_names().
*set_names is more complicated than setNames*

4. Create functions that take a vector as input and returns:
a - The last value. Should you use [ or [[?
b - The elements at even numbered positions.
c - Every element except the last value.
d - Only even numbers (and no missing values).
```{r}
x <- c(10, 3, NA, 5, 8, 1, NA)
y <- sample(20, 100, replace = TRUE)

# a
tail(y, n=1)
# already exists, called tail

last_value <- function(x) {
  # check for case with no length
  if (length(x)) {
    x[[length(x)]]
  } else {
    x
  }
}

# b
even_indices <- function(x) {
  if (length(x)) {
    x[seq_along(x) %% 2 == 0]
  } else {
    x
  }
}

even_indices(y)

# c
all_but_last <- function(x) {
  n <- length(x)
  if (n) {
    x[-n]
  } else {
    x
  }
}
    
all_but_last(x)
all_but_last(y)

# d
# not accounting for NAs
only_even_numbers <- function(x) {
  x[x %% 2 == 0]
}
  
only_even_numbers(y)
only_even_numbers(x)
```

5. Why is x[-which(x > 0)] not the same as x[x <= 0]?
*they differ in how they treat missing values*

6. What happens when you subset with a positive integer that’s bigger than the 
  length of the vector? What happens when you subset with a name that doesn’t 
  exist?
*returns NA*


--------------------------------------------------------------------------------
# 20.5 Recursive vectors (lists)

lists are a stepup in complexity from vectors
can contain mix of objects
can even contain other lists!
use str() b/c focuses on structure, not contents

## 20.5.1 Visualizing lists

3 principles of visualizing lists
- lists have rounded corners, atomic vectors have square corners
- children drawn inside their parent, slightly darker background
- orientation of children isn't important

## 20.5.2 Subsetting

## RESUME HERE
3 ways to subset a list
```{r}

```






