---
title: "Ch20_lab"
author: "Holden Jones"
date: '2022-12-21'
output: html_document
---

# Vectors


--------------------------------------------------------------------------------
# 20.1 Introduction

vectors are objects that underlie vectors

## 20.1.1 Prerequisites

```{r}
library(tidyverse)
```

## 20.1.2 Vector basics

2 main types of vectors:
- Atomic vectors, of which there are six types: logical, integer, double,
  character, complex, and raw. Integer and double vectors are collectively known
  as numeric vectors.
- Lists, which are sometimes called recursive vectors because lists can contain
  other lists.

atomic vectors are homogenous
lists can be heterogenous
NULL is a related object, means absence of vector

every vector has two key properties:
1. type, which determine with typeof()
2. length, which determine with length()
```{r}
# type
typeof(letters)
typeof(1:10)

# length
x <- list("a", "b", 1:10)
length(x)
```

vectors can also contain arbitrary additional metadata in form of attributes
used to create augmented vectors which build on additional behaviour
- factors are built on top of integer vectors
- dates and date-times are built on top of numeric vectors
- data frames and tibbles are built on top of lists


--------------------------------------------------------------------------------
# 20.3 Important types of atomic vectors

4 most important types of atomic vector
- logical
- integer
- double
- character

## 20.3.1 Logical

Logical vectors are simplest type of atomic vector b/c can take only 3 values:
- FALSE
- TRUE
- NA
logical vectors usually constructed with comparison operators, also can use c()
```{r}
# comparison operator
1:10 %% 3 == 0

# c()
c(TRUE, TRUE, FALSE, NA)
```

## 20.3.2 Numeric

Integer and double vectors known collectively as numeric vectors
to make an integer, place L after the number
```{r}
typeof(1)
typeof(1L)
```

distinction b/ween integers and doubles is not usually important, but:
- doubles are approximations
- integers have one special value NA, while doubles have four: NA, NAN, Inf, -Inf

## 20.3.3 Character

most complex type of atomic vector, b/c each element is a string
already learned about strings in strings!
R uses global string pool - saves memory!

## 20.3.4 Missing values

each type of atomic vector has its own missing value
- logical = NA
- integer = NA_integer_
- double = NA_real_
- character = NA_character_
but don't worry about this too much, R will implicitly correct to vector type

## 20.3.5 Exercises

## RESUME HERE ##

1. Describe the difference between is.finite(x) and !is.infinite(x).

2. Read the source code for dplyr::near() (Hint: to see the source code, 
  drop the ()). How does it work?

3. A logical vector can take 3 possible values. How many possible values can 
  an integer vector take? How many possible values can a double take? Use google
  to do some research.

4. Brainstorm at least four functions that allow you to convert a double to an
  integer. How do they differ? Be precise.

5. What functions from the readr package allow you to turn a string into logical,
  integer, and double vector?








