---
title: "Exploratory Data Analysis (EDA)"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

# Exploratory Data Analysis

--------------------------------------------------------------------------------
# 7.1 Introduction

1. Generate questions about your data.

2. Search for answers by visualising, transforming, and modelling your data.

3. Use what you learn to refine your questions and/or generate new questions.

```{r}
library(tidyverse)
library(nycflights13)
```


--------------------------------------------------------------------------------
# 7.2 Questions

Develop an understanding of the data. Ask lots of questions to get good questions.

Good questions to start with:

1. What type of variation occurs within my variables?

2. What type of covariation occurs between my variables?

Definitions:

*A variable* - a quantity, quality, or property that you can measure.

*A value* - the state of a variable when you measure it. The value of a variable may change from measurement to measurement.

*An observation* - a set of measurements made under similar conditions (you usually make all of the measurements in an observation at the same time and on the same object). An observation will contain several values, each associated with a different variable. I’ll sometimes refer to an observation as a data point.

*Tabular data* - a set of values, each associated with a variable and an observation. Tabular data is tidy if each value is placed in its own “cell”, each variable in its own column, and each observation in its own row.


--------------------------------------------------------------------------------
# 7.3 Variation

######################################################################
https://r4ds.had.co.nz/exploratory-data-analysis.html#exercises-15##
######################################################################

## 7.3.1 Visualising distributions

*use a bar chart to visualise a categorical variable*
```{r}
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut))

# can also compute how many observations occured with count()
diamonds %>% 
  count(cut)
```

*use a histogram to visualise a continuous variable*
```{r}
ggplot(data = diamonds) +
  geom_histogram(mapping = aes(x = carat), binwidth = 0.5)

# can compute by hand using cut_width()
diamonds %>% 
  count(cut_width(carat, 0.5))

# important to explore a variety of binwidths
smaller <- diamonds %>% 
  filter(carat < 3)
  
ggplot(data = smaller, mapping = aes(x = carat)) +
  geom_histogram(binwidth = 0.1)

# use geom_freqpoly to overlay multiple histograms on top of each other, colored by cut in this example
ggplot(data = smaller, mapping = aes(x = carat, colour = cut)) +
  geom_freqpoly(binwidth = 0.1)
```

## 7.3.4 Exercises

1. Explore the distribution of each of the `x`, `y`, and `z` variables in `diamonds`. What do you learn? Think about a diamond and how you might decide which dimension is the length, width, and depth. 
*looks like z has less variation than x and y*
```{r}
# histogram of each
ggplot(diamonds) + 
  geom_histogram(mapping = aes(x = x), binwidth = 0.5)

ggplot(diamonds) + 
  geom_histogram(mapping = aes(x = y), binwidth = 0.5)

ggplot(diamonds) + 
  geom_histogram(mapping = aes(x = z), binwidth = 0.5)

# summary of each
summary(select(diamonds, x, y, z))
```

2. Explore the distribution of `price`. Do you discover anything unusual or surprising? (Hint: Carefully think about the `binwidth` and make sure you try a wide range of values.)
*not surprisingly a huge range of prices, with some absurdly expensive but the majority much cheaper - also see that there are no diamonds priced at $1,500 - super weird!*
```{r}
ggplot(diamonds) + 
  geom_histogram(mapping = aes(x = price), binwidth = 10)

summary(select(diamonds, price))
```

3. How many diamonds are 0.99 carat? How many are 1 carat? What do you think is the cause of the difference?
*23 are 0.99, over 1500 are 1 carat - I think price is the cause*
```{r}
ggplot(diamonds) + 
  geom_histogram(mapping = aes(x = carat), binwidth = 0.01)

(diamonds %>%
  filter(carat == 0.99))
#23 are .99

(diamonds %>% 
  filter(carat == 1.00))
#wow over 1500 are one carat! clearly there's a big difference here!
```

4. Compare and contrast `coord_cartesian()` vs `xlim()` or `ylim()` when zooming in on a histogram. What happens if you leave `binwidth` unset? What happens if you try and zoom so only half a bar shows?
*coord_cartesian zooms in so allows us to remove outliers*
*xlim and y lim does something similar but removes the data, whereas coord cart just zooms in the hist*
```{r}
ggplot(data=diamonds) + 
  geom_histogram(aes(carat)) +
  coord_cartesian(xlim=c(1,3))

ggplot(data=diamonds) + 
  geom_histogram(aes(carat)) +
  xlim(1,3) +
  ylim(0, 1000)
```


--------------------------------------------------------------------------------
# 7.4 Missing values

Most datasets have missing values, its always important to remember the difference between missing values and zeros, which can sometimes be confused.

######################################################################
https://r4ds.had.co.nz/exploratory-data-analysis.html#exercises-16##
######################################################################

ifelse() has three arguments. The first argument test should be a logical vector. The result will contain the value of the second argument, yes, when test is TRUE, and the value of the third argument, no, when it is false.

*Hadley recommends replacing unusual values with missing values instead of deleting the entire row with strange values. Use mutate() to replace variables with a modified copy*
```{r}
diamonds %>% 
  mutate(y = ifelse(y < 3 | y > 20, NA, y)) %>% # change to NA here
  ggplot(mapping = aes(x = x, y = y)) + 
  geom_point()
```
*ggplot and R believe that missing values should not be removed! - can suppress warning with na.RM = TRUE*

*sometimes the missing values are important. In flights the missing values are cancelled flights. These are important! can compare to normal flights*
```{r}
flights %>% 
  mutate(
    cancelled = is.na(dep_time),
    sched_hour = sched_dep_time %/% 100,
    sched_min = sched_dep_time %% 100,
    sched_dep_time = sched_hour + sched_min / 60
  ) %>% 
  ggplot(mapping = aes(sched_dep_time)) + 
    geom_freqpoly(mapping = aes(colour = cancelled), binwidth = 1/4)
```

## 7.4.1 Exercises

1. What happens to missing values in a histogram? What happens to missing values in a bar chart? Why is there a difference?
*geom_bar puts NAs as a category while histogram removes them. This makes sense because histogram is continuous and barplot is categorical*
*values removed with non-finite values (stat_bin) for hist*

#### RESUME HERE###
```{r}
#### look at answer sheets
```

2. What does `na.rm = TRUE` do in `mean()` and `sum()`?

```{r}
mean(diamonds2$y, na.rm = TRUE) #allows us to take mean by removing NA values

sum(diamonds2$y, na.rm = TRUE) # this doesn't seem like it has an impact. perhaps this is default for sum?
```



# 7.5 Covariation, part I

######################################################################
https://r4ds.had.co.nz/exploratory-data-analysis.html#exercises-17##
######################################################################

*if we display density instead of count on the y axis it will make the area under the curve equal to one - seems like this can be a way to standardize different observations that may have vastly different counts*

*in diamonds, cut is an ordered factor: fiar is worse than good, which is worse than very good etc. So the order of these categorical variables is intrinsic and does matter. We can use the reorder() function to reorder the variables*

```{r}
ggplot(data = mpg) +
  geom_boxplot(mapping = aes(x = reorder(class, hwy, FUN = median), y = hwy))
  # here reorder by the median hwy mpg for each class

ggplot(data = mpg) +
  geom_boxplot(mapping = aes(x = reorder(class, hwy, FUN = median), y = hwy)) +
  coord_flip() # cord_flip() flips the ggplot 90 degrees. Good if have long variable names
```


1. Use what you’ve learned to improve the visualisation of the departure times of cancelled vs. non-cancelled flights dataset from  the `nycflights13` library. Hint, think boxplot!

```{r}
library(nycflights13)

flights2 <- flights %>% # now can use this df with new variable cancelled
  mutate(
    cancelled = is.na(dep_time),
    sched_hour = sched_dep_time %/% 100,
    sched_min = sched_dep_time %% 100,
    sched_dep_time = sched_hour + sched_min / 60
  )

ggplot(data = flights2, mapping = aes(x = cancelled, y = sched_dep_time)) +
  geom_boxplot()
# I think this works?
```


2. What variable in the diamonds dataset is most important for predicting the price of a diamond? How is that variable correlated with cut? Why does the combination of those two relationships lead to lower quality diamonds being more expensive?

*I guess we'll just use a corplot here? seems kinda weird that this would be asked*
```{r}
#install.packages("corrplot")
library("corrplot")

diamond.cor <-cor(diamonds[,-(2:4)])
corrplot.mixed(diamond.cor)

# looks like carat, x, y, z are strongly correlated with price
```

3. Install the ggstance package, and create a horizontal boxplot. How does this compare to using `coord_flip()`?
**SKIP**


4. One problem with boxplots is that they were developed in an era of much smaller datasets and tend to display a prohibitively large number of “outlying values”. One approach to remedy this problem is the letter value plot. Install the lvplot package, and try using `geom_lv()` to display the distribution of price vs cut. What do you learn? How do you interpret the plots?
**SKIP**


5. Compare and contrast `geom_violin()` with a facetted `geom_histogram()`, or a coloured `geom_freqpoly()`. What are the pros and cons of each method?

```{r}
#geom_violin:
ggplot(data = diamonds, mapping = aes(x = cut, y = price)) +
  geom_violin() # I think geom_violin needs to use a categorical x variable
# but here it doesn't look like cut is a super reliable predictor of price
# tells us a bit about distribution but we don't have any idea of how many are in each category

# geom_histogram:
ggplot(data = diamonds, mapping = aes(x = price)) +
  geom_histogram() +
  facet_wrap(~cut)
# faceting by a categorical variable shows us the distributions of each cat. and it also shows us the frequency

# geom_freqpoly():
ggplot(data = diamonds, mapping = aes(x = price, y = ..count..)) + 
  geom_freqpoly(mapping = aes(colour = cut), binwidth = 500)
# looks pretty cool and easier to interpret than violin plot. Can change y to be either density or count (..density..)
# This means we can input y = ..count.. if we care about the absolute numbers of each cat variable, or can leave as density if that's not important. So we've got a lot of information in this one little plot. Don't need to facet like we did for geom_histogram and it looks cool 
```


6. If you have a small dataset, it’s sometimes useful to use `geom_jitter()` to see the relationship between a continuous and categorical variable. The ggbeeswarm package provides a number of methods similar to `geom_jitter()`. List them and briefly describe what each one does.
**SKIP**

# 7.5 Covariation, part II

######################################################################
https://r4ds.had.co.nz/exploratory-data-analysis.html#exercises-18##
######################################################################

1. How could you rescale the count dataset given below to more clearly show the distribution of cut within colour, or colour within cut?

*Not sure what the question is asking here. Here's a plot!*
```{r}
diamonds %>% 
  count(color, cut) %>%  
  ggplot(mapping = aes(x = color, y = cut)) +
    geom_tile(mapping = aes(fill = n))
```

2. Use `geom_tile()` together with `dplyr` to explore how average flight delays in the `nycflights13` library vary by destination and month of year. What makes the plot difficult to read? How could you improve it?

*count month, year, destination? that's gonna be a ton of "bins"*
*then plot this with geom_tile against flight delays?*

```{r}
library(nycflights13)

flights %>% 
  count(month, dest) %>%  
  ggplot(mapping = aes(x = month, y = dest)) +
    geom_tile(mapping = aes(fill = n))

# make dataframe grouped by destination and month, summarize to get avg_delay

avg.delay <- flights %>% group_by(dest, month) %>% summarize(avg_delay=mean(dep_delay, na.rm=TRUE))
        
ggplot(data=avg.delay, aes(x=dest, y=month)) + 
  geom_tile(aes(fill=avg_delay))

# still probably too much going on for geom_tile to make sense here
```


3. Why is it slightly better to use `aes(x = color, y = cut)` rather than `aes(x = cut, y = color)` in the example below?

*this way has smaller count values on the bottom and larger ones on the top, the other way has the reverse arrangement*
```{r}
diamonds %>% 
  count(color, cut) %>%  
  ggplot(mapping = aes(x = color, y = cut)) +
    geom_tile(mapping = aes(fill = n))
```

# 7.5 Covariation, part III

######################################################################
https://r4ds.had.co.nz/exploratory-data-analysis.html#exercises-19##
######################################################################

*geom_bin2d() and geom_hex() bins in two dimensions*
*need hexbin package*

```{r}
ggplot(data = diamonds) +
  geom_bin2d(mapping = aes(x = carat, y = price))

install.packages("hexbin")
ggplot(data = diamonds) +
  geom_hex(mapping = aes(x = carat, y = price))
```


1. Instead of summarising the conditional distribution with a boxplot (given below), you could use a frequency polygon. What do you need to consider when using `cut_width()` vs `cut_number()`? How does that impact a visualisation of the 2d distribution of carat and price?

*cut_width() makes every boxplot the same width, cut_number() determines boxplot width based on the distribution of the number of observations specified*
```{r}
smaller <- diamonds %>% 
  filter(carat < 3)

ggplot(data = smaller, mapping = aes(x = carat, y = price)) + 
  geom_boxplot(mapping = aes(group = cut_number(carat, 20)))

ggplot(data = smaller, mapping = aes(x = price)) + 
  geom_freqpoly(mapping = aes(group = cut_width(carat, 0.5)))
```


2. Visualise the distribution of `carat`, partitioned by `price`.
*aes has group by and color by price. keep both within aes to retain the label info which is important*
```{r}
ggplot(data = diamonds, mapping = aes(x = carat)) + 
  geom_freqpoly(mapping = aes(group = cut_width(price, 5000), col=cut_width(price, 5000)))
```

3. How does the price distribution of very large diamonds compare to small diamonds? Is it as you expect, or does it surprise you?
*I think this is as we'd expect. Small diamonds have  apeak at lower prices, medium ones have a peak at higher prices, large ones have a peak at higher prices and really large ones have smaller sample sizes and quite a bit of variation*
```{r}
ggplot(data = diamonds, mapping = aes(x = price)) + 
  geom_freqpoly(mapping = aes(group = cut_width(carat, 1), col=cut_width(carat, 1)))
```

4. Combine two of the techniques you’ve learned to visualise the combined distribution of `cut`, `carat`, and `price`.

```{r}
ggplot(diamonds, aes(x = cut_number(carat, 8), y = price, color = cut)) +
  geom_boxplot()
```

5. Two dimensional plots reveal outliers that are not visible in one dimensional plots. For example, some points in the plot below have an unusual combination of x and y values, which makes the points outliers even though their x and y values appear normal when examined separately. Why is a scatterplot a better display than a binned plot for this case?
*I think a bin would just throw these outliers either in with the normal group or perhaps give them their own bin but it would really depend on how the bins were divided. You could potentially lose a lot of information if the outliers got lumped in with the more normal values.*
```{r}
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = x, y = y)) +
  coord_cartesian(xlim = c(4, 11), ylim = c(4, 11))
```