---
title: "Tibble"
author: "Holden Jones"
date: '2022-08-24'
output: html_document
---

# Tibbles


--------------------------------------------------------------------------------
# 10.1 Introduction

Tibbles are data frames, but are tweaked to make life a bit easier. Looking into
the tibble package

Most of the time will refer to data frames and tibbles interchangeably

```{r}
library(tidyverse)
```


--------------------------------------------------------------------------------
# 10.2 Creating tibbles

Almost all functions in this book create tibbles - tibbles are a unifying 
feature of the diyverse

as_tibble() coerces data frame into a tibble, tribble() for data entry in code
```{r}
as_tibble(iris)

tribble(
  ~x, ~y, ~z,
  #--|--|----
  "a", 2, 3.6,
  "b", 1, 8.5
)
```


--------------------------------------------------------------------------------
# 10.3 Tibbles vs. data.frame

Two main differences: printing and subsetting

## 10.3.1 Printing

Tibbles only show the first 10 rows and the columns that fit on the screen
Also each column reports its type
You can alter the default print settings tho, pretty easily
```{r}
#runif() gives random number 0-1 (default), length specified
tibble(
  a = lubridate::now() + runif(1e2) * 86400,
  b = lubridate::today() + runif(1e2) * 30, 
  c = 1:1e2,
  d = runif(1e2),
  e = sample(letters, 1e2, replace = TRUE)
)
```

## 10.3.2 Subsetting

$ and [[ are tools for pulling out single variables
```{r}
df <- tibble(
  x = runif(5),
  y = rnorm(5)
)

# Extract by name
df$x
#> [1] 0.73296674 0.23436542 0.66035540 0.03285612 0.46049161
df[["x"]]
#> [1] 0.73296674 0.23436542 0.66035540 0.03285612 0.46049161

# Extract by position
df[[1]]
#> [1] 0.73296674 0.23436542 0.66035540 0.03285612 0.46049161

# Try with mpg datset
mpg$hwy
mpg[["cty"]]
```

To use these in a pipe, you’ll need to use the special placeholder .:
```{r}
df %>% .$x
#> [1] 0.73296674 0.23436542 0.66035540 0.03285612 0.46049161
df %>% .[["x"]]
#> [1] 0.73296674 0.23436542 0.66035540 0.03285612 0.46049161

mpg %>% .$hwy
```

Compared to a data.frame, tibbles are more strict: they never do partial matching, and they will generate a warning if the column you are trying to access does not exist.
  - Otherwise they seem pretty similar so far

  
------------------------------------------------------------------------------
# 10.4 Interacting with older code

Some older functions don’t work with tibbles. If you encounter one of these functions, use as.data.frame() to turn a tibble back to a data.frame:

The main reason that some older functions don’t work with tibble is the [ function. We don’t use [ much in this book because dplyr::filter() and dplyr::select() allow you to solve the same problems with clearer code (but you will learn a little about it in vector subsetting). With base R data frames, [ sometimes returns a data frame, and sometimes returns a vector. With tibbles, [ always returns another tibble.


--------------------------------------------------------------------------------
# 10.5 Exercises

1. How can you tell if an object is a tibble? (Hint: try printing mtcars, which is a regular data frame).
*It tells you right at the top in print! A tibble
```{r}
(mtcars)
(mpg)
```

2. Compare and contrast the following operations on a data.frame and equivalent tibble. What is different? Why might the default data frame behaviours cause you frustration?
*In this example there is an observation and column that both contain "a" - so could be confusing regarding what is returned*
*Aah okay, df$x also finds the xyz column b/c it begins with x, so this could be confusing down the line!*
```{r}
df <- data.frame(abc = 1, xyz = "a")
df$x
df[, "xyz"]
df[, c("abc", "xyz")]

as_tibble(df)
df$x
df[, "xyz"]
df[, c("abc", "xyz")]
```

3. If you have the name of a variable stored in an object, e.g. var <- "mpg", how can you extract the reference variable from a tibble?
*Just use the [[]] - the $ would look for a column*

4. Practice referring to non-syntactic names in the following data frame by:

  1. Extracting the variable called 1.

  2. Plotting a scatterplot of 1 vs 2.

  3. Creating a new column called 3 which is 2 divided by 1.

  4. Renaming the columns to one, two and three.
```{r}
annoying <- tibble(
  `1` = 1:10,
  `2` = `1` * 2 + rnorm(length(`1`))
)

#1.
annoying[["1"]]

#2.
ggplot(annoying, mapping = aes(annoying$"1", annoying$"2")) +
  geom_point()
# yeah this is annoying, warning issued but this works

#3.
annoying$"3" <- 2/1

#4.
annoying %>%
  rename(one = 1,
         two = 2,
         three = 3
         )
```
  
5. What does tibble::enframe() do? When might you use it?
*converts named atomic vectors or lists to one or two column data frames*
*maybe you want to put a bunch of strings into a df b/c data was entered weirdly?*

6. What option controls how many additional column names are printed at the footer of a tibble?
*print() n_extra argument for printing tibble rows*