---
title: "Data import"
author: "Holden Jones"
date: '2022-08-28'
output: html_document
---

# Data import


--------------------------------------------------------------------------------
# 11.1 Prerequisites

Use the readr package which is part of the core tidyverse
```{r}
library(tidyverse)
```


--------------------------------------------------------------------------------
# 11.2 Getting started

Most readr functions turn flat files into data frames
```{r}
# can also supply inline csv file - important for reproducing code
read_csv("a,b,c
1,2,3
4,5,6")
# it's sensitive to line placement
```

Can skip comments or specific lines when reading inline csvs
*can use \n to add a new line*

## 11.2.1 Compared to base R

readr much faster, produces tibbles instead of dataframes, more reproducible

## 11.2.2 Exercises

1. What function would you use to read a file where fields were separated with
“|”?
*read_delim()*

2. Apart from file, skip, and comment, what other arguments do read_csv() and read_tsv() have in common?
*lots of them, can use following code to get them*
intersect(names(formals(read_csv)), names(formals(read_tsv)))

3. What are the most important arguments to read_fwf()?
*probably col_positions and col_types, but also locale - regional defaults*

4. Sometimes strings in a CSV file contain commas. To prevent them from causing problems they need to be surrounded by a quoting character, like " or '. By default, read_csv() assumes that the quoting character will be ". What argument to read_csv() do you need to specify to read the following text into a data frame?
test <- c(x,y\n1,'a,b')
*specificy quote argument*
```{r}
x <- "x,y\n1,'a,b'"

read_csv(x, quote = "'")
```

5. Identify what is wrong with each of the following inline CSV files. What happens when you run the code?
*hard to tell b/c we don't know what they're going for here...*
```{r}
read_csv("a,b\n1,2,3\n4,5,6")
read_csv("a,b,c\n1,2,3\n4,5,6") # added c column

read_csv("a,b,c\n1,2\n1,2,3,4")
read_csv("a,b,c,d\n1,2,,\n1,2,3,4") # added d column, put ,, for NAs 

read_csv("a,b\n\"1")
read_csv("a,b\n,,\n1") # fixed quotes, inserted second row, ,, for NAs

read_csv("a,b\n1,2\na,b")
read_csv("a,b\n1,2\n'a','b'") 
# idk this looks good! but values have same value as columns so put ''

read_csv("a;b\n1;3")
read_csv("a,b\n1,3") # removed semicolon
```


--------------------------------------------------------------------------------
# 11.3 Parsing a vector

Little detour to talk about parse_*() functions - take a character vector and 
return a more specialized vector like a logical, integer, or date
```{r}
str(parse_logical(c("TRUE", "FALSE", "NA")))
#>  logi [1:3] TRUE FALSE NA
str(parse_integer(c("1", "2", "3")))
#>  int [1:3] 1 2 3
str(parse_date(c("2010-01-01", "1979-10-14")))
#>  Date[1:2], format: "2010-01-01" "1979-10-14"
```

parse functions are all uniform 
- 1st argument is character vector to parse
- na argument says what should we treat as missing

8 different types of parsers

## 11.3.1 Numbers

3 problems with parsing a number
- people write numbers differently; . or , for decimal
- numbers surrounded by characters to provide context ie $, %
- grouping characters to make easier to read (ie . or , for 1,000)

locale notion - parsing options set by place. default is US-centric

parse_number() to remove currency and other things, just get the number even if
in text
```{r}
parse_number("$100")
#> [1] 100
parse_number("20%")
#> [1] 20
parse_number("It cost $123.45")
#> [1] 123.45
```

## 11.3.2 Strings

seems like parse_character() should be really simple - should just return its 
input - but alas! many different ways to represent the same string

need to dive into how computers represent strings
```{r}
charToRaw("Holden")
```

each hexadecimal number is represented by a byte of information
readr uses UTF-8 system of encoding

other languages may need other encoding systems, which can be specified in 
parse_character
```{r}
#parse_character(x1, locale = locale(encoding = "Latin1"))
#> [1] "El Niño was particularly bad this year"
#parse_character(x2, locale = locale(encoding = "Shift-JIS"))
#> [1] "こんにちは"
```

can also use guess_encoding - may take a few tries tho!

# 11.3.3 Factors

factors represent categorical variables with a known set of possible values
if a value is outside an expected level will return an error
*seems like the logic here is similar to excel formatting*
```{r}
fruit <- c("apple", "banana")
parse_factor(c("apple", "banana", "bananana"), levels = fruit) # uh oh!
```

## 11.3.4 Dates, date-times, and times

parse_datetime(), parse_date(), and hms::parse_time() each expect and give diff things
use class from hms package for parsing times

if non english language and using month names, can set lang via locale

## 11.3.5 Exercises

1. What are the most important arguments to locale()?
*probably date_names and date_format to set language*

2. What happens if you try and set decimal_mark and grouping_mark to the same character? What happens to the default value of grouping_mark when you set decimal_mark to “,”? What happens to the default value of decimal_mark when you set the grouping_mark to “.”?
*decimal_mark and grouping_mark must be different, so default is for the other to be diff*
```{r}
locale(decimal_mark = ".", grouping_mark = ".")
```

3. I didn’t discuss the date_format and time_format options to locale(). What do they do? Construct an example that shows when they might be useful.
*important in L.A. and military for example*
```{r}
dates <- c("30-08-2022", "07-08-2022")
times <- c("14:47, ")

# to see what default formats are
locale()

# can also change these to set custom parameters
locale_custom <- locale(date_format = "Day %d Mon %M Year %y",
                 time_format = "Sec %S Min %M Hour %H")
date_custom <- c("Day 01 Mon 02 Year 03", "Day 03 Mon 01 Year 01")

parse_date(date_custom)
#> Warning: 2 parsing failures.
#> row col   expected                actual
#>   1  -- date like  Day 01 Mon 02 Year 03
#>   2  -- date like  Day 03 Mon 01 Year 01
#> [1] NA NA
parse_date(date_custom, locale = locale_custom)
#> [1] "2003-01-01" "2001-01-03"
time_custom <- c("Sec 01 Min 02 Hour 03", "Sec 03 Min 02 Hour 01")
parse_time(time_custom)
#> Warning: 2 parsing failures.
#> row col   expected                actual
#>   1  -- time like  Sec 01 Min 02 Hour 03
#>   2  -- time like  Sec 03 Min 02 Hour 01
#> NA
#> NA
parse_time(time_custom, locale = locale_custom)
#> 03:02:01
#> 01:02:03
```

4. If you live outside the US, create a new locale object that encapsulates the settings for the types of file you read most commonly.
*for Ecuador*
```{r}
locale() # check that default is US-centric

parse_date("30/08/2022") # try with default, no work

ecuador_locale <- locale( # set custom for EC
  date_format = "%d/%m/%Y",
)

parse_date("30/08/2022", locale = ecuador_locale)
# works!
```

5. What’s the difference between read_csv() and read_csv2()?
*values are separated with ; instead of ,*

6. What are the most common encodings used in Europe? What are the most common encodings used in Asia? Do some googling to find out.
*Europe: ISO-8859, Asia: GB18030*

7. Generate the correct format string to parse each of the following dates and times:
```{r}
d1 <- "January 1, 2010"
parse_date(d1, "%B %d, %Y")

d2 <- "2015-Mar-07"
parse_date(d2, "%Y-%b-%d")

d3 <- "06-Jun-2017"
parse_date(d3, "%d-%b-%Y")

d4 <- c("August 19 (2015)", "July 1 (2015)")
parse_date(d4, "%B %d (%Y)")

d5 <- "12/30/14" # Dec 30, 2014
parse_date(d5, "%m/%d/%y")

t1 <- "1705"
parse_time(t1, "%H%M")

t2 <- "11:15:10.12 PM"
parse_time(t2, "%I:%M:%M.%S %p")
```


--------------------------------------------------------------------------------
# 11.4 Parsing a file

## RESUME HERE

