---
title: "Ch12_lab"
author: "Holden Jones"
date: '2022-09-02'
output: html_document
---

# Tidy data


--------------------------------------------------------------------------------
# 12.1 Introduction

we'll learn a consistent way to organise data in R - called tidy data
more info in Tidy Data paper http://www.jstatsoft.org/v59/i10/paper

## 12.1.1

tidyr is in tidyverse
```{r}
library(tidyverse)
```


--------------------------------------------------------------------------------
# 12.2 Tidy data

lots of ways to represent data in table form - but only the tidy dataset will be
best to work with in tidyverse! muah ha ha!

Rules for a tidy dataset:
1. Each variable much have its own column
2. Each observation must have its own row
3. Each value must have its own cell

even simpler set of instructions:
1. Put each dataset in a tibble
2. Put each variable in a column

Two main advantages for ensuring data is tidy:
1. General advantage for picking one consistent way of storing data. Learn the
tools that work for that data structure and stick with em!
2. Specific advantage to sticking variables in columns because it allows R's
vectorized nature to shine. Most built-in R functions work with vectors
of values, making transforming tidy data feel particularly natural.

## 12.2.1 Exercises

1. Using prose, describe how the variables and observations are organised in each of the sample tables.
*1 - each variable has a column, each observation has a row*
*2 - case and population should have their own columns, each observation has multiple rows - ex. Brazil 1999 should just be one year*
*3 - rate is really a combo of two variables but each observation is in its own row which is a start I guess!*
*4 - easier to have these in the same table*

2. Compute the rate for table2, and table4a + table4b. You will need to perform four operations:

- Extract the number of TB cases per country per year.
- Extract the matching population per country per year.
- Divide cases by population, and multiply by 10000.
- Store back in the appropriate place.
- Which representation is easiest to work with? Which is hardest? Why?
```{r}
# calculate rate for table 2
table2a <- table2 %>%
  group_by(country, year) %>%
  mutate(rate = count[type == 'cases'] / count[type == 'population']
  )
# this provides duplicate rates - 1 each for cases and population rows

# calculate rate for table4a + table4b
# from answers
table4c <-
  tibble(
    country = table4a$country,
    `1999` = table4a[["1999"]] / table4b[["1999"]] * 10000,
    `2000` = table4a[["2000"]] / table4b[["2000"]] * 10000
  )

table4c
# I like this approach - build out a 3rd dataframe instead of manipulating existing ones
```

3. Recreate the plot showing change in cases over time using table2 instead of table1. What do you need to do first?
*1st need to remove population rows, filter them out and use a smaller df*
```{r}
# ggplot code using tidy table1
ggplot(table1, aes(year, cases)) + 
  geom_line(aes(group = country), colour = "grey50") + 
  geom_point(aes(colour = country))

# filter out pop, then plot
table2 %>%
  filter(type == 'cases') %>%
  ggplot(aes(year, count)) +
  geom_line(aes(group = country), color = "grey49") +
  geom_point(aes(color = country)) +
  labs(y = 'cases')
```


--------------------------------------------------------------------------------
# 12.3 Pivoting

Unfortunately most data is untidy

Tidying - first steps: what are the observations and variables? Then, resolve one
of two common problems.
- one variable might be spread across multiple columns
- one observation might be scattered across multiple rows
to fix these use two most important tidyr functions: pivot_longer() and pivot_wider()

## 12.3.1 Longer

example, table4a each row has two observations (1 for each year), not one! 

to tidy, need to pivot offending columns into a new pair of variables, we need
three parameters to do so
- set of columns whose names are values not variables
- name of variable to move column names to - in above example it is year
- name of variable to move column name values to - in above example it is cases
*surround non-syntactic column names in `` - backticks!*
```{r}
table4a_long <- table4a %>% 
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "cases")
# wow this code is pretty straightforward!

# and same thing with table4b
table4b_long <- table4b %>%
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "population")

left_join(table4a_long, table4b_long)
```

pivet_longer() makes datasets longer by reducing number of columns

## 12.3.2 Wider

pivot_wider() is opposite of pivot_longer() - use when an observation is scattered across rows

example, table 2 - each country and year is spread across two rows
to tidy this up, first analyse representation similar to pivot_longer(), but only
need two parameters:
- column to take variable names from - here it's type
- column to take values from - here it's count
```{r}
table2 %>%
  pivot_wider(names_from = year, values_from = count)
# columns not in "" b/c already in dataframe
```

pivot_wider() makes long tables shorter and wider, it is a 
compliment to pivot_longer()

## 12.3.3 Exercises

1. Why are pivot_longer() and pivot_wider() not perfectly symmetrical?
Carefully consider the following example:
(Hint: look at the variable types and think about column names.)
pivot_longer() has a names_ptypes argument, e.g.  names_ptypes = list(year = double()). What does it do?
*not symmetrical b/c pivot_longer() creates new columns and pivot_wider() uses columns that are already created and replaces them*
*hmm, also originally variables are in double form, and after converted back from pivot_longer, year is in chr form*

from answers:
*pivot_wider() loses column type information - always makes them into characters*
*use the names_transform argument to pivot_longer(), which provides a function to coerce the column to a different data type*
```{r}
stocks <- tibble(
  year   = c(2015, 2015, 2016, 2016),
  half  = c(   1,    2,     1,    2),
  return = c(1.88, 0.59, 0.92, 0.17)
)

# this code shows that the two are reversible tho! except for variable types! can see how that is important
stocks %>% 
  pivot_wider(names_from = year, values_from = return) %>% 
  pivot_longer(`2015`:`2016`, names_to = "year", values_to = "return")
```

2. Why does this code fail?
*the year columns aren't recognized b/c they are numerical and not common column names - need ``*
*in this case, pivot_longer trying to find the 1999th and 2000th columns, which obvi don't exist in this small df!*
```{r}
# error in loc_validate() - can't find columns
#table4a %>% 
  #pivot_longer(c(1999, 2000), names_to = "year", values_to = "cases")

# appropriately define numerical columns
table4a %>% 
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "cases")
```

3. What would happen if you widen this table? Why? How could you add a new column to uniquely identify each value?
*issue likely from there being multiple entries for age per person - yep, warning: values not uniquely identified, output contains list-cols*
*with unique combo of name and obs, we can do pivot_wider()! Had to add this column in*
```{r}
# woah, interesting formatting here!
people <- tribble(
  ~name,             ~names,  ~values,
  #-----------------|--------|------
  "Phillip Woods",   "age",       45,
  "Phillip Woods",   "height",   186,
  "Phillip Woods",   "age",       50,
  "Jessica Cordero", "age",       37,
  "Jessica Cordero", "height",   156
)

# pivot_wider on og
people %>% 
  pivot_wider(names_from = names, values_from = values)

# from answers - add row with distinct observation count for each combo of name and key
people2 <- people %>%
  group_by(name, names) %>%
  mutate(obs = row_number())
people2

# pivot_wider on people2
people2 %>% 
  pivot_wider(names_from = names, values_from = values)
```

4. Tidy the simple tibble below. Do you need to make it wider or longer? What are the variables?
*variables are sex, and if pregnant or not, need to make longer*
```{r}
preg <- tribble(
  ~pregnant, ~male, ~female,
  "yes",     NA,    10,
  "no",      20,    12
)

preg %>%
  pivot_longer(c(male,female), names_to = "sex", values_to = "number", 
               values_drop_na = TRUE) # can drop male preg row
# this is an example of turning explicit missing value to implicit missing
```


--------------------------------------------------------------------------------
# 12.4. Separating and uniting

Haven't learned how to tidy table3 yet which has a different problem - one column 
(rate) has two variables (cases and population). We need to use the separate 
function(), unite() is the complement of separate()

## 12.4.1 Separate

separate() pulls apart one column into multiple columns, splitting whenever a 
separator appears - default is wheneber non alphanumeric character appears
```{r}
table3 %>% 
  separate(rate, into = c("cases", "population"))
# result is a tidy df

# can specify what the separator is
table3 %>% 
  separate(rate, into = c("cases", "population"), sep = "/")

# default new column types are kept from og (in this ex char), can tell separate
# to guess new column types by using convert = TRUE
```

can also pass a vector of integers to sep. separate() will interpret the integers 
as positions to split at. Positive values start at 1 on the far-left of the strings;
negative value start at -1 on the far-right of the strings. When using integers to
separate strings, the length of sep should be one less than the number of names 
in into - in this example makes df nontidy, but good to understand ability 
```{r}
table3 %>% 
  separate(year, into = c("century", "year"), sep = 2)
```

## 12.4.2 Unite

unite() is the inverse of separate(): it combines multiple columns into a single 
column. You’ll need it much less frequently than separate(), but it’s still a 
useful tool to have in your back pocket.

use unite() to join century and year columns from last example (saved as table5)
```{r}
table5 %>% 
  unite(new, century, year)
# interesting, default places _ between combined values

# here specify sep as "" ie nothing
table5 %>% 
  unite(new, century, year, sep = "")
```

## 12.4.3 Exercises

1. What do the extra and fill arguments do in separate()?
Experiment with the various options for the following two
toy datasets.
*extra - if sep is char vector, this arg controls what happens when too many pieces, three options*
*fill - if sep is char vector, this arg controls what happens when not enough pieces, also three options*
```{r}
# extra
tibble(x = c("a,b,c", "d,e,f,g", "h,i,j")) %>% 
  separate(x, c("one", "two", "three"))
# default sep was by , default was to drop g

tibble(x = c("a,b,c", "d,e,f,g", "h,i,j")) %>%
  separate(x, c("one", "two", "three"), extra = "merge")
# only "merge" retains g

# fill
tibble(x = c("a,b,c", "d,e", "f,g,i")) %>% 
  separate(x, c("one", "two", "three"))
# default is to put NA in

tibble(x = c("a,b,c", "d,e", "f,g,i")) %>% 
  separate(x, c("one", "two", "three"), fill = "left")
# this code fills the NA on the left
```

2. Both unite() and separate() have a remove argument. What
does it do? Why would you set it to FALSE?
*remove input column from output df - perhaps that info is also unique / important*

3. Compare and contrast separate() and extract(). Why are
there three variations of separation (by position, by
separator, and with groups), but only one unite?
*extract is more flexible - uses regular expression to specify groups in character vector and split that vector into multiple columns* - answers
```{r}
# example with separators
tibble(x = c("X_1", "X_2", "AA_1", "AA_2")) %>%
  extract(x, c("variable", "id"), regex = "([A-Z])_([0-9])")

# example that separate could not parse
tibble(x = c("X1", "X20", "AA11", "AA2")) %>%
  extract(x, c("variable", "id"), regex = "([A-Z]+)([0-9]+)")
```

# 12.5 Missing values

data can be explicitly or implicitly missing (NA or just absent)

One way to think about the difference is with this Zen-like koan: An explicit 
missing value is the presence of an absence; an implicit missing value is the 
absence of a presence.

ex:
```{r}
stocks <- tibble(
  year   = c(2015, 2015, 2015, 2015, 2016, 2016, 2016),
  qtr    = c(   1,    2,    3,    4,    2,    3,    4),
  return = c(1.88, 0.59, 0.35,   NA, 0.92, 0.17, 2.66)
)

# can pivot to make implicit missing values explicit
stocks %>% 
  pivot_wider(names_from = year, values_from = return)

# can turn back to longer, make explicit missings back to implicit
# with values_drop_na
stocks %>% 
  pivot_wider(names_from = year, values_from = return) %>% 
  pivot_longer(
    cols = c(`2015`, `2016`), 
    names_to = "year", 
    values_to = "return", 
    values_drop_na = TRUE
  )
```

Another important tool for making missing values explicit in tidy data is complete():
complete() takes all unique combos of sets of columns and fills in missing ones
with NAs

Also fill if previous entries need to be filled in
```{r}
# complete()
stocks %>% 
  complete(year, qtr)

# fill()
treatment <- tribble(
  ~ person,           ~ treatment, ~response,
  "Derrick Whitmore", 1,           7,
  NA,                 2,           10,
  NA,                 3,           9,
  "Katherine Burke",  1,           4
)

treatment %>% 
  fill(person)
```

## 12.5.1 Exercises

1. Compare and contrast the fill arguments to pivot_wider() and complete().
*fill seems pretty simple, pivot_wider()
*pivot_longer creates more rows, separates out observations if multiple stored in single column, can make explicit missing data implicit*
*pivot_wider creates more columns, separating out variables if one observation spread across multiple rows, makes implicit missing data explicit*
*fill fills in missing data with actual data*

2. What does the .direction argument to fill() do?
*fill based on should values be copied up or down?*
```{r}
treatment %>% 
  fill(person, .direction = "down")

treatment %>% 
  fill(person, .direction = "up")
```


# 12.6 Case Study

The tidyr::who dataset contains tuberculosis (TB) cases broken down by year, 
country, age, gender, and diagnosis method. The data comes from the 2014 World 
Health Organization Global Tuberculosis Report, available at
http://www.who.int/tb/country/data/download/en/.

who is typical dataset - messy!
*1st thoughts - pivot_wider() for year, what are all these columns? iso2 and iso3 seem redundant*
```{r}
who
```

To start, gather columns that are not variables
- seems like country, that's an observation!

## RESUME HERE









