---
title: "Ch14_lab"
author: "Holden Jones"
date: '2022-09-28'
output: html_document
---

# Strings


--------------------------------------------------------------------------------
# 14.1 Introduction

intro to string manipulation. focus is on regular expressions
regexps are concise language for describing patterns in strings - they look crazy
  at first!
  
## 14.1.1 Prerequisites
```{r}
library(tidyverse)
```


--------------------------------------------------------------------------------
# 14.2 String basics

can create strings with "" or '', usually do "" unless need multiple quotes
```{r}
string1 <- "This is a string"
string2 <- 'If I want to include a "quote" inside a string, I use single quotes'
```

have to be really specific with quotes within strings, and backslashes

use writeLines() to see the actual string contents

"\n" for newline and "\n" for tab

## 14.2.1 String length

use str_length
```{r}
str_length(c("a", "R for data science", NA))
```

## 14.2.2 Combining strings

use str_c
```{r}
str_c("x", "y")

str_c("x", "y", "z")
```

missing values contagious, use str_replace_na(x) if want to print as NA

objects of length 0 silently dropped, useful with if
```{r}
name <- "Holden"
time_of_day <- "night"
birthday <- FALSE

str_c(
  "Good ", time_of_day, " ", name,
  if (birthday) " and HAPPY BIRTHDAY",
  "."
)
```

collapse to make a vector a single string
```{r}
str_c(c("x", "y", "z"), collapse = ", ")
```

## 14.2.3 Subsetting strings

## RESUME HERE