---
title: "Ch15_lab"
author: "Holden Jones"
date: '2022-10-27'
output: html_document
---

# Factors


--------------------------------------------------------------------------------
# 15.1 Introduction

factors used to work with categorical variables, factors easier to work with so
base R automatically converts categorical variables to factors, but not in tidy

## 15.1.1 Prerequisites

use the forcats package
```{r}
library(tidyverse)
```

## 15.1.2 Learning more

Read Amelia McNamara and Nicholas Horton's paper; Wrangling categorical data in R


--------------------------------------------------------------------------------
# 15.2 Creating factors

Imagine that you have a variable that records month:
```{r}
x1 <- c("Dec", "Apr", "Jan", "Mar")

# Using a string to record this variable has two problems:
# only 12 months, and can make typos
x2 <- c("Dec", "Apr", "Jam", "Mar")

# and doesn't sort in good way:
sort(x1)
```

You can fix both of these problems with a factor. To create a factor you must 
start by creating a list of the valid levels:
```{r}
month_levels <- c(
  "Jan", "Feb", "Mar", "Apr", "May", "Jun", 
  "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"
)

# now can create a factor:
y1 <- factor(x1, levels = month_levels)
y1
sort(y1)

# any values not in levels set will silently become NA:
y2 <- factor(x2, levels = month_levels)
y2

# if we omit levels, will be taken from data in alphabetical order:
factor(x1)

# sometimes prefer that order of levels matches order of first appearence:
# use unique() or after the fact with fct_inorder():
f1 <- factor(x1, levels = unique(x1))
f1

f2 <- x1 %>% factor() %>% fct_inorder()
f2

# if need to access levels, can do so with levels():
levels(f2)
```


--------------------------------------------------------------------------------
# 15.3 General Social Survey

focusing rest of chapter on data from GSS, U of Chicago survey
```{r}
gss_cat

# quick view of different variables with count:
gss_cat %>%
  count(race)

gss_cat  %>%
  count(relig)

# or bar chart:
ggplot(gss_cat, aes(race)) +
  geom_bar()

ggplot(gss_cat, aes(relig)) +
  geom_bar()

# by default ggplot will drop levels without any values, can force them to show:
ggplot(gss_cat, aes(race)) +
  geom_bar() +
  scale_x_discrete(drop = FALSE)
```

## 15.3.1 Exercises

1. Explore the distribution of rincome (reported income). What makes the default 
  bar chart hard to understand? How could you improve the plot?
```{r}
# default includes lots of things like don't know, NA, and the ordering is off
gss_cat %>%
  ggplot(aes(rincome)) +
   geom_bar()

# can remove non numerical cases
gss_cat %>%
  filter(!rincome == 'Not applicable',
         !rincome == 'No answer',
         !rincome == "Don't know",
         !rincome == 'Refused') %>%
  ggplot(aes(rincome)) +
   geom_bar() +
   scale_x_discrete(limits = rev(levels(gss_cat$rincome)), drop = TRUE)

# also can reorder x axes to be in descending order
gss_cat %>%
  ggplot(aes(rincome)) +
   geom_bar() +
   scale_x_discrete(limits = rev(levels(gss_cat$rincome)), drop = FALSE)
```

2. What is the most common relig in this survey? What’s the most common partyid?
```{r}
# Protestant most common relig
gss_cat %>%
  count(relig) %>%
  arrange(desc(n)) %>%
  head(1)

# Independent most common partyid
gss_cat %>%
  count(partyid) %>%
  arrange(desc(n)) %>%
  head(1)
```

3. Which relig does denom (denomination) apply to? How can you find out with a 
  table? How can you find out with a visualisation?
*answer sheet*
```{r}
# filter out no answer responses to denomination, all remaining are Protestant
gss_cat %>%
  filter(!denom %in% c(
    "No answer", "Other", "Don't know", "Not applicable",
    "No denomination"
  )) %>%
  count(relig)

# also clear in scatterplot
gss_cat %>%
  count(relig, denom) %>%
  ggplot(aes(x = relig, y = denom, size = n)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90))
```


--------------------------------------------------------------------------------
# 15.4 Modifying factor order

take the same plot but with and without reordered levels of relig
```{r}
# default:
relig_summary <- gss_cat %>%
  group_by(relig) %>%
  summarise(
    age = mean(age, na.rm = TRUE),
    tvhours = mean(tvhours, na.rm = TRUE),
    n = n()
  )

ggplot(relig_summary, aes(tvhours, relig)) + geom_point()

# with reordered:
ggplot(relig_summary, aes(tvhours, fct_reorder(relig, tvhours))) +
  geom_point()
```

fct_reorder() takes 3 arguments:
- f, factor whose levels to modify
- x, numeric vector used to reorder levels
- fun, optional function used if multiple values of x for each f, default median

as transformations get more complicated, better to move them into mutate()
  instead of aes()
```{r}
# same plot as above:
relig_summary %>%
  mutate(relig = fct_reorder(relig, tvhours)) %>%
  ggplot(aes(tvhours, relig)) +
    geom_point()
```

similar plot, how does avg age vary across reported income level?
```{r}
rincome_summary <- gss_cat %>%
  group_by(rincome) %>%
  summarise(
    age = mean(age, na.rm = TRUE),
    tvhours = mean(tvhours, na.rm = TRUE),
    n = n()
  )

ggplot(rincome_summary, aes(age, fct_reorder(rincome, age))) + geom_point()
```

here shouldn't reorder x axis b/c rincome has principled order that shouldn't 
  be changed
*only use fct_reorder() for factors whose levels are arbitrarily ordered*
but it does make sense to reorder y axis a bit
```{r}
ggplot(rincome_summary, aes(age, fct_relevel(rincome, "Not applicable"))) +
  geom_point()
```

reordering also useful when colouring lines on plot
fct_reorder2 reorders factor by y values associated with highest x values
  means that line colours line up with legend!
```{r}
by_age <- gss_cat %>%
  filter(!is.na(age)) %>%
  count(age, marital) %>%
  group_by(age) %>%
  mutate(prop = n / sum(n))

ggplot(by_age, aes(age, prop, colour = marital)) +
  geom_line(na.rm = TRUE)

ggplot(by_age, aes(age, prop, colour = fct_reorder2(marital, age, prop))) +
  geom_line() +
  labs(colour = "marital")
```

for barplots can use fct_infreq() to order levels in increasing freq
may want to combine with fct_rev()
```{r}
gss_cat %>%
  mutate(marital = marital %>% fct_infreq() %>% fct_rev()) %>%
  ggplot(aes(marital)) +
    geom_bar()
```

## 15.4.1 Exercises

1. There are some suspiciously high numbers in tvhours. Is the mean a 
  good summary?
*well it doesn't seem like they're typos as there are multiples*
*median shows them all as the same, so tend to stick with mean here*
```{r}
gss_cat %>%
  count(tvhours)

# tv hours by relig
relig_summary <- gss_cat %>%
  group_by(relig) %>%
  summarise(
    age = mean(age, na.rm = TRUE),
    tvhours_med = median(tvhours, na.rm = TRUE),
    n = n()
  )

# using median shows the majority at 2 hours despite the means being v diff.
ggplot(relig_summary, aes(tvhours_med, fct_reorder(relig, tvhours_med))) +
  geom_point()
```

2. For each factor in gss_cat identify whether the order of the levels is       arbitrary or principled.
*principled: year, age, rincome, rest are arbitrary*

3. Why did moving “Not applicable” to the front of the levels move it to 
  the bottom of the plot?
*b/c gives the 'Not applicable' factor an integer value of 1*


--------------------------------------------------------------------------------
# 15.5 Modifying factor levels

More powerful than changing orders of levels is changing their values
can clarify labels for publication, and collapse levels for high-level display
use fct_recode() for this:
```{r}
# partid levels terse and inconsistent
gss_cat %>%
  count(partyid)

# tweak to be longer and use parallel construction
gss_cat %>%
  mutate(partyid = fct_recode(partyid,
    "Republican, strong"    = "Strong republican",
    "Republican, weak"      = "Not str republican",
    "Independent, near rep" = "Ind,near rep",
    "Independent, near dem" = "Ind,near dem",
    "Democrat, weak"        = "Not str democrat",
    "Democrat, strong"      = "Strong democrat"
  )) %>%
  count(partyid)
```

to combine groups, can assign multiple old levels to same new level:
```{r}
gss_cat %>%
  mutate(partyid = fct_recode(partyid,
    "Republican, strong"    = "Strong republican",
    "Republican, weak"      = "Not str republican",
    "Independent, near rep" = "Ind,near rep",
    "Independent, near dem" = "Ind,near dem",
    "Democrat, weak"        = "Not str democrat",
    "Democrat, strong"      = "Strong democrat",
    "Other"                 = "No answer",
    "Other"                 = "Don't know",
    "Other"                 = "Other party"
  )) %>%
  count(partyid)
```

have to be careful with this! fct_collapse() to provide vector of old levels:
```{r}
gss_cat %>%
  mutate(partyid = fct_collapse(partyid,
    other = c("No answer", "Don't know", "Other party"),
    rep = c("Strong republican", "Not str republican"),
    ind = c("Ind,near rep", "Independent", "Ind,near dem"),
    dem = c("Not str democrat", "Strong democrat")
  )) %>%
  count(partyid)
```

sometimes lump together all small groups to make plot or table simpler, 
  use fct_lump():
```{r}
gss_cat %>%
  mutate(relig = fct_lump(relig)) %>%
  count(relig)
```

we probably over collapsed here, instead can use n to specify how many groups:
```{r}
gss_cat %>%
  mutate(relig = fct_lump(relig, n = 10)) %>%
  count(relig)
```

## 15.5.1 Exercises

1. How have the proportions of people identifying as Democrat, Republican, 
  and Independent changed over time?
*ind increase, dem constant, rep decrease*
```{r}
gss_cat %>%
  mutate(partyid = fct_collapse(partyid,
    other = c("No answer", "Don't know", "Other party"),
    rep = c("Strong republican", "Not str republican"),
    ind = c("Ind,near rep", "Independent", "Ind,near dem"),
    dem = c("Not str democrat", "Strong democrat")
  )) %>%
  filter(!partyid == 'other') %>%
  count(year, partyid) %>%
  group_by(year) %>%
  mutate(prop = n / sum(n)) %>% 
  ggplot(aes(year, prop, color = fct_reorder2(partyid, year, prop))) +
  geom_line() +
  labs(color = "partyid")
```

2. How could you collapse rincome into a small set of categories?
```{r}
gss_cat %>%
  mutate(rincome = fct_collapse(rincome,
    other = c("Refused", "Don't know", "No answer", "Not applicable"),
    'less than $10000' = c("$8000 to 9999", "$7000 to 7999", "$6000 to 6999",
                           "$5000 to 5999", "$4000 to 4999", "$3000 to 3999",
                           "$1000 to 2999", "Lt $1000")
  )) %>%
  count(rincome)
```



