---
title: "Ch16_lab"
author: "Holden Jones"
date: '2022-10-31'
output: html_document
---

# Dates and times


--------------------------------------------------------------------------------
# 16.1 Introduction

dates and times hard b/c have to reconcile physical with geopolitical things

## 16.1.1 Prerequisites

lubridate package
```{r}
library(tidyverse)
library(lubridate)
library(nycflights13)
```


--------------------------------------------------------------------------------
# 16.2 Creating date/times

date, time and date-time(<dttm>)
use hms package to store times

three ways likely to create a date/time:
- string
- individual date-time components
- from existing date/time object

## 16.2.1 From strings

use lubridate functions, specify order / format of date from string
```{r}
ymd("2017-01-31")
mdy("January 31st, 2017")
dmy("31-Jan-2017")
```

also take unquoted numbers which is usually best/easiest

ymd() with _ to create datetimes
```{r}
ymd("2017-01-31")
ymd_hms("2017-01-31 20:11:59")
```

## 16.2.2 From individual components

sometimes will have ind componets of date-time spread across multiple columns
```{r}
# ex. flights data
flights %>% 
  select(year, month, day, hour, minute)
```

use make_date() for dates, make_datetime() for date-times:
```{r}
flights %>% 
  select(year, month, day, hour, minute) %>% 
  mutate(departure = make_datetime(year, month, day, hour, minute))

# can do same thing for each of 4 time columns in flights, use modulus:
make_datetime_100 <- function(year, month, day, time) {
  make_datetime(year, month, day, time %/% 100, time %% 100)
}

flights_dt <- flights %>% 
  filter(!is.na(dep_time), !is.na(arr_time)) %>% 
  mutate(
    dep_time = make_datetime_100(year, month, day, dep_time),
    arr_time = make_datetime_100(year, month, day, arr_time),
    sched_dep_time = make_datetime_100(year, month, day, sched_dep_time),
    sched_arr_time = make_datetime_100(year, month, day, sched_arr_time)
  ) %>% 
  select(origin, dest, ends_with("delay"), ends_with("time"))

flights_dt

# and can visualize the distribution of departure times across year
flights_dt %>% 
  ggplot(aes(dep_time)) + 
  geom_freqpoly(binwidth = 86400) # 86400 seconds = 1 day

# or single day
flights_dt %>% 
  filter(dep_time < ymd(20130102)) %>% 
  ggplot(aes(dep_time)) + 
  geom_freqpoly(binwidth = 600) # 600 s = 10 minutes
```

when use date-times in numeric context (like histogram), 1 means 1 second, so
bindwidth of 86400 equals one day. for dates, 1 means 1 day

## 16.2.3 From other types

May want to switch between date-time and date. use as_datetime() and as_date()

## 16.2.4 Exercises

1. What happens if you parse a string that contains invalid dates?
*fail to parse the invalid dates but parse the rest*
```{r}
x <- ymd(c("2010-10-10", "bananas"))
x
```

2. What does the tzone argument to today() do? Why is it important?
*assigns time zone, obvi important b/c date can change*

3. Use the appropriate lubridate function to parse each of the
following dates:
```{r}
d1 <- "January 1, 2010"
mdy(d1)

d2 <- "2015-Mar-07"
ymd(d2)

d3 <- "06-Jun-2017"
dmy(d3)

d4 <- c("August 19 (2015)", "July 1 (2015)")
mdy(d4)

d5 <- "12/30/14"
mdy(d5)
```


--------------------------------------------------------------------------------
# 16.3 Date-time components

focus on accessor functions to get and set individual components

## 16.3.1 Getting components

can pull out individual parts of date with accessor functions
```{r}
datetime <- ymd_hms("2016-07-08 12:34:56")

year(datetime)
month(datetime)
mday(datetime)
yday(datetime)
wday(datetime)

# for month() and wday() can set label as true
month(datetime, label = TRUE)
wday(datetime, label = TRUE)
```

use wday() to see that more flights depart during week than on weekend
```{r}
flights_dt %>% 
  mutate(wday = wday(dep_time, label = TRUE)) %>%
  filter(!is.na(air_time)) %>%
  ggplot(aes(x = wday)) +
    geom_bar()
```

can look at average departure delay by minute within the hour:
```{r}
flights_dt %>% 
  mutate(minute = minute(dep_time)) %>% 
  group_by(minute) %>% 
  summarise(
    avg_delay = mean(arr_delay, na.rm = TRUE),
    n = n()) %>% 
  ggplot(aes(minute, avg_delay)) +
    geom_line()
# interesting that there are peaks and values of this throughout the hour

# but if we look at scheduled departure we don't see this pattern:
sched_dep <- flights_dt %>% 
  mutate(minute = minute(sched_dep_time)) %>% 
  group_by(minute) %>% 
  summarise(
    avg_delay = mean(arr_delay, na.rm = TRUE),
    n = n())

ggplot(sched_dep, aes(minute, avg_delay)) +
  geom_line()
```

interesting, reason for discrepency b/ween scheduled and departure time due to 
human judgement - preference for 'nice' departure times
```{r}
ggplot(sched_dep, aes(minute, n)) +
  geom_line()
```

## 16.3.2 Rounding

instead of plotting individual components, can round to nearby unit of time
can use floord_date(), round_date(), ceiling_date()
```{r}
# with week can plot number of flights per week:
flights_dt %>% 
  count(week = floor_date(dep_time, "week")) %>% 
  ggplot(aes(week, n)) +
    geom_line()

# but we can also do this with just week() from lubridate: notice difference in 
# plots, why is this?
flights_dt %>% 
  count(week = week(dep_time)) %>% 
  ggplot(aes(week, n)) +
    geom_line()
```

computing difference between rounded and unrounded date can be useful

## 16.3.3 Setting components

can use each accessor function to set components of date/time and update them
```{r}
(datetime <- ymd_hms("2016-07-08 12:34:56")) # () to print result

(year(datetime) <- 2020)

(month(datetime) <- 01)

(hour(datetime) <- hour(datetime) + 1)

(update(datetime, year = 2020, month = 2, mday = 2, hour = 2))
```

values roll over if too big:
```{r}
ymd("2015-02-01") %>% 
  update(mday = 30)

ymd("2015-02-01") %>% 
  update(hour = 400)
```

use update() to show dist. of flights across course every day of the year
```{r}
flights_dt %>% 
  mutate(dep_hour = update(dep_time, yday = 1)) %>% # make day = 1 for all
  ggplot(aes(dep_hour)) +
    geom_freqpoly(binwidth = 300)
```

setting larger components of a date constant is a powerful way to explore
techniques in smaller components

## 16.3.4 Exercises

1. How does the distribution of flight times within a day change over the 
  course of the year?
*group_by month, recreate prior figure for each month of year*
*no big difference across months*
```{r}
flights_dt %>% 
  group_by(month(dep_time)) %>%
  mutate(dep_hour = update(dep_time, yday = 1)) %>%
  mutate(month = factor(month(dep_time))) %>%
  ggplot(aes(dep_hour)) +
  geom_freqpoly(aes(group = month, colour = month)) +
  theme(axis.text.x = element_blank())
```

2. Compare dep_time, sched_dep_time and dep_delay. Are they consistent? 
  Explain your findings.
*wow this is funky! not sure if my plot / code is doing what I think it is*
```{r}
flights_dt %>%
  filter(!is.na(air_time)) %>%
  mutate(dep_time = update(dep_time, yday = 1),
         sched_dep_time = update(sched_dep_time, yday = 1)) %>%
  ggplot(aes(sched_dep_time, dep_time, color = dep_delay)) +
  geom_point()
# funky! so this shows that vast vast majority of sched and actual dep time are
# closely correlated, but there's a strange group of outliers below, these are
# flights delayed from the previous day whose departure was pushed back a day!

# answer sheet took it a different way:
flights_dt %>%
  mutate(dep_time_ = sched_dep_time + dep_delay * 60) %>%
  filter(dep_time_ != dep_time) %>%
  select(dep_time_, dep_time, sched_dep_time, dep_delay)
# these discrepencies are the outliers in my above figure
```

3. Compare air_time with the duration between the departure and arrival. 
  Explain your findings. (Hint: consider the location of the airport.)
*expect that for flights b/ween airports in same time zone, air_time will equal*
*arrival - departure, how to control for different time zones?*
```{r}
# currently using abs value of arr_time - dep_time, and <1000 dif so not long
flights_dt %>%
  mutate(arr_minus_dep = abs(arr_time - dep_time)) %>%
  filter(arr_minus_dep < 1000) %>%
  ggplot(aes(air_time, arr_minus_dep)) +
    geom_point()
# interesting! trend goes up, but then seems to reset regularl intervals

# from answer sheet:
flights_dt %>%
  mutate(
    flight_duration = as.numeric(arr_time - dep_time),
    air_time_mins = air_time,
    diff = flight_duration - air_time_mins
  ) %>%
  select(origin, dest, flight_duration, air_time_mins, diff) %>%
  ggplot(aes(diff)) +
  geom_histogram()
```

4. How does the average delay time change over the course of a day? Should 
  you use dep_time or sched_dep_time? Why?
*def use sched_dep_time b/c that's what's relevant for delays*
*need to control for late flights getting delayed into the next day, makes
```{r}
# going to look at diff between dep_time and sched_dep_time for delay
flights_dt %>% 
  mutate(sched_dep_hour = hour(update(sched_dep_time, yday = 1)),
         dep_delay = as.numeric(abs(dep_time - sched_dep_time))) %>%
  group_by(sched_dep_hour) %>%
  mutate(avg_dep_delay = mean(dep_delay)) %>%
  ggplot(aes(sched_dep_hour, avg_dep_delay)) +
    geom_point(aes())
# generally the later it gets the larger the dep_delay is, late flights super
# likely to get delayed to the next day which makes the graph look crazy

# need to find a way to control for changes in the day; ie 23:30 - 00:30
# being a delay of 60 min instead of 23 hours

# from answer sheet:
flights_dt %>%
  mutate(sched_dep_hour = hour(sched_dep_time)) %>%
  group_by(sched_dep_hour) %>%
  summarise(dep_delay = mean(dep_delay)) %>%
  ggplot(aes(y = dep_delay, x = sched_dep_hour)) +
  geom_point() +
  geom_smooth()
# lesson here is always review the df!!! there was already a delay column
# that had differences in the day accounted for!!!! doophus!
```

5. On what day of the week should you leave if you want to minimise the chance 
  of a delay?
*fly on saturdays boys*
```{r}
flights_dt %>%
  mutate(weekday = weekdays(sched_dep_time)) %>%
  group_by(weekday) %>%
  ggplot(aes(weekday, arr_delay)) +
  geom_boxplot()
# can't really discern any difference between weekdays visually

# try with just compare means
flights_dt %>%
  mutate(weekday = weekdays(sched_dep_time)) %>%
  group_by(weekday) %>%
  summarize(mean(arr_delay, na.rm = TRUE))
# this shows that Saturdays have the lowest mean arr_delay
# actually mean is ahead of sched by one minute!
```

6. What makes the distribution of diamonds$carat and flights$sched_dep_time
  similar?
*I think human preference explains the similarity in both dist*
*flights more likely to be scheduled and diamonds more likely to be graded*
*at round numbers*
```{r}
diamonds %>%
  ggplot(aes(carat)) +
  geom_histogram(binwidth = 0.05)

flights %>%
  ggplot(aes(sched_dep_time)) +
  geom_histogram(binwidth = 10)
```

7. Confirm my hypothesis that the early departures of flights in minutes 20-30
  and 50-60 are caused by scheduled flights that leave early. Hint: create a
  binary variable that tells you whether or not a flight was delayed.
```{r}
# I misread the question, looked for flights leaving in those time windows
# as being delayed from earlier
flights %>%
  group_by(arr_delay) %>%
  mutate(status = ifelse(arr_delay > 1, "D", "OT")) %>%
  ggplot(aes(minute, color = status)) +
    geom_freqpoly()
# This doesn't support that hypothesis, identical patterns

# but here's what was actually meant, from answers:
flights_dt %>% 
  mutate(minute = minute(dep_time),
         early = dep_delay < 0) %>% # cool this is how to make binary var.
  group_by(minute) %>% 
  summarise(
    early = mean(early, na.rm = TRUE),
    n = n()) %>% 
  ggplot(aes(minute, early)) +
    geom_line()
```


--------------------------------------------------------------------------------
# 16.4 Time spans

arithmetic with dates, and 3 important classes with time spans:
- durations
- periods
- intervals

## 16.4.1 Durations

subtracting two dates gets a difftime object
```{r}
# How old am I?
(h_age <- today() - ymd(19960710))
# wow haha
```

difftimes are annoying, so use duration from lubridate:
```{r}
as.duration(h_age)

# durations have a bunch of convenient constructors:
dseconds(15)
dminutes(10)
dhours(c(12, 24))
ddays(0:5)
dweeks(3)
dyears(1)
# always recorded in seconds

# can add and multiply durations:
2 * dyears(1)
dyears(1) + dweeks(12) + dhours(15)

# add and subtract durations:
tomorrow <- today() + ddays(1)
last_year <- today() - dyears(1)

# but take care if working across DST
(one_pm <- ymd_hms("2016-03-12 13:00:00", tz = "America/New_York"))
one_pm + ddays(1)
```

## 16.4.2 Periods

use periods to solve problem of working across DST - work with "human" times
```{r}
one_pm
one_pm + days(1)

# like durations, periods can be created with friendly constructors
seconds(15)
minutes(10)
hours(c(12, 24))
days(7)
months(1:6)
weeks(3)
years(1)

# can add and multiply periods
10 * (months(6) + days(1))
days(50) + hours(25) + minutes(2)
```

and of course can add them to dates, periods more likely to do what expected:
```{r}
ymd("2016-01-01") + dyears(1)
ymd("2016-01-01") + years(1)

# Daylight Savings Time
one_pm + ddays(1)
one_pm + days(1)
```

use periods to fix oddity in flight dates: flights that arrived before departed
```{r}
flights_dt %>% 
  filter(arr_time < dep_time)
# these are ovwernight flights

# can fix by adding days(1) to arrival time of each overnight flight:
flights_dt <- flights_dt %>% 
  mutate(
    overnight = arr_time < dep_time, # binary T/F overnight variable!
    arr_time = arr_time + days(overnight * 1), # multiply by value of overnight
    sched_arr_time = sched_arr_time + days(overnight * 1)
  )

# now all flights obey laws of physics!
flights_dt %>% 
  filter(overnight, arr_time < dep_time) 
```

## 16.4.3 Intervals

leap years are tricky, lubridate() will give an estimate, use intervals!
*%--% seems to be the interval symbol!*
```{r}
years(1) / days(1)

# use an interval to get more accurate measurement:
next_year <- today() + years(1)
(today() %--% next_year) / ddays(1)

# integer division to see how many periods fall into an interval:
(today() %--% next_year) %/% days(1)
```

## 16.4.4 Summary

How to pick between duration, periods and intervals?
as always, pick simplest data structure that solves problem
- duration if only care about physical time
- period if care about human times
- interval if need to know how long a span is in human units

# 16.4.5 Exercises

1. Why is there months() but no dmonths()?
*months have different lengths, so can't assign a standard duration*

2. Explain days(overnight * 1) to someone who has just started learning R. 
  How does it work?
*in the flights dataset, overnight is either true (1) or false (0), to make the*
*arrival time match reality, we add a day to the arrival time only when ov is T*

3. Create a vector of dates giving the first day of every month in 2015. Create 
  a vector of dates giving the first day of every month in the current year.
```{r}
x <- mdy("January 1st 2015")
first_2015 <- c(x, x + months(1:11))
first_2015

y <- mdy("January 1st 2022")
first_2022 <- c(y, y + months(1:11))
first_2022
```

4. Write a function that given your birthday (as a date), returns how old you 
  are in years.
```{r}
birthday <- mdy("July 10th, 1996")

abs((today() %--% birthday) / years(1))
```

5. Why can’t (today() %--% (today() + years(1))) / months(1) work?
*use %/% to find number of months in an interval instead of /*
```{r}
(today() %--% (today() + years(1))) / months(1)
```


--------------------------------------------------------------------------------
# 16.5 Time zones

time zones are super complicated
R uses international standard IANA time zones, continent/city

find current time zone
```{r}
Sys.timezone()

# complete list of time zone names:
head(OlsonNames())

# can have same instant of time in different time zones:
(x1 <- ymd_hms("2015-06-01 12:00:00", tz = "America/New_York"))
(x2 <- ymd_hms("2015-06-01 18:00:00", tz = "Europe/Copenhagen"))
(x3 <- ymd_hms("2015-06-02 04:00:00", tz = "Pacific/Auckland"))

x1 - x2
x1 - x3
```

unless otherwise specified, lubridate always uses UTC, no use DST

can change time zone in two ways
- keep instant in time the same, just change how it's displayed
- change underlying instant in time, use when have time zone mislabelled
```{r}
x4 <- c(x1, x2, x3)
x4

# keep instant the same, change display:
x4a <- with_tz(x4, tzone = "Australia/Lord_Howe")
x4a
x4a - x4

# change underlying instant in time:
x4b <- force_tz(x4, tzone = "Australia/Lord_Howe")
x4b
x4b - x4
```