---
title: "Ch18_lab"
author: "Holden Jones"
date: '2022-11-06'
output: html_document
---

# Pipes


--------------------------------------------------------------------------------
# 18.1 Introduction

pipes are powerful tool for clearly expressing sequence of multiple operations

## 18.1.1 Prerequisites

pipe comes from magrittr package, loaded by tidyverse automatically tho
```{r}
library(magrittr)
```


--------------------------------------------------------------------------------
# 18.2 Piping alternatives

point of pipe is to help write code that is easier to read and understand

little bunny Foo Foo example
```{r}
foo_foo <- little_bunny()
```

four ways we could tell the story in code:
  1. save each intermediate step as new object
  2. overwrite original object many times
  3. compose functions
  4. use the pipe
  
here's each approach:

## 18.2.1 Intermediate steps

simplest approach
```{r}
foo_foo_1 <- hop(foo_foo, through = forest)
foo_foo_2 <- scoop(foo_foo_1, up = field_mice)
foo_foo_3 <- bop(foo_foo_2, on = head)
```

downsides:
- code cluttered with unimportant names
- have to carefully increment suffix on each line

don't worry about memory tho, that's a waste of time!

```{r}
diamonds <- ggplot2::diamonds
diamonds2 <- diamonds %>% 
  dplyr::mutate(price_per_carat = price / carat)
```

if columns are shared between dataframes they will be stored as one!
if we modify one then thats no longer the case

## 18.2.2 Overwrite the original

```{r}
foo_foo <- hop(foo_foo, through = forest)
foo_foo <- scoop(foo_foo, up = field_mice)
foo_foo <- bop(foo_foo, on = head)
```

less typing and thinking so less likely to make mistakes, but 2 problems:
- debugging painful! if make mistake need to re-run complete pipeline
- repetition of object being transformed obscures what's changing on each line

## 18.2.3 Function composition

another approach to abandon assignment and string function calls together
```{r}
bop(
  scoop(
    hop(foo_foo, through = forest),
    up = field_mice
  ), 
  on = head
)
```

disadvantage is that have to read from inside-out, right-to-left:
- dagwood sandwhich problem! from Blondie comics

## 18.2.4 Use the pipe

```{r}
foo_foo %>%
  hop(through = forest) %>%
  scoop(up = field_mice) %>%
  bop(on = head)
```

great b/c focuses on verbs, not nouns. set of imperative actions
%>% is pretty intuitive / easy to explain to others

pipe works by performing "lexical transformation" behind scenes
magrittr reassembles code in pipe to form that overwrites intermediate object:
```{r}
my_pipe <- function(.) {
  . <- hop(., through = forest)
  . <- scoop(., up = field_mice)
  bop(., on = head)
}
my_pipe(foo_foo)
```

pipe won't work for two classes of functions:
1. functons that use current envirionment, ex. assign(), get(), load()
2. functions that use lazy evaluation, ex. tryCatch()


--------------------------------------------------------------------------------
# 18.3 When not to use the pipe

powerful but not only tool at disposal
best for writing short linear sequence of operations

use another tool when:
- pipes are longer than (say) ten steps
  - in this case create intermediate objects with meaningful names
    - easier to check results and understand code
- have multiple inputs or outputs
  - if 2 or more objects being transformed, don't use the pipe
- if thinking about directed graph with complicated dependency structure
  - pipes fundamentally linear so complex relationships will yield confusion
  

--------------------------------------------------------------------------------
# 18.4 Other tools from magrittr

%>% automatically loaded in tidyverse, but other magrittr tools can be useful

when working w/ complex pipes, sometimes useful to call function for side effect
- maybe print current object, plot it, or save it to disk
- use "tee" pipe: %T%, returns left-hand side instead of right-hand side
```{r}
rnorm(100) %>%
  matrix(ncol = 2) %>%
  plot() %>%
  str() # structure of above is NULL

rnorm(100) %>%
  matrix(ncol = 2) %T>% # save this so structure at end can use
  plot() %>%
  str()
```

if working w/ functions that don't have data frame based API
-%$% may be useful, it explodes out variables in dataframe so can
  refer to them explicitly
- useful with many functions in baseR
```{r}
mtcars %$%
  cor(disp, mpg)
```

%<>% operator for assignment which allows for following:
```{r}
# instead of:
mtcars <- mtcars %>% 
  transform(cyl = cyl * 2)

# use %<>% for:
mtcars %<>% transform(cyl = cyl * 2)
```

Hadley isn't a fan tho b/c assignment is so important that it's worth
emphasizing!
- I agree!