---
title: "Ch19_lab"
author: "Holden Jones"
date: '2022-12-08'
output: html_document
---

# Functions

going to move a bit quicker through this


--------------------------------------------------------------------------------
# 19.1 Introduction

3 big advantages over copy and paste:
- give a function a name that makes code easier to understand
- as requirements change, only need to update code in one place
- eliminate chance of making incidental mistakes 

## 19.1.1 Prerequisites

base R so no need


--------------------------------------------------------------------------------
# 19.2 When should you write a function?

if copy and pasting a bunch of code more than twice, write a function!

ex. this code randomizes a tibble using same parameters each time
```{r}
df <- tibble::tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)

df$a <- (df$a - min(df$a, na.rm = TRUE)) / 
  (max(df$a, na.rm = TRUE) - min(df$a, na.rm = TRUE))
df$b <- (df$b - min(df$b, na.rm = TRUE)) / 
  (max(df$b, na.rm = TRUE) - min(df$b, na.rm = TRUE)) # og was mistake here!
df$c <- (df$c - min(df$c, na.rm = TRUE)) / 
  (max(df$c, na.rm = TRUE) - min(df$c, na.rm = TRUE))
df$d <- (df$d - min(df$d, na.rm = TRUE)) / 
  (max(df$d, na.rm = TRUE) - min(df$d, na.rm = TRUE))
```

how many inputs does this code have?
- just one for each line

to make inputs more clear, good idea to rewrite code using temp vars
```{r}
x <- df$a
(x - min(x, na.rm = TRUE)) / (max(x, na.rm = TRUE) - min(x, na.rm = TRUE))

# there's duplication here, compute range of data once:
rng <- range(x, na.rm = TRUE)
(x - rng[1]) / (rng[2] - rng[1])
```

after siplifying code and checking that it still works, can turn into function
```{r}
rescale01 <- function(x) {
  rng <- range(x, na.rm = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
}
rescale01(c(0, 5, 10))
```

3 steps to creating new function:
- pick a name for the function, rescale01 b/c rescales b/ween 0 and 1
- list inputs or arguments to function inside function(), just one here, but
  with more, call looks like function(x,y,z)
- place code that have developed in body of function, a {} that immediately
  follows function(...)

easier to start with working code and turn it into a function, harder to create
  function first and then try to make it work
  
good to check function with a few different inputs
```{r}
rescale01(c(-10, 0, 10))
rescale01(c(1, 2, 3, NA, 5))
```

eventually will want to do formal, automated tests, called unit tests
but outside scope of this book

can simplify og code now that we have a function:
```{r}
df$a <- rescale01(df$a)
df$b <- rescale01(df$b)
df$c <- rescale01(df$c)
df$d <- rescale01(df$d)
```

this code easier to understand, but still quite a bit of duplication
will learn to improve this in iteration later

ex.if some var include infinite values, rescale01() will fail
- but can easily and efficiently correct it as a function
```{r}
x <- c(1:10, Inf)

rescale01 <- function(x) {
  rng <- range(x, na.rm = TRUE, finite = TRUE)
  y <- (x - rng[1]) / (rng[2] - rng[1])
  y[y == -Inf] <- 0
  y[y == Inf] <- 1
  y
}

(rescale01(x))
```

do not repeat yourself! DRY! principle: 
less repition = fewer needs to update = fewer bugs and errors

## 19.2.1 Exercises
*all from answers*

1. Why is TRUE not a parameter to rescale01()? What would happen if x 
  contained a single missing value, and na.rm was FALSE?
*b/c even if na.rm is FALSE, will return non-missing value*

2. In the second variant of rescale01(), infinite values are left unchanged.
  Rewrite rescale01() so that -Inf is mapped to 0, and Inf is mapped to 1.
```{r}
rescale01 <- function(x) {
  rng <- range(x, na.rm = TRUE, finite = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
}
```

3. Practice turning the following code snippets into functions. Think about 
  what each function does. What would you call it? How many arguments does it
  need? Can you rewrite it to be more expressive or less duplicative?
```{r}
x <- c(1:10)

# takes proportion of nas in a vector
mean(is.na(x))

# call prop_na:
prop_na <- function(x) {
  mean(is.na(x))
}
prop_na(x)

# look at prop of overall sum for each value in vector
x / sum(x, na.rm = TRUE)

prop_sum <- function(x) {
  x / sum(x, na.rm = TRUE)
}
prop_sum(x)

# calculates sd / mean aka coefficient of variation for a vector
sd(x, na.rm = TRUE) / mean(x, na.rm = TRUE)

coeff_var <- function(x, na.rm = TRUE) {
  sd(x, na.rm = na.rm) / mean(x, na.rm = na.rm)
}
coeff_var(x)
```

4. Write your own functions to compute the variance and skewness of a numeric
  vector.
*from answers*
```{r}
# variance
variance <- function(x, na.rm = TRUE) {
  n <- length(x)
  m <- mean(x, na.rm = TRUE)
  sq_err <- (x - m)^2
  sum(sq_err) / (n - 1)
}

variance(x)
var(x) # check

# skewness
skewness <- function(x, na.rm = FALSE) {
  n <- length(x)
  m <- mean(x, na.rm = na.rm)
  v <- var(x, na.rm = na.rm)
  (sum((x - m) ^ 3) / (n - 2)) / v ^ (3 / 2)
}

skewness(x)
```

5. Write both_na(), a function that takes two vectors of the same length and
  returns the number of positions that have an NA in both vectors.
```{r}
both_na <- function(x, y) {
  sum(is.na(x) & is.na(y))
}

both_na(
  c(NA, NA, 1, 2),
  c(NA, 1, NA, 2)
)

both_na(
  c(NA, NA, 1, 2, NA, NA, 1),
  c(NA, 1, NA, 2, NA, NA, 1)
)
```

6. What do the following functions do? Why are they useful even though they are
  so short?
*always good to make sure directory and file access is in order!*
```{r}
is_directory <- function(x) file.info(x)$isdir
is_readable <- function(x) file.access(x, 4) == 0
```

7. Read the complete lyrics to “Little Bunny Foo Foo”. There’s a lot of
  duplication in this song. Extend the initial piping example to recreate the
  complete song, and use functions to reduce the duplication.
*didn't end up working*


--------------------------------------------------------------------------------
# 19.3 Functions are for humans and computers

function names should be verbs, arguments should be nouns
use common prefixes if functions are related
use comments to explain "why", try to write code so that "what/how" are clear
break code up with dashes, like I've been doing!

## 19.3.1 Exercises

1. Read the source code for each of the following three functions, puzzle out 
  what they do, and then brainstorm better names.
```{r}
f1 <- function(string, prefix) {
  substr(string, 1, nchar(prefix)) == prefix
}

f1(c("hey", "hi", "what", "is", "up", "Holden"), "h")
# tests to see if a string has a given prefix
# rename has_prefix()

f2 <- function(x) {
  if (length(x) <= 1) return(NULL)
  x[-length(x)]
}

f2(1:6)
# drops last value of vector
# rename drop_last_value

f3 <- function(x, y) {
  rep(y, length.out = length(x))
}

f3(1:5, 1:3)
# repeats y so that is matches length of x
# rename repeat_to_match
```

2. Take a function that you’ve written recently and spend 5 minutes 
  brainstorming a better name for it and its arguments.

3. Compare and contrast rnorm() and MASS::mvrnorm(). How could you make them 
  more consistent?
*mvrnorm() should follow rnorm() naming conventions b/c it is more used*

4. Make a case for why norm_r(), norm_d() etc would be better than rnorm(), 
  dnorm(). Make a case for the opposite.
*similar functions that are closely related should have same prefix*


## 19.4 Conditional execution


