---
title: "Ch19_lab"
author: "Holden Jones"
date: '2022-12-08'
output: html_document
---

# Functions

going to move a bit quicker through this


--------------------------------------------------------------------------------
# 19.1 Introduction

3 big advantages over copy and paste:
- give a function a name that makes code easier to understand
- as requirements change, only need to update code in one place
- eliminate chance of making incidental mistakes 

## 19.1.1 Prerequisites

base R so no need


--------------------------------------------------------------------------------
# 19.2 When should you write a function?

if copy and pasting a bunch of code more than twice, write a function!

ex. this code randomizes a tibble using same parameters each time
```{r}
df <- tibble::tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)

df$a <- (df$a - min(df$a, na.rm = TRUE)) / 
  (max(df$a, na.rm = TRUE) - min(df$a, na.rm = TRUE))
df$b <- (df$b - min(df$b, na.rm = TRUE)) / 
  (max(df$b, na.rm = TRUE) - min(df$b, na.rm = TRUE)) # og was mistake here!
df$c <- (df$c - min(df$c, na.rm = TRUE)) / 
  (max(df$c, na.rm = TRUE) - min(df$c, na.rm = TRUE))
df$d <- (df$d - min(df$d, na.rm = TRUE)) / 
  (max(df$d, na.rm = TRUE) - min(df$d, na.rm = TRUE))
```

how many inputs does this code have?
- just one for each line

to make inputs more clear, good idea to rewrite code using temp vars
```{r}
x <- df$a
(x - min(x, na.rm = TRUE)) / (max(x, na.rm = TRUE) - min(x, na.rm = TRUE))

# there's duplication here, compute range of data once:
rng <- range(x, na.rm = TRUE)
(x - rng[1]) / (rng[2] - rng[1])
```

after siplifying code and checking that it still works, can turn into function
```{r}
rescale01 <- function(x) {
  rng <- range(x, na.rm = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
}
rescale01(c(0, 5, 10))
```

3 steps to creating new function:
- pick a name for the function, rescale01 b/c rescales b/ween 0 and 1
- list inputs or arguments to function inside function(), just one here, but
  with more, call looks like function(x,y,z)
- place code that have developed in body of function, a {} that immediately
  follows function(...)

easier to start with working code and turn it into a function, harder to create
  function first and then try to make it work
  
good to check function with a few different inputs
```{r}
rescale01(c(-10, 0, 10))
rescale01(c(1, 2, 3, NA, 5))
```

eventually will want to do formal, automated tests, called unit tests
but outside scope of this book

can simplify og code now that we have a function:
```{r}
df$a <- rescale01(df$a)
df$b <- rescale01(df$b)
df$c <- rescale01(df$c)
df$d <- rescale01(df$d)
```

this code easier to understand, but still quite a bit of duplication
will learn to improve this in iteration later

ex.if some var include infinite values, rescale01() will fail
- but can easily and efficiently correct it as a function
```{r}
x <- c(1:10, Inf)

rescale01 <- function(x) {
  rng <- range(x, na.rm = TRUE, finite = TRUE)
  y <- (x - rng[1]) / (rng[2] - rng[1])
  y[y == -Inf] <- 0
  y[y == Inf] <- 1
  y
}

(rescale01(x))
```

do not repeat yourself! DRY! principle: 
less repition = fewer needs to update = fewer bugs and errors

## 19.2.1 Exercises
*all from answers*

1. Why is TRUE not a parameter to rescale01()? What would happen if x 
  contained a single missing value, and na.rm was FALSE?
*b/c even if na.rm is FALSE, will return non-missing value*

2. In the second variant of rescale01(), infinite values are left unchanged.
  Rewrite rescale01() so that -Inf is mapped to 0, and Inf is mapped to 1.
```{r}
rescale01 <- function(x) {
  rng <- range(x, na.rm = TRUE, finite = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
}
```

3. Practice turning the following code snippets into functions. Think about 
  what each function does. What would you call it? How many arguments does it
  need? Can you rewrite it to be more expressive or less duplicative?
```{r}
x <- c(1:10)

# takes proportion of nas in a vector
mean(is.na(x))

# call prop_na:
prop_na <- function(x) {
  mean(is.na(x))
}
prop_na(x)

# look at prop of overall sum for each value in vector
x / sum(x, na.rm = TRUE)

prop_sum <- function(x) {
  x / sum(x, na.rm = TRUE)
}
prop_sum(x)

# calculates sd / mean aka coefficient of variation for a vector
sd(x, na.rm = TRUE) / mean(x, na.rm = TRUE)

coeff_var <- function(x, na.rm = TRUE) {
  sd(x, na.rm = na.rm) / mean(x, na.rm = na.rm)
}
coeff_var(x)
```

4. Write your own functions to compute the variance and skewness of a numeric
  vector.
*from answers*
```{r}
# variance
variance <- function(x, na.rm = TRUE) {
  n <- length(x)
  m <- mean(x, na.rm = TRUE)
  sq_err <- (x - m)^2
  sum(sq_err) / (n - 1)
}

variance(x)
var(x) # check

# skewness
skewness <- function(x, na.rm = FALSE) {
  n <- length(x)
  m <- mean(x, na.rm = na.rm)
  v <- var(x, na.rm = na.rm)
  (sum((x - m) ^ 3) / (n - 2)) / v ^ (3 / 2)
}

skewness(x)
```

5. Write both_na(), a function that takes two vectors of the same length and
  returns the number of positions that have an NA in both vectors.
```{r}
both_na <- function(x, y) {
  sum(is.na(x) & is.na(y))
}

both_na(
  c(NA, NA, 1, 2),
  c(NA, 1, NA, 2)
)

both_na(
  c(NA, NA, 1, 2, NA, NA, 1),
  c(NA, 1, NA, 2, NA, NA, 1)
)
```

6. What do the following functions do? Why are they useful even though they are
  so short?
*always good to make sure directory and file access is in order!*
```{r}
is_directory <- function(x) file.info(x)$isdir
is_readable <- function(x) file.access(x, 4) == 0
```

7. Read the complete lyrics to “Little Bunny Foo Foo”. There’s a lot of
  duplication in this song. Extend the initial piping example to recreate the
  complete song, and use functions to reduce the duplication.
*didn't end up working*


--------------------------------------------------------------------------------
# 19.3 Functions are for humans and computers

function names should be verbs, arguments should be nouns
use common prefixes if functions are related
use comments to explain "why", try to write code so that "what/how" are clear
break code up with dashes, like I've been doing!

## 19.3.1 Exercises

1. Read the source code for each of the following three functions, puzzle out 
  what they do, and then brainstorm better names.
```{r}
f1 <- function(string, prefix) {
  substr(string, 1, nchar(prefix)) == prefix
}

f1(c("hey", "hi", "what", "is", "up", "Holden"), "h")
# tests to see if a string has a given prefix
# rename has_prefix()

f2 <- function(x) {
  if (length(x) <= 1) return(NULL)
  x[-length(x)]
}

f2(1:6)
# drops last value of vector
# rename drop_last_value

f3 <- function(x, y) {
  rep(y, length.out = length(x))
}

f3(1:5, 1:3)
# repeats y so that is matches length of x
# rename repeat_to_match
```

2. Take a function that you’ve written recently and spend 5 minutes 
  brainstorming a better name for it and its arguments.

3. Compare and contrast rnorm() and MASS::mvrnorm(). How could you make them 
  more consistent?
*mvrnorm() should follow rnorm() naming conventions b/c it is more used*

4. Make a case for why norm_r(), norm_d() etc would be better than rnorm(), 
  dnorm(). Make a case for the opposite.
*similar functions that are closely related should have same prefix*


# 19.4 Conditional execution

an if statement allows you to conditionally execute code
if else
takes advantage of standard return rule: a function returns last value computed

## 19.4.1 Conditions

condition must evaluate to either true or false
use || (or) and && (and) to combine multiple logical expressions

## 19.4.2 Multiple conditions

can chain multiple if statements together
but if super long, should rewrite
can use switch() or cut() to help with this

## 19.4.3 Code style

Both if and function should (almost) always be followed by squiggly brackets 
({}), and the contents should be indented by two spaces. This makes it easier 
to see the hierarchy in your code by skimming the left-hand margin.

An opening curly brace should never go on its own line and should always be 
followed by a new line. A closing curly brace should always go on its own line,
unless it’s followed by else. Always indent the code inside curly braces.
```{r}
# Good
if (y < 0 && debug) {
  message("Y is negative")
}

if (y == 0) {
  log(x)
} else {
  y ^ x
}
```

okay to drop curly brackets if statement can fit on one line, otherwise keep

## 19.4.4 Exercises

1. What’s the difference between if and ifelse()? Carefully read the help and
  construct three examples that illustrate the key differences.
*if tests single condition, ifelse tests each element*

2. Write a greeting function that says “good morning”, “good afternoon”, or 
  “good evening”, depending on the time of day. (Hint: use a time argument that
  defaults to lubridate::now(). That will make it easier to test your function.)
*from answers, modified based on my thoughts*
```{r}
greet <- function(time = lubridate::now()) {
  hr <- lubridate::hour(time)
  if (hr < 5) {
    print("good night")
  } else if(hr < 12) {
    print("good morning")
  } else if (hr < 17) {
    print("good afternoon")
  } else if(hr < 20) {
    print("good evening")
  } else {
    print("good night")
  }
}

greet()
greet(lubridate::ymd_h("2017-01-08:20")) # this is confusing! hour is :20
greet(lubridate::ymd_h("2017-01-08:02"))
greet(lubridate::ymd_h("2017-01-08:05"))
```

3. Implement a fizzbuzz function. It takes a single number as input. If the 
  number is divisible by three, it returns “fizz”. If it’s divisible by five it
  returns “buzz”. If it’s divisible by three and five, it returns “fizzbuzz”.
  Otherwise, it returns the number. Make sure you first write working code 
  before you create the function.
```{r}
fizzbuzz <- function(x) {
     if (!(x %% 3) && !(x %% 5)) {
        print("fizzbuzz")
   } else if (!(x %% 3)) {
        print("fizz")
   } else if (!(x %% 5)) {
        print("buzz")
   } else {
        print(x)
   }  
}

fizzbuzz(15)
fizzbuzz(3)
fizzbuzz(5)
fizzbuzz(358)
fizzbuzz(1934)
fizzbuzz(2100)
```

4. How could you use cut() to simplify this set of nested if-else statements?
```{r}
if (temp <= 0) {
  "freezing"
} else if (temp <= 10) {
  "cold"
} else if (temp <= 20) {
  "cool"
} else if (temp <= 30) {
  "warm"
} else {
  "hot"
}

# from answers:
temp <- seq(-10, 50, by = 5)
cut(temp, c(-Inf, 0, 10, 20, 30, Inf),
  right = FALSE,
  labels = c("freezing", "cold", "cool", "warm", "hot")
)
```

How would you change the call to cut() if I’d used < instead of <=? What is 
  the other chief advantage of cut() for this problem? (Hint: what happens if 
  you have many values in temp?)
*cut works on vectors, unlike if which only works on one value*
  
5. What happens if you use switch() with numeric values?
*need to make sure they're integer values, b/c it does not round*
```{r}
switch(1, "apple", "banana", "cantaloupe")
switch(2, "apple", "banana", "cantaloupe")
switch(1.2, "apple", "banana", "cantaloupe")
switch(2.8, "apple", "banana", "cantaloupe")
```

6. What does this switch() call do? What happens if x is “e”?
*SKIP*
```{r}
switch(x, 
  a = ,
  b = "ab",
  c = ,
  d = "cd"
)

# don't understand, skip
```


# 19.5 Function arguments

arguments in two broad sets:
- supplies data to compute on
- control details of computation
data arguments first, controls at end which should have default values
```{r}
# Compute confidence interval around mean using normal approximation
mean_ci <- function(x, conf = 0.95) {
  se <- sd(x) / sqrt(length(x))
  alpha <- 1 - conf
  mean(x) + se * qnorm(c(alpha / 2, 1 - alpha / 2))
}

x <- runif(100)
mean_ci(x) # uses default unless otherwise specified
mean_ci(x, conf = 0.99)
```

default value should be the most common use, except when due to safety
ex. - na.rm = FALSE is default, b/c dangerous to default ignore missing values

when calling function, typically omit names of data arguments, b/c so common
if overriding default value of detail argument, then need to use full name

use whitespace, makes it easier to skim the function for important components

## 19.5.1 Choosing names

generally best to use longer, more descriptive names, but following are common:
- x, y, z: vectors.
- w: a vector of weights.
- df: a data frame.
- i, j: numeric indices (typically rows and columns).
- n: length, or number of rows.
- p: number of columns.
otherwise match existing argument names ex. na.rm to remove missing values

## 19.5.2 Checking values

eventually when write enough functions, will forget how each one works
becomes easy to call function with invalid inputs == VERY BAD!!
*to avoid this problem, make constraints explicit*
```{r}
# without constraints:
wt_mean <- function(x, w) {
  sum(x * w) / sum(w)
}

# what if x and w are not same length?
wt_mean(1:6, 1:3)
# OH NO! It still works and doesn't give us an error

# using constraints:
wt_mean <- function(x, w) {
  if (length(x) != length(w)) {
    stop("`x` and `w` must be the same length", call. = FALSE)
  }
  sum(w * x) / sum(w)
}

wt_mean(1:6, 1:3)
```

but be sure not to take this too far! have to balance time vs reward
instead, use stopifnot(), checks that each argument is TRUE, generic error if not
```{r}
wt_mean <- function(x, w, na.rm = FALSE) {
  stopifnot(is.logical(na.rm), length(na.rm) == 1)
  stopifnot(length(x) == length(w)) # assert what should be true here
  
  if (na.rm) {
    miss <- is.na(x) | is.na(w)
    x <- x[!miss]
    w <- w[!miss]
  }
  sum(w * x) / sum(w)
}

wt_mean(1:6, 6:1, na.rm = "foo")
```

## 19.5.3 Dot-dot-dot(...)

... is special argument that captures any number of arguments that aren't 
  otherwise matched
  
useful b/c can send those ... to another function, helpful if wrapping functions
```{r}
commas <- function(...) stringr::str_c(..., collapse = ", ")
commas(letters[1:10])

rule <- function(..., pad = "-") {
  title <- paste0(...)
  width <- getOption("width") - nchar(title) - 5
  cat(title, " ", stringr::str_dup(pad, width), "\n", sep = "")
}
rule("Important output")
```

basically can punt any arguments on to the next function (in above ex. str_c)
but comes at price, an misspelled arguments will not raise an error
so, easy for typos to go unnoticed

## 19.5.4 Lazy evaluation

arguments aren't computed until they're needed. so if never used, never called

## 19.5.5 Exercises

1. What does commas(letters, collapse = "-") do? Why?
*seems like its belnding arguments from 2 different functions that are separate*
*would need to add a separate collapse argument to commas to fix error*

2. It’d be nice if you could supply multiple characters to the pad argument, 
  e.g. rule("Title", pad = "-+"). Why doesn’t this currently work? How could 
  you fix it?
*can supply multiple chars to pad but output will be twice as long*
*can fix using str_trunc() to truncate string, str_length() to find chars in pad*
```{r}
# this uses dplyr
library(tidyverse)

rule <- function(..., pad = "-") {
  title <- paste0(...)
  width <- getOption("width") - nchar(title) - 5 # strange way to get width no?
  padding <- str_dup(
    pad,
    ceiling(width / str_length(title))
  ) %>%
    str_trunc(width)
  cat(title, " ", padding, "\n", sep = "")
}

# figured out width from above?: current R script width (80 for me) / nchar - 5
# then the longest it can be is the width / the length of the title?
rule("Important output")
rule("hi")
rule("Valuable output", pad = "-+")
rule("Vital output", pad = "-+-")
```

3. What does the trim argument to mean() do? When might you use it?
*cuts off observations from being included, could use w/ huge data, or if errors*

4. The default value for the method argument to cor() is c("pearson", "kendall",
  "spearman"). What does that mean? What value is used by default?
*which correlation coefficient to be computed, pearson is default*


## 19.6 Return values

Things to think about when considering what your function should return:
- Does returning early make your function easier to read?
- Can you make your function pipeable?

# 19.6.1 Explicit return statements

can choose to return early by using return()
tends to make really long code easier to understand, b/c don't need as much
context to understand

## 19.6.2 Writing pipeable functions

if you want to write pipeable functions, think about return value
knowing return value's object type means pipeline will "just work"

2 basic types of pipeable functions:
- transformations
- side-effects


# 19.7 Environment

Last component of a function is its environment

not really that important for now
- lexical scoping, R will lok in environment to find names of things in functions
for example can override '+", devious! but also super powerful and advanced



