---
title: "Ch21_lab"
author: "Holden Jones"
date: '2023-01-17'
output: html_document
---

# Iteration


--------------------------------------------------------------------------------
# 21.1 Introduction

Super important to reduce code duplication. 3 main benefits:
- Easier to see intent of code, eyes drawn to differences
- Easier to respond to changes in requirements, changes in one place only
- Likely to have fewer bugs b/c each line of code used in more places

Functions are one tool for reducing duplication, iteration is another one!
- here will learn about imperative programming and functional programming

imperative - for and while loops, good places to start but verbose
functional - more complicated, but once mastered can solve many iteration probs

## 21.1.1 Prerequisites

will use base R and purr
```{r}
library(tidyverse)
```


--------------------------------------------------------------------------------
# 21.2 For loops

simple tibble
```{r}
df <- tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)
```

want to compute median of each column, use for loop instead of copy and paste
```{r}
output <- vector("double", ncol(df))  # 1. output

for (i in seq_along(df)) {            # 2. sequence
  output[[i]] <- median(df[[i]])      # 3. body
}

output
```

every for loop has three components:
1. output 
  - before start loop must allocate sufficient space for output
  - super important for efficiency
2. sequence
  - determines what to loop over: each run of for loop will assign i to diff
    value from a sequence. Think of i as a pronoun, like "it"
  - seq_along() is safer version of familiar 1:length(), but does right thing
    for a zero-length vector
3. body
  - this is the code that does the work
  - run repeatedly, each time with different value for i
  
## 21.2.1 Exercises

1. Write for loops to:

- Compute the mean of every column in mtcars.
```{r}
output <- vector("double", ncol(mtcars))

for (i in seq_along(mtcars)) {
  output[[i]] <- mean(mtcars[[i]])
}

output
```

- Determine the type of each column in nycflights13::flights.
```{r}
library(nycflights13)

# still getting some errors here:
ouput <- vector("list", ncol(flights)) # used list not char

for (i in seq_along(flights)) {
  output[[i]] <- str(flights[[i]])
}

output
```
- Compute the number of unique values in each column of iris.
```{r}
# this is mine, gets right answer but also printing leftovers from flights
ouput <- vector("double", ncol(iris))
for (i in seq_along(iris)) {
  output[[i]] <- n_distinct(iris[[i]])
}
output

# answer sheets, prints right amount - why different?
iris_uniq <- vector("double", ncol(iris))
names(iris_uniq) <- names(iris)
for (i in names(iris)) {
  iris_uniq[i] <- n_distinct(iris[[i]])
}
iris_uniq
```

- Generate 10 random normals from distributions with means of -10, 0, 10, and 100.
```{r}
n <- 10
mu <- c(-10, 0, 10, 100)

output <- vector("list", length(mu))  # 1. output

for (i in seq_along(output)) {            # 2. sequence
  output[[i]] <- rnorm(n, mu[i])      # 3. body
}

output
```

Think about the output, sequence, and body before you start writing the loop.

2. Eliminate the for loop in each of the following examples by taking advantage 
  of an existing function that works with vectors:
```{r}
out <- ""
for (x in letters) {
  out <- stringr::str_c(out, x)
}
out

# use str_c
str_c(letters, collapse = "")

x <- sample(100)
sdev <- 0
for (i in seq_along(x)) {
  sdev <- sdev + (x[i] - mean(x)) ^ 2
}
sdev <- sqrt(sdev / (length(x) - 1))
sdev

# use sd
sd(x)

# this is calculating a cumulative sum
x <- runif(100)
out <- vector("numeric", length(x))
out[1] <- x[1]
for (i in 2:length(x)) {
  out[i] <- out[i - 1] + x[i]
}
out

# use cumsum
all.equal(cumsum(x), out)
```

3. Combine your function writing and for loop skills:
- Write a for loop that prints() the lyrics to the children’s song “Alice the
  camel”.
- Convert the nursery rhyme “ten in the bed” to a function. Generalise it to any
  number of people in any sleeping structure.
- Convert the song “99 bottles of beer on the wall” to a function. Generalise to
  any number of any vessel containing any liquid on any surface.
```{r}
# camels
humps <- c("five", "four", "three", "two", "one", "no")

for (i in humps) {
  cat(str_c("Alice the camel has ", rep(i, 3), " humps.",
    collapse = "\n"
  ), "\n")
  if (i == "no") {
    cat("Now Alice is a horse.\n")
  } else {
    cat("So go, Alice, go.\n")
  }
  cat("\n")
}

# sleeping
structure <- "beefmarphshlarpdong"
numbers <- c("ten", "nine", "eight", "seven", "six", "five", "four", "three", 
             "two", "one", "no")

for (i in numbers) {
  cat(str_c("There were ", i, " in the ", print(structure)))
  cat("\nand the little one said\n")
  if (i == "one") {
    cat("I'm lonely...")
  } else {
    cat("Roll over, roll over\n")
    cat("So they all rolled over and one fell out.\n")
  }
  cat("\n")
}

# wall drinking - from answers
bottles <- function(n) {
  if (n > 1) {
    str_c(n, " bottles")
  } else if (n == 1) {
    "1 bottle"
  } else {
    "no more bottles"
  }
}

beer_bottles <- function(total_bottles) {
  # print each lyric
  for (current_bottles in seq(total_bottles, 0)) {
    # first line
    cat(str_to_sentence(str_c(bottles(current_bottles), " of beer on the wall, ",
                              bottles(current_bottles), " of beer.\n")))   
    # second line
    if (current_bottles > 0) {
      cat(str_c(
        "Take one down and pass it around, ", bottles(current_bottles - 1),
        " of beer on the wall.\n"
      ))          
    } else {
      cat(str_c("Go to the store and buy some more, ", bottles(total_bottles), 
                " of beer on the wall.\n"))                }
    cat("\n")
  }
}

beer_bottles(99)
```

4. It’s common to see for loops that don’t preallocate the output and instead
  increase the length of a vector at each step:
```{r}
output <- vector("integer", 0)
for (i in seq_along(x)) {
  output <- c(output, lengths(x[[i]]))
}
output
```

How does this affect performance? Design and execute an experiment.
*Allocation is way faster!*
```{r}
# from answers
library('microbenchmark')

# w/out preallocation
add_to_vector <- function(n) {
  output <- vector("integer", 0)
  for (i in seq_len(n)) {
    output <- c(output, i)
  }
  output
}

# w/ preallocation
add_to_vector_2 <- function(n) {
  output <- vector("integer", n)
  for (i in seq_len(n)) {
    output[[i]] <- i
  }
  output
}

timings <- microbenchmark(add_to_vector(10000), add_to_vector_2(10000), times = 10)
timings
```


--------------------------------------------------------------------------------
# 21.3 For loop variations

4 variations on basic theme of for loop
- modify existing object, instead of create new object
- looping over names or values, instead of indices
- handling outputs of unknown length
- handling sequences of unknown length

## 21.3.1 Modifying an existing object

think of rescaling columns example from functions:
```{r}
df <- tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)

rescale01 <- function(x) {
  rng <- range(x, na.rm = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
}

df$a <- rescale01(df$a)
df$b <- rescale01(df$b)
df$c <- rescale01(df$c)
df$d <- rescale01(df$d)
```

think about the three components:
- output - same as then input!
- sequence - iterate over each column with seq_along(df)
- body - apply rescale01()
```{r}
for (i in seq_along(df)) {
  df[[i]] <- rescale01(df[[i]])
}
```

Hadley recommends using [[]] for all for loops, makes clear working with single
  element
  
## 21.3.2 Looping patterns

3 basic ways to loop over a vector, for (i in seq_along(xs)) is most general
- loop over elements: for (x in xs). most useful if only care about side-effects
  like plotting or saving a file
- loop over names: for (nm in names(xs)). gives you name, which can use to
  access value with x[[nm]]. useful if want to use name in plot title or file.

## 21.3.3 Unknown output length

sometimes don't know how long output will be
ex. simulate random vectors of random lengths
```{r}
means <- c(0, 1, 2)

output <- double()
for (i in seq_along(means)) {
  n <- sample(100, 1)
  output <- c(output, rnorm(n, means[[i]]))
}
str(output)
```

growing vector might seem good but will get quadratic behavior = SUPER SLOW

better solution is to save results in list, then combine into single vector
```{r}
out <- vector("list", length(means))

for(i in seq_along(means)) {
  n <- sample(100, 1)
  out[[i]] <- rnorm(n, means[[i]])
}

str(out)

str(unlist(out))
```

here used unlist(), but stricter would be to use purr::flatten_dbl()

## 21.3.4 Unkown sequence length

Sometimes don't know how long input sequence should run for. Common w/ sims
ex. want to loop until get three heads in a row
  - can't do it with for loop, need to use while loop
  - while loop is simpler than for loop - only 2 components: condition and body
  
while (condition) {
  # body
}

while loops also more general than for loops
- can rewrite any for loop as a while loop
- but can't rewrite every whil loop as a for loop

use while loop to see how many tries to flip 3 heads in a row
```{r}
flip <- function() sample(c("T", "H"), 1)

flips <- 0
nheads <- 0

while (nheads < 3) { # only while you have less than 3 H
  if (flip() == "H") { # if you get a head
    nheads <- nheads + 1 # add to the nheads counter
  } else {
    nheads <- 0 # if it's not a head, counter resets to 0
  }
  flips <- flips + 1 # flips counter grows by one each time
}

flips
```

## 21.3.5 Exercises

1. Imagine you have a directory full of CSV files that you want to read in. You 
  have their paths in a vector, 
  files <- dir("data/", pattern = "\\.csv$", full.names = TRUE), 
  and now want to read each one with read_csv(). Write the for loop that will 
  load them into a single data frame.
```{r}
output <- vector("list", length(files))

for (i in seq_along(files)) {
  output[[files]] <- read_csv(files[[i]])
}
```

2. What happens if you use for (nm in names(x)) and x has no names? What if only
  some of the elements are named? What if the names are not unique?
*guessing that error if no names, also if some are not named, repeat if overlap*
*pretty close - 0 length loop if no names, uses 1st name if overlap*

3. Write a function that prints the mean of each numeric column in a data frame, 
  along with its name. For example, show_mean(iris) would print:
*from answers*
```{r}
show_mean <- function(df, digits = 2) {
  # Get max length of all variable names in the dataset
  maxstr <- max(str_length(names(df)))
  for (nm in names(df)) {
    if (is.numeric(df[[nm]])) {
      cat(
        str_c(str_pad(str_c(nm, ":"), maxstr + 1L, side = "right"),
          format(mean(df[[nm]]), digits = digits, nsmall = digits),
          sep = " "
        ),
        "\n"
      )
    }
  }
}

show_mean(iris)
```

4. What does this code do? How does it work?
*mutates display and am columns of mtcars*
```{r}
trans <- list( 
  disp = function(x) x * 0.0163871,
  am = function(x) {
    factor(x, labels = c("auto", "manual"))
  }
)
for (var in names(trans)) {
  mtcars[[var]] <- trans[[var]](mtcars[[var]])
}
```


--------------------------------------------------------------------------------
# 21.4 For loops vs. functionals

for loops not as important in R than other languages b/c R is a functional
  programming language - means can wrap for loops in a function, and call the
  function instead of using for loop directly
  
ex. simple data frame
```{r}
df <- tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)
```

compute mean of every column, can do with a for loop
```{r}
output <- vector("double", length(df))

for (i in seq_along(df)) {
  output[[i]] <- mean(df[[i]])
}

output
```

also can make it into a function b/c doing it frequently
```{r}
col_mean <- function(df) {
  output <- vector("double", length(df))
  for (i in seq_along(df)) {
    output[i] <- mean(df[[i]])
  }
  output
}
```

but now what if we want to get median, sd? don't copy and paste! generalise...

here's an example - want to make changes in code super obvious
```{r}
# instead of:
f1 <- function(x) abs(x - mean(x)) ^ 1
f2 <- function(x) abs(x - mean(x)) ^ 2
f3 <- function(x) abs(x - mean(x)) ^ 3

# extract into single argument, remove lots of duplication:
f <- function(x, i) abs(x - mean(x)) ^ i
```

so let's do that with our col_mean(), col_median(), col_sd() code:
```{r}
col_summary <- function(df, fun) { # make the function an argument
  out <- vector("double", length(df))
  for (i in seq_along(df)) {
    out[i] <- fun(df[[i]]) # perform the function
  }
  out
}

# this is cool! can get lots more flexibility / use out of this code now
col_summary(df, median)

col_summary(df, mean)

col_summary(df, max)
```

*idea of passing a function to another function is extremely powerful idea!!!!!*

rest of chapter will learn about purr package - functions that eliminate need
  for many common for loops
- the apply family in base R solve similar problem, but purr is more consistent

goal of using purr functions instead of for loops - break common list manipulation
  into independent pieces
- how can solve problem for single element of list? once solved, purr generalises
  solution for every element in list
- if solving complex problem, can break it down into bite-sized pieces
  with purr get lots of small pieces that can compose together with pipe
  
## 21.4.1 Exercises

1. Read the documentation for apply(). In the 2d case, what two for loops does 
  it generalise?
*replaces looping over the columns and rows in a df*

2. Adapt col_summary() so that it only applies to numeric columns You might want 
  to start with an is_numeric() function that returns a logical vector that has 
  a TRUE corresponding to each numeric column.
```{r}
# my attempt - IT WORKS!
col_summary <- function(df, fun) { 
  df <- select_if(df, is.numeric)
  out <- vector("double", length(df))
  for (i in seq_along(df)) {
    out[i] <- fun(df[[i]])
  }
  out
}

col_summary(mpg, mean)
```


--------------------------------------------------------------------------------
# 21.5 The map functions

The pattern of looping over a vector, doing something, and saving results is
  so common purr has family of functions that do it for you
- map() makes a list
- map_lgl() makes logical vector
- map_int() makes an integer vector
- map_dbl() makes a double vector
- map_chr() makes character vector

each function takes vector, applies function, returns new vector of same length
  that is type determined by the suffix attached to the map function
  
mastering these makes iteration problems way faster

*most important thing is to solve problem you're working on!!!!!*
- not write the most concise and elegant code - although that's good to strive for

map makes code easier to write and read

use these functions to perform same computations as last for loop
```{r}
map_dbl(df, mean)

map_dbl(df, median)

map_dbl(df, sd)
```

compared to for loop, the focus is on operation being performed, not bookkeeping
- even more so if use the pipe!

differences between map_*() compared to col_summary()
- all purrr functions implemented in C - faster but less readability
- second arg, .f can be formula, character vector, or integer vector
- map_*() uses ... to pass along additional arguments to .f each time called

## 21.5.1 Shortcuts

can use shortcuts with .f to save a little typing

imagine want to fit linear model to each group in dataset
this ex. splits mtcars into 3 pieces, for each value of cylinder, fit same model
  for each
```{r}
models <- mtcars %>% 
  split(.$cyl) %>% # split by value of cyl
  map(function(df) lm(mpg ~ wt, data = df)) # run linear model for each!
```

*split seems like a great function to know about!*

or what about looking at many models - extract summary statistic like R2
first run summary() then extract component called r.squared
```{r}
models %>% 
  map(summary) %>% 
  map_dbl(~.$r.squared)

# can also use string to extract named components - super common!
models %>% 
  map(summary) %>% 
  map_dbl("r.squared")
```

## 21.5.2 Base R

if familiar with apply functions in base R, might have noticied similarities
  with purr
- lappply() basically identical to map(), except for shortcuts
- base sapply() is a wrapper around lapply() automatically simpliefies output
- vapply() safe alternative to sapply() b/c supply additional argument that 
  defines type
  
but here focuses on purrr functions b/c more consistent names and args
  
## 21.5.3 Exercises

1. Write code that uses one of the map functions to:
- Compute the mean of every column in mtcars.
```{r}
map_dbl(mtcars, mean)
```

- Determine the type of each column in nycflights13::flights.
```{r}
map_chr(flights, typeof)
```

- Compute the number of unique values in each column of iris.
```{r}
map_dbl(iris, n_distinct)
```

- Generate 10 random normals from distributions with means of -10, 0, 10, and 100.
```{r}
mu <- c(-10, 0, 10, 100)

map(mu, ~rnorm(10, mean = mu)) # not sure why we need to ~ here, or only use map
```

2. How can you create a single vector that for each column in a data frame 
  indicates whether or not it’s a factor?
```{r}
map_lgl(diamonds, is.factor)
```

3. What happens when you use the map functions on vectors that aren’t lists? 
  What does map(1:5, runif) do? Why?
*map can take vectors but output will be a list*
*result of this code is length five list with numeric vector sizes 1 through 5*
```{r}
map(1:5, runif)
```

4. What does map(-2:2, rnorm, n = 5) do? Why? 
  What does map_dbl(-2:2, rnorm, n = 5) do? Why?
```{r}
# seems to keep n for rnorm constant
map(-2:2, rnorm, n = 5)

# error b/c not returning numeric vector of length 1 as required by map_dbl()
map_dbl(-2:2, rnorm, n = 5)
```

5. Rewrite map(x, function(df) lm(mpg ~ wt, data = df)) to eliminate the 
  anonymous function.
*doesn't run, use other code instead*
```{r}
x <- split(mtcars, mtcars$cyl)
map(x, function(df) lm(mpg ~ wt, data = df))

# can eliminate use of anonymosu function using ~
map(x, ~ lm(mpg ~ wt, data = .))
```


--------------------------------------------------------------------------------
# 21.6 Dealing with failure

if using map functions to repeat many operations, likely one will fail
when this happens, will get an error message with no output

how to ensure that one bad apple doesn't ruin the whole bunch?

use new function: safely()
- it is an adverb, takes a function (verb) and returns a modified version

modified function will never throw an error. instead returns list w/ 2 elements
- result is OG result. if there was an error this will be NULL
- error is an error object. if operation was successful, this will be NULL

w/ simple example: log()
```{r}
safe_log <- safely(log)

# have value in result, NULL in error
str(safe_log(10))

# have NULL in result, value in error
str(safe_log("a"))
```

safely() is designed to work with map
```{r}
x <- list(1, 10, "a")
y <- x %>% map(safely(log))

# a list for as long as the object being passed through is
str(y)
```

would be easier to work with if had two lists: all errors and all output
- use purr::transpose() to get it
```{r}
y <- y %>% transpose()

# so a max list of 2 here
str(y)
```

depends how to deal with errors. typically will either:
- look at values of x where y is an error
- or work with values of y that are ok
```{r}
is_ok <- y$error %>% map_lgl(is_null)
x[!is_ok]

y$result[is_ok] %>% flatten_dbl()
```

possibly() and quietly() function similarly to safely()


--------------------------------------------------------------------------------
# 21.7 Mapping over multiple arguments

so far have only mapped along a single input
- if mapping along related inputs in parallel, use map2() and pmap() functions

ex. simulating random normals with different means
- can do this with map()
```{r}
mu <- list(5, 10, -3)

mu %>% 
  map(rnorm, n = 5) %>% 
  str()
```

how about varying the standard deviation?
- one way would be to iterate over indices and index into vectors of means and sds
```{r}
sigma <- list(1, 5, 10)

seq_along(mu) %>% 
  map(~rnorm(5, mu[[.]], sigma[[.]])) %>% 
  str()
```

but better to use map2() which is clearer, iterates over 2 vectors in parallel
```{r}
map2(mu, sigma, rnorm, n = 5) %>% str()
```

*Note that the arguments that vary for each call come before the function;* 
  *arguments that are the same for every call come after.*

map3,4,5 etc would get tedious quickly! use pmap, takes list of arguments
```{r}
n <- list(1, 3, 5)

args1 <- list(n, mu, sigma)

#notic that output varies based on n
args1 %>%
  pmap(rnorm) %>% 
  str()
```
better to not depend on positional matching, so name the arguments!
```{r}
args2 <- list(mean = mu, sd = sigma, n = n)

args2 %>% 
  pmap(rnorm) %>% 
  str()
```

args are all same length, so makes sense to store them in data frame
```{r}
params <- tribble(
  ~mean, ~sd, ~n,
    5,     1,  1,
   10,     5,  3,
   -3,    10,  5
)

params %>% 
  pmap(rnorm)
```

once code gets complicated, a data frame is a good approach
- ensures each column has a name
- and each column is same length as all the other columns

## 21.7.1 Invoking different functions

could also vary the function itself, not just the arguments to the function
```{r}
f <- c("runif", "rnorm", "rpois")

param <- list(
  list(min = -1, max = 1), 
  list(sd = 5), 
  list(lambda = 10)
)

# can use invoke map to handle this case
invoke_map(f, param, n = 5) %>% str()

# can really scale up complexity this way!
```

use tribble to make creating these matching pairs a bit easier
```{r}
sim <- tribble(
  ~f,      ~params,
  "runif", list(min = -1, max = 1),
  "rnorm", list(sd = 5),
  "rpois", list(lambda = 10)
)

sim %>% 
  mutate(sim = invoke_map(f, params, n = 10))
```


--------------------------------------------------------------------------------
# 21.8 Walk

use walk when want to call function for side effects, not return value
- typically when want to render output to screen or save files to disk
- important thing is action, not return value

```{r}
x <- list(1, "a", 3)

x %>% 
  walk(print)
```

walk() not useful compared to walk2() or pwalk()
- ex. list of plots and vector of file names, pwalk() to save each file
```{r}
library(ggplot2)

plots <- mtcars %>% 
  split(.$cyl) %>% 
  map(~ggplot(., aes(mpg, wt)) + geom_point())

paths <- stringr::str_c(names(plots), ".pdf")

pwalk(list(paths, plots), ggsave, path = tempdir())
```


--------------------------------------------------------------------------------
# 21.9 Other patterns of for loops

number of other functions abstract over other types of for loops
goal here is to previously introduce them so if run into similar prob in future

## 21.9.1 Predicate functions

predicate functions return either a single TRUE or FALSE
- keep() and discard() keep elements of input where predicate is TRUE or FALSE

```{r}
iris %>% 
  keep(is.factor) %>% 
  str()

iris %>% 
  discard(is.factor) %>% 
  str()
```

some() and every() determine if predicate is true for any or for all of elements
```{r}
x <- list(1:5, letters, list(10))

x %>% 
  some(is_character)

x %>% 
  every(is_vector)
```

detect() finds first element where predicate is true, detect_index() returns pos
```{r}
(x <- sample(10))

x %>% 
  detect(~ . > 5)

x %>% 
  detect_index(~ . > 5)

# head_while() and tail_while() take elements from start or end of vector
# while predicate is true
x %>% 
  head_while(~ . > 5)

x %>% 
  tail_while(~ . > 5)
```

## 21.9.2 Reduce and accumulate

sometimes have complex list that want to reduce to simple list
- by repeatedly applying a function that reduces pair to singleton

ex list of data frames, want to reduce to a single data frame joining together
*this seems useful!*
```{r}
dfs <- list(
  age = tibble(name = "John", age = 30),
  sex = tibble(name = c("John", "Mary"), sex = c("M", "F")),
  trt = tibble(name = "Mary", treatment = "A")
)

dfs %>% reduce(full_join)
```

or maybe have list of vectors, want to find intersection
```{r}
vs <- list(
  c(1, 3, 5, 6, 10),
  c(1, 2, 3, 7, 8, 10),
  c(1, 2, 3, 4, 8, 9, 10)
)

vs %>% reduce(intersect)
```

reduce() takes binary function, applies repeatedly to list until only single
  element left
  
accumulate() similar but keeps all interim results. could use it to implement a
  cumulative sum
```{r}
x <- sample(10)
x

x %>% accumulate(`+`)
```

## 21.9.3 Exercises

1. Implement your own version of every() using a for loop. Compare it with
  purrr::every(). What does purrr’s version do that your version doesn’t?
*purr's version does fancy stuff w/ predicate*
```{r}
# from answers

# Use ... to pass arguments to the function
every2 <- function(.x, .p, ...) {
  for (i in .x) {
    if (!.p(i, ...)) {
      # If any is FALSE we know not all of then were TRUE
      return(FALSE)
    }
  }
  # if nothing was FALSE, then it is TRUE
  TRUE
}

every2(1:3, function(x) {
  x > 1
})

every2(1:3, function(x) {
  x > 0
})
```

2. Create an enhanced col_summary() that applies a summary function to every
  numeric column in a data frame.
```{r}
# the col_summary function from the chapter
col_summary <- function(df, fun) {
  out <- vector("double", length(df))
  for (i in seq_along(df)) {
    out[i] <- fun(df[[i]])
  }
  out
}

# from answers, use map and keep
col_sum2 <- function(df, f, ...) {
  map(keep(df, is.numeric), f, ...)
}

col_sum2(iris, mean)
```


3. A possible base R equivalent of col_summary() is:
```{r}
col_sum3 <- function(df, f) {
  is_num <- sapply(df, is.numeric)
  df_num <- df[, is_num]

  sapply(df_num, f)
}
```

But has a number of bugs as illustrated with the following inputs:
```{r}
df <- tibble(
  x = 1:3, 
  y = 3:1,
  z = c("a", "b", "c")
)
# OK
col_sum3(df, mean)
# Has problems: don't always return numeric vector
col_sum3(df[1:2], mean)
col_sum3(df[1], mean)
col_sum3(df[0], mean)
```

What causes the bugs?
*sapply causes the bugs - answers*