---
title: "Ch21_lab"
author: "Holden Jones"
date: '2023-01-17'
output: html_document
---

# Iteration


--------------------------------------------------------------------------------
# 21.1 Introduction

Super important to reduce code duplication. 3 main benefits:
- Easier to see intent of code, eyes drawn to differences
- Easier to respond to changes in requirements, changes in one place only
- Likely to have fewer bugs b/c each line of code used in more places

Functions are one tool for reducing duplication, iteration is another one!
- here will learn about imperative programming and functional programming

imperative - for and while loops, good places to start but verbose
functional - more complicated, but once mastered can solve many iteration probs

## 21.1.1 Prerequisites

will use base R and purr
```{r}
library(tidyverse)
```


--------------------------------------------------------------------------------
# 21.2 For loops

simple tibble
```{r}
df <- tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)
```

want to compute median of each column, use for loop instead of copy and paste
```{r}
output <- vector("double", ncol(df))  # 1. output

for (i in seq_along(df)) {            # 2. sequence
  output[[i]] <- median(df[[i]])      # 3. body
}

output
```

every for loop has three components:
1. output 
  - before start loop must allocate sufficient space for output
  - super important for efficiency
2. sequence
  - determines what to loop over: each run of for loop will assign i to diff
    value from a sequence. Think of i as a pronoun, like "it"
  - seq_along() is safer version of familiar 1:length(), but does right thing
    for a zero-length vector
3. body
  - this is the code that does the work
  - run repeatedly, each time with different value for i
  
## 21.2.1 Exercises

1. Write for loops to:

- Compute the mean of every column in mtcars.
```{r}
output <- vector("double", ncol(mtcars))

for (i in seq_along(mtcars)) {
  output[[i]] <- mean(mtcars[[i]])
}

output
```

- Determine the type of each column in nycflights13::flights.
```{r}
library(nycflights13)

# still getting some errors here:
ouput <- vector("list", ncol(flights)) # used list not char

for (i in seq_along(flights)) {
  output[[i]] <- str(flights[[i]])
}

output
```
- Compute the number of unique values in each column of iris.
```{r}
# this is mine, gets right answer but also printing leftovers from flights
ouput <- vector("double", ncol(iris))
for (i in seq_along(iris)) {
  output[[i]] <- n_distinct(iris[[i]])
}
output

# answer sheets, prints right amount - why different?
iris_uniq <- vector("double", ncol(iris))
names(iris_uniq) <- names(iris)
for (i in names(iris)) {
  iris_uniq[i] <- n_distinct(iris[[i]])
}
iris_uniq
```

## RESUME HERE

- Generate 10 random normals from distributions with means of -10, 0, 10, and 100.
Think about the output, sequence, and body before you start writing the loop.

2. Eliminate the for loop in each of the following examples by taking advantage 
  of an existing function that works with vectors:
```{r}
out <- ""
for (x in letters) {
  out <- stringr::str_c(out, x)
}

x <- sample(100)
sd <- 0
for (i in seq_along(x)) {
  sd <- sd + (x[i] - mean(x)) ^ 2
}
sd <- sqrt(sd / (length(x) - 1))

x <- runif(100)
out <- vector("numeric", length(x))
out[1] <- x[1]
for (i in 2:length(x)) {
  out[i] <- out[i - 1] + x[i]
}
```

3. Combine your function writing and for loop skills:
- Write a for loop that prints() the lyrics to the children’s song “Alice the
  camel”.
- Convert the nursery rhyme “ten in the bed” to a function. Generalise it to any
  number of people in any sleeping structure.
- Convert the song “99 bottles of beer on the wall” to a function. Generalise to
  any number of any vessel containing any liquid on any surface.

4. It’s common to see for loops that don’t preallocate the output and instead
  increase the length of a vector at each step:
```{r}
output <- vector("integer", 0)
for (i in seq_along(x)) {
  output <- c(output, lengths(x[[i]]))
}
output
```

How does this affect performance? Design and execute an experiment.


--------------------------------------------------------------------------------
# 21.3 For loop variations

