---
title: "Ch24_lab"
author: "Holden Jones"
date: '2023-01-30'
output: html_document
---

# Model building


--------------------------------------------------------------------------------
# 24.1 Introduction

this chapter shows how to progressively build up model to aid understanding of 
  data
  
a challenge to know when to stop building models

## 24.1.1 Prerequisites

```{r}
library(tidyverse)
library(modelr)
options(na.action = na.warn)

library(nycflights13)
library(lubridate)
```


--------------------------------------------------------------------------------
# 24.2 Why are low quality diamonds more expensive?

in previous chapters have seen weird relationship b/ween quality and price
- ie poor cuts, bad colours, inferior clarity, have higher prices
```{r}
ggplot(diamonds, aes(cut, price)) + geom_boxplot()
ggplot(diamonds, aes(color, price)) + geom_boxplot()
ggplot(diamonds, aes(clarity, price)) + geom_boxplot()
```

## 24.2.1 Price and carat

important confounding variable - weight (carat) of diamond
- weight is single most important factor for determining price
- lower quality diamonds tend to be larger
```{r}
ggplot(diamonds, aes(carat, price)) + 
  geom_hex(bins = 50)
```

fit a model to separate out effect of carat, but first tweak dataset to make
  easier to work with
- focus on diamonds smaller than 2.5 carats (99.7% of data)
- log-transform carat and price variables
```{r}
diamonds2 <- diamonds %>% 
  filter(carat <= 2.5) %>% 
  mutate(lprice = log2(price), lcarat = log2(carat))

# makes it easier to see relationship b/ween carat and price
ggplot(diamonds2, aes(lcarat, lprice)) + 
  geom_hex(bins = 50)
```

log-transformation is particularly useful b/c makes pattern linear
*linear patterns are easiest to work with!*

let's take next step and remove strong linear pattern
```{r}
mod_diamond <- lm(lprice ~ lcarat, data = diamonds2)

# look at what model teslls us about data
# note that back transform predictions, undoing log transform
# so can overlay predictions on raw data
grid <- diamonds2 %>% 
  data_grid(carat = seq_range(carat, 20)) %>% 
  mutate(lcarat = log2(carat)) %>% 
  add_predictions(mod_diamond, "lprice") %>% 
  mutate(price = 2 ^ lprice)

ggplot(diamonds2, aes(carat, price)) + 
  geom_hex(bins = 50) + 
  geom_line(data = grid, colour = "red", size = 1)
```

now can lok at residuals, to verify that have successfully removed strong
  linear pattern
```{r}
diamonds2 <- diamonds2 %>% 
  add_residuals(mod_diamond, "lresid")

ggplot(diamonds2, aes(lcarat, lresid)) + 
  geom_hex(bins = 50)
```

IMPORTANT - now can re-do motivating plots using these residuals instead of price
- I'm not sure why we do this
- So that linear effect of carat has been taken out of it?
```{r}
ggplot(diamonds2, aes(cut, lresid)) + geom_boxplot()
ggplot(diamonds2, aes(color, lresid)) + geom_boxplot()
ggplot(diamonds2, aes(clarity, lresid)) + geom_boxplot()
```

Yes, no we see what we'd expect - as quality of diamond increases, so does price
- y axis need to think about what residuals are telling us, and what scale on
- residual of -1 indicates lprice was 1 unit lower than prediction based solely
  on weight
- residuals with value 1 are twice the predicted price

## 24.2.2 A more complicated model

If wanted to, could continue to build up model, moving effects observed into
  model to make them explicit
- ex. could include color, cut, clarity so also make explicit effect of these
  3 categorical variables
```{r}
mod_diamond2 <- lm(lprice ~ lcarat + color + cut + clarity, data = diamonds2)
```

but this model now has 4 predictors! and is harder to visualize
- use model argument to data_grid
```{r}
grid <- diamonds2 %>% 
  data_grid(cut, .model = mod_diamond2) %>% 
  add_predictions(mod_diamond2)
grid

ggplot(grid, aes(cut, pred)) + 
  geom_point()
```

if model needs variables that haven't explicitly supplied, data_grid() will
  automatically fill them in with "typical" value
- for continuous uses median
- for categorical uses most common value
```{r}
diamonds2 <- diamonds2 %>% 
  add_residuals(mod_diamond2, "lresid2")

ggplot(diamonds2, aes(lcarat, lresid2)) + 
  geom_hex(bins = 50)
```

this plot indicates some diamonds have quite large residuals!
```{r}
diamonds2 %>%
  filter(abs(lresid2) > 1) %>%
  add_predictions(mod_diamond2) %>%
  mutate(pred = round(2 ^ pred)) %>%
  select(price, pred, carat:table, x:z) %>%
  arrange(price)
```

notice that some of these are way higher, some way lower than predictions
- opportunity to review these 'outliers' against the model predictions

## 24.2.3 Exercises

1. In the plot of lcarat vs. lprice, there are some bright vertical strips. 
  What do they represent?
*spots where there are lots of diamonds - round numbers!*
```{r}
# this is the plot
ggplot(diamonds2, aes(lcarat, lprice)) + 
  geom_hex(bins = 50)
```

2. If log(price) = a_0 + a_1 * log(carat), what does that say about the 
  relationship between price and carat?
*there is a nonlinear relationship b/ween price and carat*

3. Extract the diamonds that have very high and very low residuals. Is there 
  anything unusual about these diamonds? Are they particularly bad or good, or 
  do you think these are pricing errors?

4. Does the final model, mod_diamond2, do a good job of predicting diamond 
  prices? Would you trust it to tell you how much to spend if you were buying 
  a diamond?





