---
title: "Ch24_lab"
author: "Holden Jones"
date: '2023-01-30'
output: html_document
---

# Model building


--------------------------------------------------------------------------------
# 24.1 Introduction

this chapter shows how to progressively build up model to aid understanding of 
  data
  
a challenge to know when to stop building models

## 24.1.1 Prerequisites

```{r}
library(tidyverse)
library(modelr)
options(na.action = na.warn)

library(nycflights13)
library(lubridate)
```


--------------------------------------------------------------------------------
# 24.2 Why are low quality diamonds more expensive?

in previous chapters have seen weird relationship b/ween quality and price
- ie poor cuts, bad colours, inferior clarity, have higher prices
```{r}
ggplot(diamonds, aes(cut, price)) + geom_boxplot()
ggplot(diamonds, aes(color, price)) + geom_boxplot()
ggplot(diamonds, aes(clarity, price)) + geom_boxplot()
```

## 24.2.1 Price and carat

important confounding variable - weight (carat) of diamond
- weight is single most important factor for determining price
- lower quality diamonds tend to be larger
```{r}
ggplot(diamonds, aes(carat, price)) + 
  geom_hex(bins = 50)
```

fit a model to separate out effect of carat, but first tweak dataset to make
  easier to work with
- focus on diamonds smaller than 2.5 carats (99.7% of data)
- log-transform carat and price variables
```{r}
diamonds2 <- diamonds %>% 
  filter(carat <= 2.5) %>% 
  mutate(lprice = log2(price), lcarat = log2(carat))

# makes it easier to see relationship b/ween carat and price
ggplot(diamonds2, aes(lcarat, lprice)) + 
  geom_hex(bins = 50)
```

log-transformation is particularly useful b/c makes pattern linear
*linear patterns are easiest to work with!*

let's take next step and remove strong linear pattern
```{r}
mod_diamond <- lm(lprice ~ lcarat, data = diamonds2)

# look at what model teslls us about data
# note that back transform predictions, undoing log transform
# so can overlay predictions on raw data
grid <- diamonds2 %>% 
  data_grid(carat = seq_range(carat, 20)) %>% 
  mutate(lcarat = log2(carat)) %>% 
  add_predictions(mod_diamond, "lprice") %>% 
  mutate(price = 2 ^ lprice)

ggplot(diamonds2, aes(carat, price)) + 
  geom_hex(bins = 50) + 
  geom_line(data = grid, colour = "red", size = 1)
```

now can lok at residuals, to verify that have successfully removed strong
  linear pattern
```{r}
diamonds2 <- diamonds2 %>% 
  add_residuals(mod_diamond, "lresid")

ggplot(diamonds2, aes(lcarat, lresid)) + 
  geom_hex(bins = 50)
```

IMPORTANT - now can re-do motivating plots using these residuals instead of price
- I'm not sure why we do this
- So that linear effect of carat has been taken out of it?
```{r}
ggplot(diamonds2, aes(cut, lresid)) + geom_boxplot()
ggplot(diamonds2, aes(color, lresid)) + geom_boxplot()
ggplot(diamonds2, aes(clarity, lresid)) + geom_boxplot()
```

Yes, no we see what we'd expect - as quality of diamond increases, so does price
- y axis need to think about what residuals are telling us, and what scale on
- residual of -1 indicates lprice was 1 unit lower than prediction based solely
  on weight
- residuals with value 1 are twice the predicted price

## 24.2.2 A more complicated model

If wanted to, could continue to build up model, moving effects observed into
  model to make them explicit
- ex. could include color, cut, clarity so also make explicit effect of these
  3 categorical variables
```{r}
mod_diamond2 <- lm(lprice ~ lcarat + color + cut + clarity, data = diamonds2)
```

but this model now has 4 predictors! and is harder to visualize
- use model argument to data_grid
```{r}
grid <- diamonds2 %>% 
  data_grid(cut, .model = mod_diamond2) %>% 
  add_predictions(mod_diamond2)
grid

ggplot(grid, aes(cut, pred)) + 
  geom_point()
```

if model needs variables that haven't explicitly supplied, data_grid() will
  automatically fill them in with "typical" value
- for continuous uses median
- for categorical uses most common value
```{r}
diamonds2 <- diamonds2 %>% 
  add_residuals(mod_diamond2, "lresid2")

ggplot(diamonds2, aes(lcarat, lresid2)) + 
  geom_hex(bins = 50)
```

this plot indicates some diamonds have quite large residuals!
```{r}
diamonds2 %>%
  filter(abs(lresid2) > 1) %>%
  add_predictions(mod_diamond2) %>%
  mutate(pred = round(2 ^ pred)) %>%
  select(price, pred, carat:table, x:z) %>%
  arrange(price)
```

notice that some of these are way higher, some way lower than predictions
- opportunity to review these 'outliers' against the model predictions

## 24.2.3 Exercises

1. In the plot of lcarat vs. lprice, there are some bright vertical strips. 
  What do they represent?
*spots where there are lots of diamonds - round numbers!*
```{r}
# this is the plot
ggplot(diamonds2, aes(lcarat, lprice)) + 
  geom_hex(bins = 50)
```

2. If log(price) = a_0 + a_1 * log(carat), what does that say about the 
  relationship between price and carat?
*there is a nonlinear relationship b/ween price and carat*

3. Extract the diamonds that have very high and very low residuals. Is there 
  anything unusual about these diamonds? Are they particularly bad or good, or 
  do you think these are pricing errors?
*don't really notice anything strange here*
```{r}
# code is already above
diamonds2 %>%
  filter(abs(lresid2) > 1) %>%
  add_predictions(mod_diamond2) %>%
  mutate(pred = round(2^pred)) %>%
  select(price, pred, carat:table, x:z) %>%
  arrange(price)
```

4. Does the final model, mod_diamond2, do a good job of predicting diamond 
  prices? Would you trust it to tell you how much to spend if you were buying 
  a diamond?
*No I don't have a ton of confidence with this model*
```{r}
# lets plot residuals for this final model
ggplot(diamonds2, aes(lcarat, lresid2)) +
  geom_hex(bins = 50)

# quite a few residuals
```


--------------------------------------------------------------------------------
# 24.3 What affects the number of daily flights?

using really simple dataset - number of flights that leave NYC per day
```{r}
daily <- flights %>% 
  mutate(date = make_date(year, month, day)) %>% 
  group_by(date) %>% 
  summarise(n = n())

daily

ggplot(daily, aes(date, n)) + 
  geom_line()
```

## 24.3.1 Day of week

strong day of week effect that dominates subtler patterns
```{r}
daily <- daily %>%
  mutate(wday = wday(date, label = TRUE))

ggplot(daily, aes(wday, n)) +
  geom_boxplot()
# clear strong effect of weekdays having higher flights
```

one way to remove this strong pattern is to use a model
- 1st, fit model and display its predictions overlaid on original data
```{r}
mod <- lm(n ~ wday, data = daily)

grid <- daily %>% 
  data_grid(wday) %>% 
  add_predictions(mod, "n")

ggplot(daily, aes(wday, n)) + 
  geom_boxplot() +
  geom_point(data = grid, colour = "red", size = 4)
```

next, compute and visualise the residuals
```{r}
daily <- daily %>% 
  add_residuals(mod)

daily %>% 
  ggplot(aes(date, resid)) + 
  geom_ref_line(h = 0) + 
  geom_line()
```

plot above is now seeing deviation from expected number of flights, given day of
  week
- can now see some of subtler patterns that remain
  - says that model seems to fail starting in June - see strong regular pattern
    that model hasn't captured
  - looks to me that the model fails from the beginning! look at residuals up
    until June!
but I digress, let's plot day of the week over residuals
```{r}
ggplot(daily, aes(date, resid, colour = wday)) + 
  geom_ref_line(h = 0) + 
  geom_line()
```

our model fails to accurately predict number of flights on Saturday
- during summer more flights than expected
- during fall fewer

there are some days with far fewer flights than expected
```{r}
daily %>%
  filter(abs(resid) > 100)
# interesting, only one day with waaaay more flights than expected
# the largest differential residuals are negative - holidays!
```

smoother long term trend over course of year, highlight w/ geom_smooth()
```{r}
daily %>% 
  ggplot(aes(date, resid)) + 
  geom_ref_line(h = 0) + 
  geom_line(colour = "grey50") + 
  geom_smooth(se = FALSE, span = 0.20)
```

## 24.3.2 Seasonal Saturday effect

focus on Saturdays - we failed to accurately predict number of flights on Saturday
```{r}
daily %>% 
  filter(wday == "Sat") %>% 
  ggplot(aes(date, n)) + 
    geom_point() + 
    geom_line() +
    scale_x_date(NULL, date_breaks = "1 month", date_labels = "%b")
```

seems like summer breaks - nobody has problem traveling on Sat for vacation
- why more Saturday flights in Spring than Fall? more family vacays?
```{r}
# create term variable to capture three school terms
term <- function(date) {
  cut(date, 
    breaks = ymd(20130101, 20130605, 20130825, 20140101),
    labels = c("spring", "summer", "fall") 
  )
}

daily <- daily %>% 
  mutate(term = term(date)) 

daily %>% 
  filter(wday == "Sat") %>% 
  ggplot(aes(date, n, colour = term)) +
  geom_point(alpha = 1/3) + 
  geom_line() +
  scale_x_date(NULL, date_breaks = "1 month", date_labels = "%b")
```

and how does term variable affect other days of the week?
```{r}
daily %>% 
  ggplot(aes(wday, n, colour = term)) +
    geom_boxplot()
# idk, seeing a lot of overlap across terms for each day of the week
```

incorporate term into model, how does this improve fit?
```{r}
# og model
mod1 <- lm(n ~ wday, data = daily)
# new model with term interaction
mod2 <- lm(n ~ wday * term, data = daily)

# plot both residuals on same graph
daily %>% 
  gather_residuals(without_term = mod1, with_term = mod2) %>% 
  ggplot(aes(date, resid, colour = model)) +
    geom_line(alpha = 0.75)
```

overlay predictions from model onto raw data:
```{r}
grid <- daily %>% 
  data_grid(wday, term) %>% 
  add_predictions(mod2, "n")

ggplot(daily, aes(wday, n)) +
  geom_boxplot() + 
  geom_point(data = grid, colour = "red") + 
  facet_wrap(~ term)
```

we still have a lot of big outliers here! mean tends to be far away from typical
  value

so lets use a model robust to effect of outliers: MASS::rlm()
```{r}
mod3 <- MASS::rlm(n ~ wday * term, data = daily)

daily %>% 
  add_residuals(mod3, "resid") %>% 
  ggplot(aes(date, resid)) + 
  geom_hline(yintercept = 0, size = 2, colour = "white") + 
  geom_line()
```

this model does much better! now a lot easier to see long-term trend, and big
  outliers both positive and negative
  
## 24.3.3 Computed variables

good idea to bundle creation of variables into function so no change of 
  accidentally applying different transformation in different places
```{r}
compute_vars <- function(data) {
  data %>% 
    mutate(
      term = term(date), 
      wday = wday(date, label = TRUE)
    )
}

# another option to put transformations directly into model formula
wday2 <- function(x) wday(x, label = TRUE)
mod3 <- lm(n ~ wday2(date) * term(date), data = daily)
```

## 24.3.4 Time of year: an alternative approach

previously used domain knowledge to improve model
- alternative is to "give data more room to speak"
- could use more flexible model and allow that to capture pattern interested in
- simple linear trend isn't adequate, so could try to use a natural spline to fit
  a smooth curve across the year
```{r}
library(splines)
mod <- MASS::rlm(n ~ wday * ns(date, 5), data = daily)

daily %>% 
  data_grid(wday, date = seq_range(date, n = 13)) %>% 
  add_predictions(mod) %>% 
  ggplot(aes(date, pred, colour = wday)) + 
    geom_line() +
    geom_point()
```

## 24.3.5 Exercises

1. Use your Google sleuthing skills to brainstorm why there were fewer than 
  expected flights on Jan 20, May 26, and Sep 1. (Hint: they all have the same 
  explanation.) How would these days generalise to another year?
*Sundays before Monday holidays*

2. What do the three days with high positive residuals represent? How would 
  these days generalise to another year?
*weekends right after Thanksgiving, and Christmas*
```{r}
daily %>% 
  slice_max(n = 3, resid)
```

3. Create a new variable that splits the wday variable into terms, but only for 
  Saturdays, i.e. it should have Thurs, Fri, but Sat-summer, Sat-spring, 
  Sat-fall. How does this model compare with the model with every combination 
  of wday and term?
*from answers*
*honestly super small difference! also visually, not noticeable difference*
```{r}
# add new var, wday2
daily <- daily %>%
  mutate(
    wday2 =
      case_when(
        wday == "Sat" & term == "summer" ~ "Sat-summer",
        wday == "Sat" & term == "fall" ~ "Sat-fall",
        wday == "Sat" & term == "spring" ~ "Sat-spring",
        TRUE ~ as.character(wday)
      )
  )

# make new model
mod4 <- MASS::rlm(n ~ wday2, data = daily)

# comparison model
mod3 <- MASS::rlm(n ~ wday * term, data = daily)

# adds residuals of both to daily
daily %>%
  gather_residuals(sat_term = mod4, all_interact = mod3) %>%
  ggplot(aes(date, resid, colour = model)) +
  geom_line(alpha = 0.75)

# use glance to compare model fits
library(broom)

glance(mod4)

glance(mod3)

# AICs are almost identical!
```

4. Create a new wday variable that combines the day of week, term 
  (for Saturdays), and public holidays. What do the residuals of that model 
  look like?
```{r}
holidays_2013 <-
  tribble(
    ~holiday, ~date,
    "New Year's Day", 20130101,
    "Martin Luther King Jr. Day", 20130121,
    "Washington's Birthday", 20130218,
    "Memorial Day", 20130527,
    "Independence Day", 20130704,
    "Labor Day", 20130902,
    "Columbus Day", 20131028,
    "Veteran's Day", 20131111,
    "Thanksgiving", 20131128,
    "Christmas", 20131225
  ) %>%
  mutate(date = lubridate::ymd(date))

daily <- daily %>%
  mutate(
    wday3 =
      case_when(
        date %in% (holidays_2013$date - 1L) ~ "day before holiday",
        date %in% (holidays_2013$date + 1L) ~ "day after holiday",
        date %in% holidays_2013$date ~ "holiday",
        .$wday == "Sat" & .$term == "summer" ~ "Sat-summer",
        .$wday == "Sat" & .$term == "fall" ~ "Sat-fall",
        .$wday == "Sat" & .$term == "spring" ~ "Sat-spring",
        TRUE ~ as.character(.$wday)
      )
  )

mod4 <- lm(n ~ wday3, data = daily)

daily %>%
  spread_residuals(resid_sat_terms = mod3, resid_holidays = mod4) %>%
  mutate(resid_diff = resid_holidays - resid_sat_terms) %>%
  ggplot(aes(date, resid_diff)) +
  geom_line(alpha = 0.75)
```

5. What happens if you fit a day of week effect that varies by month 
  (i.e. n ~ wday * month)? Why is this not very helpful?
*from answers, running out of steam in this book!*
*likely not a great model - b/c way too many combinations*
```{r}
daily <- mutate(daily, month = factor(lubridate::month(date)))
mod6 <- lm(n ~ wday * month, data = daily)
print(summary(mod6))
```

6. What would you expect the model n ~ wday + ns(date, 5) to look like? Knowing 
  what you know about the data, why would you expect it to be not particularly 
  effective?
*probably not going to do very well b/c doesn't interact with time of year*

7. We hypothesised that people leaving on Sundays are more likely to be business 
  travellers who need to be somewhere on Monday. Explore that hypothesis by 
  seeing how it breaks down based on distance and time: if it’s true, you’d 
  expect to see more Sunday evening flights to places that are far away.
*from answers*
```{r}
flights %>%
  mutate(
    date = make_date(year, month, day),
    wday = wday(date, label = TRUE)
  ) %>%
  ggplot(aes(y = distance, x = wday)) +
  geom_boxplot() +
  labs(x = "Day of Week", y = "Average Distance")

# hide outliers
flights %>%
  mutate(
    date = make_date(year, month, day),
    wday = wday(date, label = TRUE)
  ) %>%
  ggplot(aes(y = distance, x = wday)) +
  geom_boxplot(outlier.shape = NA) +
  labs(x = "Day of Week", y = "Average Distance")

# pointrange with mean and standard error of mean
flights %>%
  mutate(
    date = make_date(year, month, day),
    wday = wday(date, label = TRUE)
  ) %>%
  ggplot(aes(y = distance, x = wday)) +
  stat_summary() +
  labs(x = "Day of Week", y = "Average Distance")

flights %>%
  mutate(
    date = make_date(year, month, day),
    wday = wday(date, label = TRUE)
  ) %>%
  filter(
    distance < 3000,
    hour >= 5, hour <= 21
  ) %>%
  ggplot(aes(x = hour, color = wday, y = ..density..)) +
  geom_freqpoly(binwidth = 1)

flights %>%
  mutate(
    date = make_date(year, month, day),
    wday = wday(date, label = TRUE)
  ) %>%
  filter(
    distance < 3000,
    hour >= 5, hour <= 21
  ) %>%
  group_by(wday, hour) %>%
  summarise(distance = mean(distance)) %>%
  ggplot(aes(x = hour, color = wday, y = distance)) +
  geom_line()

flights %>%
  mutate(
    date = make_date(year, month, day),
    wday = wday(date, label = TRUE)
  ) %>%
  filter(
    distance < 3000,
    hour >= 5, hour <= 21
  ) %>%
  group_by(wday, hour) %>%
  summarise(distance = sum(distance)) %>%
  group_by(wday) %>%
  mutate(prop_distance = distance / sum(distance)) %>%
  ungroup() %>%
  ggplot(aes(x = hour, color = wday, y = prop_distance)) +
  geom_line()
```

8. It’s a little frustrating that Sunday and Saturday are on separate ends of 
  the plot. Write a small function to set the levels of the factor so that the 
  week starts on Monday.
```{r}
monday_first <- function(x) {
  fct_relevel(x, levels(x)[-1])
}

daily <- daily %>%
  mutate(wday = wday(date, label = TRUE))
ggplot(daily, aes(monday_first(wday), n)) +
  geom_boxplot() +
  labs(x = "Day of Week", y = "Number of flights")
```

