---
title: "Ch25_lab"
author: "Holden Jones"
date: '2023-01-31'
output: html_document
---

# Many models


--------------------------------------------------------------------------------
# 25.1 Introduction

learn 3 powerful ideas to help work with large numbers of models
- using many simple models to better understand complex datasets
- using list-columns to store arbitrary data structures in data frame
  - ex. have column containing linear models
- using broom package to turn models into tidy data
  - powerful technique b/c once have tidy data, can apply all techniques
    learned earlier in this book
    
using real life example about life expectancy

following sections dive into more detail about individual techniques
- list-columns - list-column data structure, why valid to put lists into 
  data frames
- creating list-columns - learn 3 main ways in which create list-columns
- simplifying list-columns - learn how to convert list-columns back to regular
  atomic vectors so can work with them more easily
- making tidy data with broom - learn full set of tools provided by broom, how
  to apply to other types of data structure
  
## 25.1.1 Prerequisites

```{r}
library(modelr)
library(tidyverse)
```


--------------------------------------------------------------------------------
# 25.2 gapminder

from Hans Rosling - Swedish doctor and statistician

```{r}
library(gapminder)
gapminder
```

focus on: how does lifeExp change over year for each country
```{r}
gapminder %>%
  ggplot(aes(year, lifeExp, group = country)) +
  geom_line(alpha = 1/3)
# see that generally life expectancy increases over time, but some exceptions!
```

think like from last chapter - strong signal of overall linear growth makes hard
  to see subtler trends
- model captures steady growth over time, and residuals will show what's left
```{r}
# know how to do this for one country
nz <- filter(gapminder, country == "New Zealand")
nz %>% 
  ggplot(aes(year, lifeExp)) + 
  geom_line() + 
  ggtitle("Full data = ")

nz_mod <- lm(lifeExp ~ year, data = nz)
nz %>% 
  add_predictions(nz_mod) %>%
  ggplot(aes(year, pred)) + 
  geom_line() + 
  ggtitle("Linear trend + ")

nz %>% 
  add_residuals(nz_mod) %>% 
  ggplot(aes(year, resid)) + 
  geom_hline(yintercept = 0, colour = "white", size = 3) + 
  geom_line() + 
  ggtitle("Remaining pattern")
```

how to easily fit that model to every country?

## 25.2.1 Nested data

extract out common code with a function and repeat using map function from purrr
- to repeat an action for each country (a subset of rows), need a nested data frame
```{r}
by_country <- gapminder %>% 
  group_by(country, continent) %>% 
  nest()

by_country
```

this creates data frame with one row per group (per country) with unusual column
  data! data is a list of data frames (tibbles to be precise)
```{r}
# pluck out single entry in data to look at it
by_country$data[[1]]
```

## 25.2.2 List-columns

now that have nested data frame, in good position to fit some models
```{r}
country_model <- function(df) {
  lm(lifeExp ~ year, data = df)
}

# and want to apply it to every data frame. apply country_model to each element
models <- map(by_country$data, country_model)
```

better to store list of models as column in by_country data frame
- storing related objects in columns key value of data frames
- instead of creating new object in global env, create new variable in by_country
  - a job for mutate()!
```{r}
by_country <- by_country %>% 
  mutate(model = map(data, country_model))
by_country
```

makes it easier to sort and filter
```{r}
by_country %>% 
  filter(continent == "Europe")
```

## 25.2.3 Unnesting

now have 142 dataframes and 142 models, to compute residuals call add_residuals()
  with each model-data pair
```{r}
by_country <- by_country %>% 
  mutate(
    resids = map2(data, model, add_residuals)
  )
by_country

# turn list of data frames back into regular data frame, use unnest()
resids <- unnest(by_country, resids)
resids
```

now have a regular data frame, and can plot residuals
```{r}
resids %>%
  ggplot(aes(year, resid)) +
  geom_line(aes(group = country)) +
  geom_smooth()

# can facet by continent
resids %>%
  ggplot(aes(year, resid)) +
  geom_line(aes(group = country)) +
  geom_smooth() +
  facet_wrap(~continent)

# lots of residuals in Africa, Asia, rest of world looks pretty good
```

## 25.2.4 Model quality

Instead of looking at residuals from model, could look at some general measures of
  model quality
```{r}
library(broom)

glance(nz_mod)

# use mutate and unnest to create data frame with row with glance for each country
by_country %>%
  mutate(glance = map(model, glance)) %>%
  unnest(glance)

# don't want all the list columns
glance <- by_country %>% 
  mutate(glance = map(model, broom::glance)) %>% 
  unnest(glance, .drop = TRUE)

glance

# look for models that don't fit well
glance %>% 
  arrange(r.squared)

# worst models appear to be in Africa, check with plot
glance %>% 
  ggplot(aes(continent, r.squared)) + 
    geom_jitter(width = 0.5)
```

pull out countries with particularly bad R2 and plot data
```{r}
bad_fit <- filter(glance, r.squared < 0.25)

gapminder %>% 
  semi_join(bad_fit, by = "country") %>% 
  ggplot(aes(year, lifeExp, colour = country)) +
    geom_line()
# see two main effects - Rwandan genocide, HIV/AIDS epidemic
```

## 25.2.5 Exercises

1. A linear trend seems to be slightly too simple for the overall trend. Can you 
  do better with a quadratic polynomial? How can you interpret the coefficients 
  of the quadratic? (Hint you might want to transform year so that it has mean 
  zero.)
*from answers*
```{r}
lifeExp ~ poly(year, 2)

country_model <- function(df) {
  lm(lifeExp ~ poly(year - median(year), 2), data = df)
}

by_country <- gapminder %>%
  group_by(country, continent) %>%
  nest()

by_country <- by_country %>%
  mutate(model = map(data, country_model))

by_country <- by_country %>%
  mutate(
    resids = map2(data, model, add_residuals)
  )
by_country

unnest(by_country, resids) %>%
  ggplot(aes(year, resid)) +
  geom_line(aes(group = country), alpha = 1 / 3) +
  geom_smooth(se = FALSE)

by_country %>%
  mutate(glance = map(model, broom::glance)) %>%
  unnest(glance, .drop = TRUE) %>%
  ggplot(aes(continent, r.squared)) +
  geom_jitter(width = 0.5)
```

2. Explore other methods for visualising the distribution of R2 per continent. 
  You might want to try the ggbeeswarm package, which provides similar methods 
  for avoiding overlaps as jitter, but uses deterministic methods.
*from answers*
```{r}
library("ggbeeswarm")

by_country %>%
  mutate(glance = map(model, broom::glance)) %>%
  unnest(glance, .drop = TRUE) %>%
  ggplot(aes(continent, r.squared)) +
  geom_beeswarm()
```

3. To create the last plot (showing the data for the countries with the worst 
  model fits), we needed two steps: we created a data frame with one row per 
  country and then semi-joined it to the original dataset. It’s possible to avoid 
  this join if we use unnest() instead of unnest(.drop = TRUE). How?
*from answers*
```{r}
gapminder %>%
  group_by(country, continent) %>%
  nest() %>%
  mutate(model = map(data, ~lm(lifeExp ~ year, .))) %>%
  mutate(glance = map(model, broom::glance)) %>%
  unnest(glance) %>%
  unnest(data) %>%
  filter(r.squared < 0.25) %>%
  ggplot(aes(year, lifeExp)) +
  geom_line(aes(color = country))
```


--------------------------------------------------------------------------------
# 25.3 List-columns

dive back into some of the details for managing many models, after seeing workflow

```{r}
# base R doesn't make it easy to create list-columns
data.frame(x = list(1:3, 3:5))

# can prevent data.frame from doing this with I(), but result doesn't print well
data.frame(
  x = I(list(1:3, 3:5)), 
  y = c("1, 2", "3, 4, 5")
)
```

tibble() is better
```{r}
tibble(
  x = list(1:3, 3:5), 
  y = c("1, 2", "3, 4, 5")
)

# also easier with tribble(), automatically knows that need a list:
tribble(
   ~x, ~y,
  1:3, "1, 2",
  3:5, "3, 4, 5"
)
```

effective list-column pipeline:
- create list-column using one of nest(), summarise() + list(), or mutate() + 
  a map function
- create other intermediate list-columns by transforming existing list columns
  with map(), map2(), or pmap()
- simplify list-column back down to data frame or atomic vector


--------------------------------------------------------------------------------
# 25.4 Creating list-columns

typically won't create list-columns with tibble(). Instead, create from regular
  columns, using one of three methods:
- with tidyr::best() - to convert grouped data frame into nested data frame where
  have list-column of data frames
- with mutate() - and vectorised functions that return a list
- with summarise() - and summary functions that return multiple results
- alternatively, could create them from a named list, using tibble::enframe()

generally, make sure list-columns are homogenous - each element should contain
  same type of thing

## 25.4.1 With nesting

nest() creates a nested data frame, which is data frame with a list-column of 
  data frames
- each row is a meta observation - other columns give variables that define
  observation
  
2 ways to use nest
- have seen how to use it with grouped data frame
  - nest keeps grouping columns as is, and bundles everything else into list-col
- can also use it on ungrouped data frame, specifying which columns want to nest
```{r}
# nest with a grouped data frame - keeps grouping columns, bundles all else
gapminder %>% 
  group_by(country, continent) %>% 
  nest()

# nest with ungrouped data frame - specify what to nest
gapminder %>% 
  nest(data = c(year:gdpPercap))
```

## 25.4.2 From vectorised functions

some useful functions take atomic vector and return a list
```{r}
df <- tribble(
  ~x1,
  "a,b,c", 
  "d,e,f,g"
) 

df %>% 
  mutate(x2 = stringr::str_split(x1, ","))
```

unnest knows how to handle these lists of vectors
```{r}
df %>% 
  mutate(x2 = stringr::str_split(x1, ",")) %>% 
  unnest(x2)
```

## 25.4.3 From multivalued summaries

one restriction of summarise() is that only works with summary functions that
  return single value
- means can't use it with functions like quantile() that return vector of arbitrary
  length
```{r}
mtcars %>% 
  group_by(cyl) %>% 
  summarise(q = quantile(mpg))

# can wrap result in list! obeys contract of summarise
mtcars %>% 
  group_by(cyl) %>% 
  summarise(q = list(quantile(mpg)))

# to make useful results with unnest, also need capture probabilities
probs <- c(0.01, 0.25, 0.5, 0.75, 0.99)
mtcars %>% 
  group_by(cyl) %>% 
  summarise(p = list(probs), q = list(quantile(mpg, probs))) %>% 
  unnest(c(p, q))
```

## 25.4.4 From a named list

data frames with list-columns provide solution to common problem
  - what to do if want to iterate over both contents of list and its elements?
```{r}
x <- list(
  a = 1:5,
  b = 3:4, 
  c = 5:6
) 

df <- enframe(x)
df

# if want to iterate over names and values in parallel, can use map2()
df %>% 
  mutate(
    smry = map2_chr(name, value, ~ stringr::str_c(.x, ": ", .y[1]))
  )
```

## 25.4.5 Exercises

1. List all the functions that you can think of that take a atomic vector and 
  return a list.
*many functions in stringr, map() function*

2. Brainstorm useful summary functions that, like quantile(), return multiple 
  values.
*range, fivenum, I think glance, no?*

3. What’s missing in the following data frame? How does quantile() return that 
  missing piece? Why isn’t that helpful here?
*the particular quantiles of values are missing (ie 0, 25, 50, 75, 100 %)
```{r}
mtcars %>% 
  group_by(cyl) %>% 
  summarise(q = list(quantile(mpg))) %>% 
  unnest(q)

quantile(mtcars$mpg)

# since unnest function drops names of vector, not useful here
```

4. What does this code do? Why might might it be useful?
*not sure!*
```{r}
mtcars %>% 
  group_by(cyl) %>% 
  summarise_all(list(list))
```


--------------------------------------------------------------------------------
#25.5 Simplifying list-columns

to apply techniques of data manipulation and visualization learned in this book,
need to simplify list-column back to regular column or set of columns
- technique use to collapse down to simpler structure depends on whether want
  single value per element, or multiple values
  - if want single value, use mutate(), with map_lgl(), map_int(), map_dbl() and
    map_chr() to create atomic vector
  - if want many values, use unnest() to convert list-columns back to regular
    columns, repeating rows as many times as necessary
    
## 25.5.1 List to vector

if can reduce list column to atomic vector then will be regular column
- ex can always summarise object with type and length, this code will work
  regardless of what sort of list-column you have
```{r}
df <- tribble(
  ~x,
  letters[1:5],
  1:3,
  runif(5)
)
  
df %>% mutate(
  type = map_chr(x, typeof),
  length = map_int(x, length)
)

# also don't forget about map_*() shortcuts
df <- tribble(
  ~x,
  list(a = 1, b = 2),
  list(a = 2, c = 4)
)

df %>% mutate(
  a = map_dbl(x, "a"),
  b = map_dbl(x, "b", .null = NA_real_)
)
```

## 25.5.2 Unnesting

unnest() works by repeating regular columns once for each element of list-column
```{r}
tibble(x = 1:2, y = list(1:4, 1)) %>% unnest(y)

# means cant simultaneously unnest two columns that contain different # of elements

# Ok, because y and z have the same number of elements in
# every row
df1 <- tribble(
  ~x, ~y,           ~z,
   1, c("a", "b"), 1:2,
   2, "c",           3
)
df1

df1 %>% unnest(c(y, z))

# Doesn't work because y and z have different number of elements
df2 <- tribble(
  ~x, ~y,           ~z,
   1, "a",         1:2,  
   2, c("b", "c"),   3
)
df2

df2 %>% unnest(c(y, z))
```

same principle applies when unnesting list-columns or data frames
- can unnest multiple list-cols as long as all data frames in each row
  have same number of rows
  
## 25.5.3 Exercises

1. Why might the lengths() function be useful for creating atomic vector columns 
  from list-columns?
*important to know if all elements have same length for list-columns!*

2. List the most common types of vector found in a data frame. What makes lists 
  different?
*logical, numeric, integer, character, factor*
*all of common types of vectors in data frames are atomic. lsits are not atomic*


--------------------------------------------------------------------------------
# 25.6 Making tidy data with broom

broom package provides three general tools for turning models into tidy data frames
- glance(model) - returns a row for each model. each column gives a model summary,
  either a measure of model quality, or complexity, or both
- tidy(model) - returns row for each coefficient in the model. each column gives
  information about estimate or variability
- augment(model, data) - returns a row for each row in data, adding extra values
  like residuals, influence statistics





