---
title: "Ch27_lab"
author: "Holden Jones"
date: '2023-01-31'
output: html_document
---

# R Markdown


--------------------------------------------------------------------------------
# 27.1 Introduction

R Markdown provides unified framework for data science
combines code, results, and prose commentary

designed to be used in 3 ways
- communicating to decision makers, focus on conclusions not code
- collaborating with other data scientists (including future Holden!) who
  are interested in both conclusions, and how reached them (code)
- as an environment in which to do data science, modern day lab notebook to 
  capture what did, and what thinking
  
## 27.1.1 Prerequisites

rmarkdown package is automatically installed with RStudio


--------------------------------------------------------------------------------
# 27.2 R Markdown basics

R Markdown files are plain text files that have extension .Rmd

3 important types of content
- an optional YAML header surrounded by ---s
- chunks of R code surrounded by '''
- text mixed with simple formatting like # heading and _italics_

to produce complete report containing all text, code and results, click "knit"
- can do this programmatically with rmarkdown::render("1-example.Rmd")
- displays report in viewer plane, creates self-contained HTML file

## 27.2.1 Exercises

1. Create a new notebook using File > New File > R Notebook. Read the instructions. 
  Practice running the chunks. Verify that you can modify the code, re-run it, 
  and see modified output.

2. Create a new R Markdown document with File > New File > R Markdown… Knit it 
  by clicking the appropriate button. Knit it by using the appropriate keyboard 
  short cut. Verify that you can modify the input and see the output update.

3. Compare and contrast the R notebook and R markdown files you created above. 
  How are the outputs similar? How are they different? How are the inputs 
  similar? How are they different? What happens if you copy the YAML header from 
  one to the other?
*R notebook files show output of code chunks inside editor, while hiding console*

4. Create one new R Markdown document for each of the three built-in formats: 
  HTML, PDF and Word. Knit each of the three documents. How does the output 
  differ? How does the input differ? (You may need to install LaTeX in order to 
  build the PDF output — RStudio will prompt you if this is necessary.)


--------------------------------------------------------------------------------
# 27.3 Text formatting with Markdown

prose in .Rmd files is written in Markdown, lightweight set of convetions for
  formatting plain text files
- designed to be easy to learn as well


## 27.3.1 Exercises

1. Practice what you’ve learned by creating a brief CV. The title should be your 
  name, and you should include headings for (at least) education or employment. 
  Each of the sections should include a bulleted list of jobs/degrees. Highlight 
  the year in bold.

2. Using the R Markdown quick reference, figure out how to:
- Add a footnote.
- Add a horizontal rule.
- Add a block quote.

3. Copy and paste the contents of diamond-sizes.Rmd from
  https://github.com/hadley/r4ds/tree/master/rmarkdown in to a local R markdown 
  document. Check that you can run it, then add text after the frequency polygon 
  that describes its most striking features.

*checked the answer key for all of these*


--------------------------------------------------------------------------------
# 27.4 Code chunks

use keyboard shortcut!

to insert code chunks and to run entire code chunks

chunks are like functions - should be relatively self-contained, focussed
  around a single task

## 27.4.1 Chunk name

can name chunks!
```{r this is the name}
library(nycflights13)

swag <- flights
```

can navigate / find them easier, and will have graphics that are named useful

and can set up networks of cached chunks to avoid re-performing expensive
  computations on every run - more below
  
chunk name setup implies special - will be run automatically once, first

## 27.4.2 Chunk options

lots of options to customise code chunks
- eval = FALSE - prevents code form being evaluated, good for example code
- include = FALSE - runs code but doesn't show code or results in final doc
  use for setup code don't want cluttering report
- echo = FALSE - prevents code but not results from appearing in finished file
- message = FALSE prevents messages from appearing in finished file
- results = 'hide' hides printed output, fig.show = 'hide' hides plots
- error = TRUE causes render to continue even if code returns error

## 27.4.3 Table

by default, tables are printed as would in console
- if prefer data displayed with additional formatting, can use knitr::kable

## 27.4.4 Caching

normally each knit of document starts from completely clean slate
- great for reproducibility, b/c ensures have captured every important computation
  in code. but can be a pain if some computations take a long time
- so, use cache = TRUE - will save output of chunk to specially named file on
  disk
- on subsequent runs, knitr will check to see if code has changed, if it hasn't
  will reuse cached results
  
good idea to regularly clear out all caches with knitr::clean_cache()

use advice of David Robinson for naming - each chunk named after primary object
  it creates - makes easier to understand depends on specifications
  
## 27.4.5 Global options

as work with knitr, will discover that some of default chunk options don't fit
  needs and want to change them - can do with knitr::opts_chunk$set() in code
  chunk

## 27.4.6 Inline code

interesting! ccan also embed R code into R Markdown document, directly into text
  with 'r'
doing so means results of code will automatically be inserted into text when
  document is knit

ex
We have data about `r nrow(diamonds)` diamonds. Only `r nrow(diamonds) - nrow(smaller)` 
  are larger than 2.5 carats. The distribution of the remainder is shown below:

in these cases, format() is your friend, helps formatting inline code

## 27.4.7 Exercises

1. Add a section that explores how diamond sizes vary by cut, colour, and clarity. 
  Assume you’re writing a report for someone who doesn’t know R, and instead of 
  setting echo = FALSE on each chunk, set a global option.

2. Download diamond-sizes.Rmd from https://github.com/hadley/r4ds/tree/master/rmarkdown. 
  Add a section that describes the largest 20 diamonds, including a table that 
  displays their most important attributes.

3. Modify diamonds-sizes.Rmd to use comma() to produce nicely formatted output. 
  Also include the percentage of diamonds that are larger than 2.5 carats.

4. Set up a network of chunks where d depends on c and b, and both b and c depend 
  on a. Have each chunk print lubridate::now(), set cache = TRUE, then verify 
  your understanding of caching.

*copied answers, put in test documents in scripts folder*


--------------------------------------------------------------------------------
# 27.5 Troubleshooting

can be harder to troubleshoot R Markdown documents b/c no longer in interactive
  R environment, need to learn new tricks
- first thing should be to try to recreate problem in interactive session
  restart R, then "Run all chunks"
- if that doesn't help, must be something different b/ween interactive env and
  R markdown environment. most common difference is working directory
```{r}
getwd()
```

next, brainstorm all things that might cause bug
- need to systematically check that are same in R session and R markdown session
- easiest way to do that is to set error = TRUE on the chunk causing problem, 
  then use print() and str() to check that settings are as you expect


--------------------------------------------------------------------------------
# 27.6 YAML header

can control many other "whole document" settings by tweaking parameters of 
YAML header - YAML = yet another markup language

## 27.6.1 Parameters

parameters are useful when want to re-render same report with distinct values
  for various key inputs
- use params field
- good way to specify date/time parameters

can also call rmarkdown::render() with a list of params

## 27.6.2 Bibliographies and citations

Pandoc can automatically generate citations a bibliography in number of styles
- to use this feature, specify bib file using bibliography field in file's 
  header. field should contain a path from directory that contains .Rmd file to
  file that contains bibliography file


--------------------------------------------------------------------------------
# 27.7 Learning more

highly recommend Git and GitHub~~~
- books they recommend:
“Happy Git with R”: a user friendly introduction to Git and GitHub from R users, 
  by Jenny Bryan. The book is freely available online: http://happygitwithr.com

The “Git and GitHub” chapter of R Packages, by Hadley. You can also read it for 
  free online: http://r-pkgs.had.co.nz/git.html.








