---
title: "Ch28_lab"
author: "Holden Jones"
date: '2023-02-01'
output: html_document
---

# Graphics for communication


--------------------------------------------------------------------------------
# 28.1 Introduction

after understanding data, need to communicate understanding to others

## 28.1.1 Prerequisites

```{r}
library(tidyverse)
```


--------------------------------------------------------------------------------
# 28.2 Label

easiest place to upgrade graphic is with good labels
- add labels with labs() function
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  labs(title = "Fuel efficiency generally decreases with engine size")
```

purpose of plot title is to summarise main finding
- avoid titles that just describe what plot is
- subtitle adds additional detail in smaller font beneath title
- caption adds text at bottom right of plot
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  labs(
    title = "Fuel efficiency generally decreases with engine size",
    subtitle = "Two seaters (sports cars) are an exception because of their light weight",
    caption = "Data from fueleconomy.gov"
  )
```

also use labs to replace x, y variable names
- generally good to include units, make short names more descriptive, longer
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  geom_smooth(se = FALSE) +
  labs(
    x = "Engine displacement (L)",
    y = "Highway fuel economy (mpg)",
    colour = "Car type"
  )
```

## 28.2.1 Exercises

1. Create one plot on the fuel economy data with customised title, subtitle, 
  caption, x, y, and colour labels.
```{r}
ggplot(mpg, aes(cty, hwy, colour = class)) +
  geom_point() +
  labs(x = "City Miles per Gallon", y = "Highway Miles per Gallon", colour = "Car Class",
       title = "Car Stuff", subtitle = "City and highway mileage are closely related", 
       caption = "Who would've thought!") 
```

2. The geom_smooth() is somewhat misleading because the hwy for large engines 
  is skewed upwards due to the inclusion of lightweight sports cars with big 
  engines. Use your modelling tools to fit and display a better model.
```{r}
ggplot(mpg, aes(displ, hwy)) + geom_point() + geom_smooth(method = 'lm', 
                                                          se = FALSE) + 
  labs(x = "Engine Displacement", y = "Highway Miles per Gallon")
```

3. Take an exploratory graphic that you’ve created in the last month, and add 
  informative titles to make it easier for others to understand.
*SKIP*


--------------------------------------------------------------------------------
# 28.3 Annotations

often useful to label individual observations or groups of observation
- first tool at disposal is geom_text() - can add textual labels to plots
- one approach is have a tibble that provides labels
```{r}
best_in_class <- mpg %>%
  group_by(class) %>%
  filter(row_number(desc(hwy)) == 1)

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  geom_text(aes(label = model), data = best_in_class)

# use nudge_y to move labels slightly above corresponding points
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  geom_label(aes(label = model), data = best_in_class, nudge_y = 2, alpha = 0.5)

# nice, but look there's annoying overlap in labels, use ggrepel package
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  geom_point(size = 3, shape = 1, data = best_in_class) +
  ggrepel::geom_label_repel(aes(label = model), data = best_in_class)
```

note that also added second layer of large hollow points to highlight labelled ones

can use same idea to replace legen with labels directly on plot
```{r}
class_avg <- mpg %>%
  group_by(class) %>%
  summarise(
    displ = median(displ),
    hwy = median(hwy)
  )

ggplot(mpg, aes(displ, hwy, colour = class)) +
  ggrepel::geom_label_repel(aes(label = class),
    data = class_avg,
    size = 6,
    label.size = 0,
    segment.color = NA
  ) +
  geom_point() +
  theme(legend.position = "none")
```

or maybe just want to add single label to plot, but still need to create data frame
```{r}
label <- mpg %>%
  summarise(
    displ = max(displ),
    hwy = max(hwy),
    label = "Increasing engine size is \nrelated to decreasing fuel economy."
  )

ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  geom_text(aes(label = label), data = label, vjust = "top", hjust = "right")
```

*reference chapter for more detailed specifications with label placement*

additional tools to help annotate plot
- geom_hline(), geom_vline() to add reference lines
- geom_rect() to draw rectangle around points of interest
- geom_segment() with arrow argument to draw attention to point with an arrow

## 28.3.1 Exercises

*skip for now*
1. Use geom_text() with infinite positions to place text at the four corners of 
  the plot.

2. Read the documentation for annotate(). How can you use it to add a text label 
  to a plot without having to create a tibble?

3. How do labels with geom_text() interact with faceting? How can you add a 
  label to a single facet? How can you put a different label in each facet? 
  (Hint: think about the underlying data.)

4. What arguments to geom_label() control the appearance of the background box?

5. What are the four arguments to arrow()? How do they work? Create a series of 
  plots that demonstrate the most important options.


--------------------------------------------------------------------------------
# 28.4 Scales

third way to make plot better for communication is to adjust scale
- scales control mapping from data values to things that you can perceive
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class))

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  scale_x_continuous() +
  scale_y_continuous() +
  scale_colour_discrete()
```

default scale named according to type of variable they align with - continuous,
  discrete, datetime, or date
lots of non-default scales which will learn about below

## 28.4.1 Axis ticks and legend keys

2 primary args affect appearence of ticks on axes - breaks and labels
```{r}
# most common use of breaks to override default choice
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  scale_y_continuous(breaks = seq(15, 40, by = 5))

# can use labels in same way
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  scale_x_continuous(labels = NULL) +
  scale_y_continuous(labels = NULL)

# another use of breaks is when have relatively few data and want to show exactly
presidential %>%
  mutate(id = 33 + row_number()) %>%
  ggplot(aes(start, id)) +
    geom_point() +
    geom_segment(aes(xend = end, yend = id)) +
    scale_x_date(NULL, breaks = presidential$start, date_labels = "'%y")
```

## 28.4.2 Legend layout

control overall position of legend with theme() setting
- in brief, control non-data parts of the plot
```{r}
base <- ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class))

base + theme(legend.position = "left")
base + theme(legend.position = "top")
base + theme(legend.position = "bottom")
base + theme(legend.position = "right") # the default
```

can also use legend.position = "none" to suppress display of legend altogether
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  geom_smooth(se = FALSE) +
  theme(legend.position = "bottom") +
  guides(colour = guide_legend(nrow = 1, override.aes = list(size = 4)))
```

## 28.4.3 Replacing a scale

2 types of scales most likely to want to switch out
- continuous position scales
- colour scales

very usefl to plot transformations of variable
- ex as have seen in diamond prices easier to see precise relationship bween
  carat and price if we log transform them
```{r}
ggplot(diamonds, aes(carat, price)) +
  geom_bin2d()

ggplot(diamonds, aes(log10(carat), log10(price))) +
  geom_bin2d()

# disadvantage of transformaton is that axes now labelled with transformed values
# can do transformation with the scale - really?! axes labelled on og data scale
ggplot(diamonds, aes(carat, price)) +
  geom_bin2d() + 
  scale_x_log10() + 
  scale_y_log10()
```

color also frequently customised
- 2 below plots look similar, but plot on right can be distinguished by color
  blind folks
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = drv))

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = drv)) +
  scale_colour_brewer(palette = "Set1")

# also can add redundant shape mapping - ensures plot is interpretable in b/w
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = drv, shape = drv)) +
  scale_colour_brewer(palette = "Set1")
```

colourbrewer scales - in RColorBrewer package

when have predefined mapping between values and colours, use scale_colour_manual()
- ex presidential map by party
```{r}
presidential %>%
  mutate(id = 33 + row_number()) %>%
  ggplot(aes(start, id, colour = party)) +
    geom_point() +
    geom_segment(aes(xend = end, yend = id)) +
    scale_colour_manual(values = c(Republican = "red", Democratic = "blue"))
```

for continuous color - can use built-in scale_colour_gradient() or 
  scale_fill_gradient(). if have diverging scale, can use scale_colour_gradient2()
- another option is scale_colour_virdisi() from viridis package
```{r}
df <- tibble(
  x = rnorm(10000),
  y = rnorm(10000)
)
ggplot(df, aes(x, y)) +
  geom_hex() +
  coord_fixed()

ggplot(df, aes(x, y)) +
  geom_hex() +
  viridis::scale_fill_viridis() +
  coord_fixed()
```

## 28.4.4 Exercises

*skip for now*
1. Why doesn’t the following code override the default scale?
```{r}
ggplot(df, aes(x, y)) +
  geom_hex() +
  scale_colour_gradient(low = "white", high = "red") +
  coord_fixed()
```


2. What is the first argument to every scale? How does it compare to labs()?

3. Change the display of the presidential terms by:
- Combining the two variants shown above.
- Improving the display of the y axis.
- Labelling each term with the name of the president.
- Adding informative plot labels.
- Placing breaks every 4 years (this is trickier than it seems!).

4. Use override.aes to make the legend on the following plot easier to see.
```{r}
ggplot(diamonds, aes(carat, price)) +
  geom_point(aes(colour = cut), alpha = 1/20)
```


--------------------------------------------------------------------------------
# 28.5 Zooming

3 ways to control plot limits
- adjusting what data are plotted
- setting limits in each scale
- setting xlim and ylim in coord_cartesian()

use coord_cartesian() to zoom in on region of plot
```{r}
ggplot(mpg, mapping = aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth() +
  coord_cartesian(xlim = c(5, 7), ylim = c(10, 30))

mpg %>%
  filter(displ >= 5, displ <= 7, hwy >= 10, hwy <= 30) %>%
  ggplot(aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth()
```

can also set limits on individual scales - reducing limits basically 
  equivalent to subsetting data
```{r}
suv <- mpg %>% filter(class == "suv")
compact <- mpg %>% filter(class == "compact")

ggplot(suv, aes(displ, hwy, colour = drv)) +
  geom_point()

ggplot(compact, aes(displ, hwy, colour = drv)) +
  geom_point()
```

one way to overcome problem is to share scales across multiple plots
```{r}
x_scale <- scale_x_continuous(limits = range(mpg$displ))
y_scale <- scale_y_continuous(limits = range(mpg$hwy))
col_scale <- scale_colour_discrete(limits = unique(mpg$drv))

ggplot(suv, aes(displ, hwy, colour = drv)) +
  geom_point() +
  x_scale +
  y_scale +
  col_scale

ggplot(compact, aes(displ, hwy, colour = drv)) +
  geom_point() +
  x_scale +
  y_scale +
  col_scale
```


--------------------------------------------------------------------------------
# 28.6 Themes

finally, can customise non-data elements of plot with theme
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_bw()
```

8 themes by default - many more included in add-on packages like ggthemes


--------------------------------------------------------------------------------
# 28.7 Saving your plots

2 main ways to get plots out of R and into final write-up
- ggsave() and knitr
```{r}
# ggsave() saves most recent plot to disk
ggplot(mpg, aes(displ, hwy)) + geom_point()

# sweet! put path = "output" to save in output folder :)
ggsave("my-plot.pdf", path = "output")
```

## 28.7.1 Figure sizing

biggest challenge of graphics in R Markdown is getting figures right size + shape
Hadley only uses three of five options
- have plots at consistent width, set fig.width = 6 and fig.asp = 0.618 in
  defaults (the golden ratio). then only adjust fig.asp in individual chunks
- control output size with out.width and set it to percentage of line width
  default to out.width = 70% and fig.align = "center"
- to put multiple plots in single row set out.width to 50% for 2, 33% for 3 etc

if find need to squint to read text in plot, need to tweak fig.width

if want to make sure font size consistent across all figures, whenever set
  out.width, also need to adjust fig.width to maintain same ratio with default

## 28.7.2 Other important options

recommend setting fig.show = "hold" when mixing code and text like do in this
  book
use fig.cap to add caption to plot
if displaying plots with thousands of data points, set dev = "png" for speed

*good idea to name code chunks that produce figures, even if don't name others*
- chunk label used to generate file name of graphic on disk, easier to pick out
  and reuse
  

--------------------------------------------------------------------------------
# 28.8 Learning more

best place to learn more is ggplot2 book - elegant graphics for data analysis
- source code online at github

ggplot2 extensions gallery - site has many packages that extend ggplot2