---
title: "Graphics for communication"
output: html_notebook
---
  
```{r}
library(tidyverse)
#install.packages("viridis")
library(viridis) #You'll probably need to install this
#install.packages("ggrepel")
library(ggrepel)
```


```{r}
#run this stuff for section 28.4 (Scales)
df <- tibble(
  x = rnorm(10000),
  y = rnorm(10000)
)
ggplot(df, aes(x, y)) +
  geom_hex() +
  coord_fixed()

ggplot(df, aes(x, y)) +
  geom_hex() +
  viridis::scale_fill_viridis() +
  coord_fixed()
```

# 28.2 Label!
########################################################################
##https://r4ds.had.co.nz/graphics-for-communication.html#exercises-74 ##
########################################################################

```{r}
best_in_class <- mpg %>%
  group_by(class) %>%
  filter(row_number(desc(hwy)) == 1)

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  geom_text(aes(label = model), data = best_in_class)

# can nudge labels within geom_label so they don't overlap
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  geom_label(aes(label = model), data = best_in_class, nudge_y = 2, alpha = 0.5)

# but still will have overlap if labels are for same point. So can use ggrepel package which says don't make any labels overlap! useful

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  geom_point(size = 3, shape = 1, data = best_in_class) +
  ggrepel::geom_label_repel(aes(label = model), data = best_in_class)
```

```{r}
label <- mpg %>%
  summarise(
    displ = max(displ),
    hwy = max(hwy),
    label = "Increasing engine size is \nrelated to decreasing fuel economy."
  )

ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  geom_text(aes(label = label), data = label, vjust = "top", hjust = "right")
```

```{r}
label <- tibble(
  displ = Inf,
  hwy = Inf,
  label = "Increasing engine size is \nrelated to decreasing fuel economy."
)

ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  geom_text(aes(label = label), data = label, vjust = "top", hjust = "right")
```


1. Create one plot on the fuel economy data (`mpg`) with customised `title`, `subtitle`, `caption`, `x`, `y`, and `colour` labels.

```{r}
ggplot(mpg, aes(cty, hwy, colour = class)) +
  geom_point() +
  labs(x = "City Miles per Gallon", y = "Highway Miles per Gallon", colour = "Car Class",
       title = "Car Stuff", subtitle = "City and highway mileage are closely related", 
       caption = "Who would've thought!") 
```

2. The `geom_smooth()` is somewhat misleading because the `hwy` for large engines is skewed upwards due to the inclusion of lightweight sports cars with big engines. Use your modelling tools to fit and display a better model (hint use a linear model (`lm`) in geom_smooth).

```{r}
ggplot(mpg, aes(displ, hwy)) + geom_point() + geom_smooth(method = 'lm', se = FALSE) + labs(x = "Engine Displacement", y = "Highway Miles per Gallon")
```

3. Take an exploratory graphic that you’ve created in the last month, and add informative titles to make it easier for others to understand.
**SKIP**
  
#######################################################################
##https://r4ds.had.co.nz/graphics-for-communication.html#exercises-75##
#######################################################################

# 28.3 Annotations

**This section is optional. Take a look if you think annotating plots will be useful for you.**

1. Use `geom_text()` with infinite positions to place text at the four corners of the plot.

2. Read the documentation for `annotate()`. How can you use it to add a text label to a plot without having to create a tibble?
  
3. How do labels with `geom_text()` interact with faceting? How can you add a label to a single facet? How can you put a different label in each facet? (Hint: think about the underlying data.)

4. What arguments to `geom_label()` control the appearance of the background box?
  
5. What are the four arguments to `arrow()`? How do they work? Create a series of plots that demonstrate the most important options.



#######################################################################
##https://r4ds.had.co.nz/graphics-for-communication.html#exercises-76##
#######################################################################

28.4 Scales

1. Why doesn’t the following code override the default color scale?
  
```{r}
ggplot(df, aes(x, y)) +
  geom_hex() +
  scale_colour_gradient(low = "white", high = "red") +
  coord_fixed()
```

2. What is the first argument to every scale? How does it compare to `labs()`?
  
3. Change the display of the `presidential` terms by doing one or more of the below modifications:

  
- Combining the two variants shown above.
- Improving the display of the y axis.
- Labelling each term with the name of the president.
- Adding informative plot labels.
- Placing breaks every 4 years (this is trickier than it seems!).


4. Use override.aes to make the legend on the following plot easier to see.

```{r}
ggplot(diamonds, aes(carat, price)) +
  geom_point(aes(colour = cut), alpha = 1/20)
```


28.6-7 Themes and Saving

1. Modify the above the plot below by testing the different built-in themes (https://d33wubrfki0l68.cloudfront.net/6be25cfb939e69d72df9fbc68e7a0334e1febefe/10e92/images/visualization-themes.png). You can also install and mess around with  the `ggthemes` package  (https://github.com/jrnold/ggthemes)

```{r}
ggplot(diamonds, aes(carat, price)) +
  geom_point(aes(colour = cut), alpha = 1/20)
```

2. Save your above plot using a width of 6 inches (`fig.width`) and a figure aspect (`fig.asp`) of 0.618.

```{r, fig.width=6, fig.asp=0.618}
ggplot(diamonds, aes(carat, price)) +
  geom_point(aes(colour = cut), alpha = 1/20)
```
