---
title: "Ch29_lab"
author: "Holden Jones"
date: '2023-02-01'
output: html_document
---

# R Markdown formats


--------------------------------------------------------------------------------
# 29.1 Introduction

chapter gives brief overview of other types of output can produce with markdown

2 ways to set output of document
```{r}
# permanently, by modifying YAML header
title: "Viridis Demo"
output: html_document

# transiently, by calling rmarkdown::render() by hand
rmarkdown::render("diamond-sizes.Rmd", output_format = "word_document")
```

R Studio's knit button renders file to first format listed in output field
- also can render to additional formats via GUI


--------------------------------------------------------------------------------
# 29.2 Output options

each output format associated with an R function
- to override default parameter values, need to expand output field
- ex render html_document with floating table of contents
```{r}
output:
  html_document:
    toc: true
    toc_float: true
    
# can even render to multiple outputs by supplying list of formats
output:
  html_document:
    toc: true
    toc_float: true
  pdf_document: default
```


--------------------------------------------------------------------------------
# 29.3 Documents

number of variations on default html_document output, generating different
  types of documents
  

--------------------------------------------------------------------------------
# 29.4 Notebooks

- html_document focused on communicating with decision makers
- notebook focused on collaborating with other data scientists

notebook also contains full source code, both contain fully rendered output

interesting - use both html_notebook and github_document outputs
- html_notebook gives local preview, and file that can share via email
- github_document creates minimal md file that can check into git, renders nicely
  on Git
  

--------------------------------------------------------------------------------
# 29.5 Presentations

automatically inserting code and results into presentation can save huge
  amount of time
- begin new slide at # or ##
- *** makes new slide without header

comes with three presentation formats built in, with others provided by packages


--------------------------------------------------------------------------------
# 29.6 Dashboards

useful way to communicate large amounts of information visually and quickly

code in Ch29_dashboard creates nice lil dashboard for diamonds
- but need flexdashboard package, issues with internet in Galapagos


--------------------------------------------------------------------------------
# 29.7 Interactivity

any HTML format can contain interactive components

## 29.7.1 htmlwidgets

ex - leaflet map below, can interact if viewing on web
rmarkdown inserts static screenshot
```{r}
library(leaflet)
leaflet() %>%
  setView(174.764, -36.877, zoom = 16) %>% 
  addTiles() %>%
  addMarkers(174.764, -36.877, popup = "Maungawhau") 
```

great thing about widgets is that don't need to know anything about HTML or
  JavaScript to use them, all details wrapped inside package

## 29.7.2 Shiny

alternative to widgets is using shiny, a package that allows you to create
  interactivity using R code, not JavaScript
  
to call Shiny code from R Markdown doc, add runtime: shiny to header
```{r}
title: "Shiny Web App"
output: html_document
runtime: shiny
```

then can use "input" functions to add interactive components to document
```{r}
library(shiny)

textInput("name", "What is your name?")
numericInput("age", "How old are you?", NA, min = 0, max = 150)
```

can then refer to values as input$name and input$age and code that uses them
  will be automatically re-run whenever the values change!!!!
  
logistical issue - Shiny apps need a Shiny server to be run online

tradeoff - can do anything in a shiny document that can do in R, but it requires
  someone to be running R
  

--------------------------------------------------------------------------------
# 29.8 Websites

with a little additional infrastructure, can use R Markdown to generate complete
  website
- put .Rmd files in single directory. index.Rmd will become home page
- Add a YAML file named _site.yml provides navigation for site
```{r}
name: "my-website"
navbar:
  title: "My Website"
  left:
    - text: "Home"
      href: index.html
    - text: "Viridis Colors"
      href: 1-example.html
    - text: "Terrain Colors"
      href: 3-inline.html
```


--------------------------------------------------------------------------------
# 29.9 Other formats

other packages provide even more formats - review if necessary


--------------------------------------------------------------------------------
# 29.10 Learning more

to learn more about effective communication in different formats, following
  resources
- academic talks - Leek group guide to giving talks
- improve presentation skills - Presentation Patterns


