---
title: "Occurrence Maps"
output: html_document
---

```{r}
# install.packages("maptools")
# install.packages("ggplot2")
install.packages("sp") # problem child here
# install.packages("rnaturalearth")
# install.packages("rgdal")
# install.packages("ggmap")

library("maptools")
library("ggplot2")
library("sp") # error here. triggered a fatal error! session aborted. never seen that before
library("rnaturalearth") # error here. oh no!
library("rgdal") # good
library("ggmap") # good. strange!


```

# Making maps with R! with Dr. Laura Jimenez Jimenez

R has functionality of traditional GIS packages (such as ArcGIS and QGIS)
R is great tool for geospatial data analysis
many ways to handle geographic data in R, lots of packages available
R enables a reproducible workflow

# 2 steps to build a map

1. find and load the data layers
need to find region boundaries and all the data layers that you want to include in your map
these data layers usually stored in shapefiles, geoJSON files, TIFF files, raster files
these layers are geospatial objects that have specific features so you will need special R packages to load and manipulate data

2. manipulate and plot
once geo-data are loaded in R, you acn manipulate and plot the different layers with functions included in packages such as sp or ggplot2

# where do I get spatial data from
spatial data is bulky
geodatabases have capacity and capability to store virtually all types of geographic datasets. a lot are open-access
ESRI, USGS, HI statewide GIS, IUCN Red List, Global Biodiversity Information Facility (GBIF) - check this out!, WorldClim - free climate data for ecological modeling and GIS

Geocomputation with R, pacakge vignettes: dismo, terra, sp, rgdal

r-graph-gallery.com
r-spatial.org
geocompr.robinlovelace.net
rspatial.org/terra
rspatialdata.github
rstudio.github.io/leaflet

esri for GIS concepts

```{r}
spp.pnts <- read.csv("species_occurrences_GBIF.csv", header = T)
dim(spp.pnts) # give dimensions
head(spp.pnts) # look at the first few rows
```
identify differen species in dataset
```{r}
spp <- factor(spp.pnts$SpeciesName) # make species name a factor, call it spp
(spnames <- levels(spp)) # we have 9 different species
table(spp) # how many records of each species
```

plot the map and the occurrence points on top using long and lat
```{r}
data("wrld_simpl", package = "maptools") # load baseline map from package maptools

plot(wrld_simpl) # plot the map
points(spp.pnts$longitude, spp.pnts$latitude, pch = 20) # put points right on map, need to run current chunk as one
```
using different colors to identify species
```{r}
plot(wrld_simpl) # plot the map
points(spp.pnts$longitude, spp.pnts$latitude, pch = 20, col =spp) # put points right on map, col by spp (species)
```
try with a different palette
```{r}
palette(hcl.colors(length(spnames), "viridis")) # change palette - would need to change back if want to
  # palette("default") # to get default palette back
plot(wrld_simpl)
points(spp.pnts$longitude, spp.pnts$latitude, pch = 20, col =spp)
```
changing background colors
```{r}
plot(wrld_simpl, col = 'wheat1', bg = 'lightcyan1') # can change color of background map too
points(spp.pnts$longitude, spp.pnts$latitude, pch = 20, col =spp)
legend("bottomleft", legend=spnames, bty = "n", pch = 20, col =1:length(spnames))
```
above map looks good but lets remove antarctica cause no need

```{r}
#raster::extent(wrld_simpl) # no package called raster darn
plot(wrld_simpl, col = 'wheat1', bg = 'lightcyan1', ylim=c(-20, 80))
points(spp.pnts$longitude, spp.pnts$latitude, pch = 20, col =spp)
legend("bottomleft", legend=spnames, bty = "n", pch = 20, col =1:length(spnames))
```

png(filename = "global-map-1.png", width = 30, height = 15, units = 'cm', res = 300, pointsize = 10)

can make a map for a subset of species - just one here
```{r}
spnames[7]
sp.data <- subset(spp.pnts, spp.pnts$SpeciesName==spnames[7])
range(sp.data$longitude) # what is the range of the longitude and latitude? so can zoom the map into these points
range(sp.data$latitude)
```


